---
title: "aqim volume spatial layers"
author: "Dalon White"
date: "2025-08-28"
output: html_document

params:

  # Set number of simulations
  n_sims_for_volume: 100
  pathway_volumes_filename: "volume_of_risk_by_pathway" # Name of the output file for the pathway risk summary results
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using 
---
```{r setup, include=FALSE}
pacman::p_load(tidyverse, magrittr, ggplot2, DT, caret, lubridate, readr, stringdist, purrr)
pacman::p_load(DBI, odbc)
pacman::p_load(tidygeocoder, sf, tigris, SpatialKDE, here)
```



```{r params for vscode}
if(!exists("params")) {
  params <- list(
   # Set number of simulations
   n_sims_for_volume = 100,
   affected_countries = "current", # "current" or "mex_can" or "other"
   other_affected_countries = c("Country1 ISO-3 Code", "Country2 ISO-3 Code"), # If using "other", specify the countries here
   pathway_volumes_filename = "volume_of_risk_by_pathway" # Name of the output file for the pathway risk summary results
  )
}
```
# Steps to spatial mappings

Have Location ID & Location Name
These correspond to BRG2_LOCATION IDs
Can connect to CBP Port Number & IATA code (sometimes both) thruogh this BRG
CBP Port Number is geolocated in the GeoLocations database in ARM; IATA codes are geolocated elsewhere

Will have to check to see how many Location Type isn't mapped at this point; PPQ things probably won't be
    PPQ field offices/PIS are probably on GeoLocations as well - see tbl_PPQ_Facilities
    use LocationName and Longitude + Latitude

```{r setup, include=FALSE}
setwd(here::here("output","aqim based volume of at risk product approaching"))
pathway_risk_summary <- readRDS(paste(params$pathway_volumes_filename, params$n_sims_for_volume, "sims",".rds", sep = ""))
```

# Gather geolocation data
```{r geolocation data}
#-- Connect to ARM ---- 
#ARM data base version 2
db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

geo_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=GeoLocations;
                        trusted_connection=yes")


brg_loc_cols <- c("ID",
                "NAME",
                "SITE",
                "PHYSICAL_ADDRESS_ID",
                "MAIL_ADDRESS_ID",
                'CBP_PORT_NUMBER',
                "IATA_CODE")
cbp_poe_cols <- c("CBP_NAME", "CBP_CODE", "LAT", "LON", "Port of Entry Name")
ppq_facilities_cols <- c("LocationName", "PPQCode", "City", "State", "Longitude", "Latitude")
loc_280_cols <- c("LOCATION", "ARM_InspectionLocationName", "Latitude", "Longitude")


```


# gather coords from difference sources

```{r}
# BRG2_LOCATION
brg_loc <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_LOCATION]")) |> 
               dplyr::select(all_of(brg_loc_cols)) |> 
              collect()


cbp_poe <- tbl(geo_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_CBP_POEs]")) |> 
              dplyr::select(all_of(cbp_poe_cols)) |> 
              collect()


ppq_facilities <- tbl(db_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_PPQ_Facilities]")) |> dplyr::select(all_of(ppq_facilities_cols)) |> collect()


pathway_risk_brg_loc <- pathway_risk_summary |> left_join(brg_loc, by = c("LOCATION_ID" = "ID"))


# Step 1: Join with CBP POE data first
pathway_risk_geo <- pathway_risk_brg_loc %>%
  left_join(cbp_poe, by = c("CBP_PORT_NUMBER" = "CBP_CODE")) %>%
  # Create initial LAT/LON columns that might be filled
  rename(
    CBP_LAT = LAT,
    CBP_LON = LON,
    CBP_PORT_NAME = `Port of Entry Name`
  )

# Step 2: Join with PPQ facilities for rows still missing coordinates
pathway_risk_geo <- pathway_risk_geo %>%
  left_join(
    ppq_facilities %>% 
      dplyr::select(LocationName, Latitude, Longitude) %>%
      rename(
        PPQ_LAT = Latitude,
        PPQ_LON = Longitude
      ),
    by = c("LOCATION_NAME" = "LocationName")
  )

# Step 3: Create combined coordinate columns, prioritizing CBP data
pathway_risk_geo2 <- pathway_risk_geo %>%
  mutate(
    final_LAT = coalesce(CBP_LAT, PPQ_LAT),
    final_LON = coalesce(CBP_LON, PPQ_LON)
  )

# Step 4: For remaining missing coordinates, use IATA code if available
missing_coords <- pathway_risk_geo2 %>%
  filter(is.na(final_LAT) | is.na(final_LON)) %>%
  filter(!is.na(IATA_CODE))

if(nrow(missing_coords) > 0) {
  # Option 1: Use tidygeocoder package if available
  if(requireNamespace("tidygeocoder", quietly = TRUE)) {
    iata_coords <- missing_coords %>%
      dplyr::select(IATA_CODE) %>%
      distinct() %>%
      mutate(
        # Look up IATA codes - this creates lat/lng columns
        query = paste("airport", IATA_CODE)
      ) %>%
      tidygeocoder::geocode(address = query, method = "osm")
    
    # Join the results back
    pathway_risk_geo3 <- pathway_risk_geo2 %>%
      left_join(
        iata_coords %>% 
          dplyr::select(IATA_CODE, lat, long) %>%
          rename(IATA_LAT = lat, IATA_LON = long),
        by = "IATA_CODE"
      ) %>%
      mutate(
        final_LAT = coalesce(final_LAT, IATA_LAT),
        final_LON = coalesce(final_LON, IATA_LON)
      )
  } else {
    # Option 2: Create a simple example for manual IATA lookup
    cat("Found", nrow(missing_coords), "rows with IATA codes but no coordinates.\n")
    cat("Consider using the 'airportr' or 'tidygeocoder' package to look up these codes.\n")
    print(missing_coords %>% dplyr::select(IATA_CODE) %>% distinct())
  }
}

# Step 5: Evaluate final coverage
coords_summary <- pathway_risk_geo3 %>%
  summarize(
    total_rows = n(),
    rows_with_coords = sum(!is.na(final_LAT) & !is.na(final_LON)),
    percent_coverage = round(100 * rows_with_coords / total_rows, 1)
  )

print(coords_summary)

#risk coverage
risk_summary <- pathway_risk_geo3 %>%
  summarize(
    total_risk = sum(mean_positives, na.rm = TRUE),
    risk_with_coords = sum(mean_positives[!is.na(final_LAT) & !is.na(final_LON)], na.rm = TRUE),
    percent_risk = round(100 * risk_with_coords / total_risk, 1)
  )

print(risk_summary)

# DONE!



# Count which source provided the coordinates
source_summary <- pathway_risk_geo3 %>%
  mutate(
    coord_source = case_when(
      !is.na(CBP_LAT) & !is.na(CBP_LON) ~ "CBP",
      !is.na(PPQ_LAT) & !is.na(PPQ_LON) ~ "PPQ",
      !is.na(IATA_LAT) & !is.na(IATA_LON) ~ "IATA",
      TRUE ~ "Missing"
    )
  ) %>%
  count(coord_source) %>%
  mutate(percent = round(100 * n / sum(n), 1))

print(source_summary)

# Show examples of remaining locations without coordinates
if(coords_summary$rows_with_coords < coords_summary$total_rows) {
  cat("\nSample of locations still missing coordinates:\n")
  print(pathway_risk_geo3 %>% 
          filter(is.na(final_LAT) | is.na(final_LON)) %>%
          dplyr::select(LOCATION_NAME, PATHWAY) %>%
          distinct())
}

```


```{r}
if(requireNamespace("tidygeocoder", quietly = TRUE)) {
    iata_coords <- missing_coords %>%
      dplyr::select(IATA_CODE) %>%
      distinct() %>%
      mutate(
        # Look up IATA codes - this creates lat/lng columns
        query = paste("airport", IATA_CODE)
      ) %>%
      tidygeocoder::geocode(address = query, method = "osm")
}
```

# convert to spatial
```{r}
# convert to spatial
pathway_risk_geo_sf <- pathway_risk_geo3 |> filter(!is.na(final_LAT) | !is.na(final_LON)) |> 
  sf::st_as_sf(coords = c("final_LON", "final_LAT"), crs = 4326) |>
  # Then transform to NAD83/Albers (EPSG:5070)
  sf::st_transform(crs = 5070)
# save
# saveRDS(pathway_risk_geo_sf, paste("pathway_risk_with_geolocation", ".rds", sep = ""))
```

# Now, can plot spatially
Just an option proof of concept
```{r plot spatially}
# ggplot() +
#   geom_sf(data = pathway_risk_geo_sf, aes(size = mean_positives), color = "blue", alpha = 0.5) +
#   theme_minimal() +
#   labs(title = "Pathway Risk Locations with Geolocation",
#        size = "Mean Positives") +
#   theme(legend.position = "right")
```

# Create US boundaries
```{r}
# Create US boundaries with all territories
# Get all states and territories in one call
us_all <- tigris::states(cb = FALSE) 

us_bounds_sf <- us_all %>%
  st_union() %>%
  st_transform(5070) %>%  # Use same CRS as your data
  st_sf(geometry = .)

pathway_risk_geo_sf_filtered <- pathway_risk_geo_sf %>%
  st_join(us_bounds_sf, left = TRUE) 

# Count how many points were kept vs removed
total_points <- nrow(pathway_risk_geo_sf)
kept_points <- nrow(pathway_risk_geo_sf_filtered)
removed_points <- total_points - kept_points

cat(sprintf("Filtered to %d points within US boundaries (%d removed)\n", 
            kept_points, removed_points))

```

# Now, summarize data and separate for spatial layers
Summarize over year, month

Then, create a spatial layer for each combination of pathway, container, product_pathway, and product_subpathway, with each port have a single value for each combination, and each combination being in different layers
```{r summarize for spatial layers}

spatial_layers_summary <- pathway_risk_geo_sf_filtered %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(mean_kg, na.rm = TRUE)) |> # Assuming independence
  ungroup() |>
  #Filter any that are still all NA's (blanks)
  filter(!is.na(container) & !is.na(product_pathway) & !is.na(product_subpathway))
```

# Create spatial layers
```{r}
create_spatial_layers <- function(summarized_data, save_files = FALSE, output_dir = NULL) {
  # Split data by pathway combinations
  split_data <- summarized_data %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    group_split()
  
  cat(sprintf("Created %d spatial layers for different pathway combinations\n", length(split_data)))
  
  # Create list to store the layers
  spatial_layers <- list()
  
  # Create scenario string based on affected_countries parameter
  scenario_string <- case_when(
    params$affected_countries == "current" ~ "current_affected_countries",
    params$affected_countries == "mex_can" ~ "mex_can_ASF_only",
    params$affected_countries == "other" ~ paste0("custom_", 
                                                paste(params$other_affected_countries, collapse = "_")),
    TRUE ~ "unspecified_scenario"
  )
  
  # Process each split
  for (i in seq_along(split_data)) {
    current_split <- split_data[[i]]
    
    # Get combination values for naming
    combo <- current_split %>%
      dplyr::select(PATHWAY, container, product_pathway, product_subpathway) %>%
      slice(1) %>%
      as.list()
    
    # Create clean name for layer
    layer_name <- paste(
      gsub("[^[:alnum:]]", "_", combo$PATHWAY),
      gsub("[^[:alnum:]]", "_", combo$container),
      gsub("[^[:alnum:]]", "_", combo$product_pathway),
      gsub("[^[:alnum:]]", "_", combo$product_subpathway),
      sep = "_"
    )
    
    # Store in list with clean name
    spatial_layers[[layer_name]] <- current_split
    
    # Save to file if requested
    if (save_files && !is.null(output_dir)) {
      # Fix geometry types before saving - convert all to POINT
      export_layer <- current_split %>%
        st_cast("POINT")  # Convert any multi-points to simple points
      
      # Include scenario in filename
      file_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".shp"))
      
      # Try-catch to handle potential errors
      tryCatch({
        sf::st_write(export_layer, file_path, delete_layer = TRUE)
        cat(sprintf("Saved layer %d/%d: %s\n", i, length(split_data), file_path))
      }, error = function(e) {
        cat(sprintf("ERROR saving layer %d/%d: %s - %s\n", 
                   i, length(split_data), layer_name, e$message))
        
        # Try alternative format if shapefile fails
        geopackage_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".gpkg"))
        cat(sprintf("Trying to save as GeoPackage instead: %s\n", geopackage_path))
        st_write(current_split, geopackage_path, delete_layer = TRUE)
      })
    }
  }
  
  # Create summary table of the layers
  layer_summary <- data.frame(
    layer_name = names(spatial_layers),
    scenario = scenario_string,
    point_count = sapply(spatial_layers, nrow),
    total_risk = sapply(spatial_layers, function(x) sum(x$total_mean_kg, na.rm = TRUE))
  ) %>%
    arrange(desc(total_risk))
  
  # Print summary
  cat(paste0("\nSummary of spatial layers for scenario '", scenario_string, "' (top 10 by risk):\n"))
  print(head(layer_summary, 10))
  
  # Also save the summary table
  if (save_files && !is.null(output_dir)) {
    summary_path <- file.path(output_dir, paste0("layer_summary_", scenario_string, ".csv"))
    write.csv(layer_summary, summary_path, row.names = FALSE)
    cat(sprintf("\nSaved summary table to: %s\n", summary_path))
  }
  
  return(list(
    layers = spatial_layers,
    summary = layer_summary,
    scenario = scenario_string
  ))
}
```

## Create spatial layers and optionally save to files
```{r}

setwd(here::here("output","spatial layers","risk at ports"))
# Apply the function to your summarized data
spatial_layers <- create_spatial_layers(
  summarized_data = spatial_layers_summary,
  save_files = TRUE,
  output_dir = here::here("output", "spatial layers", "risk at ports")
)
```

# Summarize risk entirely by pathway combination, then apply to the appropriate hex grid for final destination

```{r summarize by pathway combination}
pathway_spatial_layers_summary <- spatial_layers_summary %>% st_drop_geometry() |>
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(total_mean_kg, na.rm = TRUE)) |> # Assuming independence
  ungroup()

#for each PATHWAY, container, product_pathway, product_subpathway combination, apply to hex grid for final destination, using the PATHWAY as the key for identifying which hex grid in 'output/proportional destination layers' folder to use
```
```{r}
# Function to distribute pathway risks to spatial grids - with proper mapping
distribute_pathway_risks <- function(pathway_summary, kde_folder_path) {
  # List all available KDE files
  kde_files <- list.files(path = kde_folder_path, pattern = "kde_scaled_.*\\.rds", full.names = TRUE)
  
  # Create list to store results
  distributed_layers <- list()
  
  # Initialize mapping_df ONCE here, outside both loops
  mapping_df <- data.frame(
    short_name = character(),
    full_name = character(),
    pathway = character(),
    container = character(),
    product_pathway = character(),
    product_subpathway = character(),
    stringsAsFactors = FALSE
  )
  
  # Process each pathway
  for (pathway in unique(pathway_summary$PATHWAY)) {
    # Find the matching KDE file
    kde_filename_pattern <- paste0("kde_scaled_", pathway, ".rds")
    matching_files <- grep(kde_filename_pattern, kde_files, value = TRUE, fixed = TRUE)
    
    if (length(matching_files) == 0) {
      cat(sprintf("No KDE file found for pathway: %s\n", pathway))
      next
    }
    
    # Load the KDE file
    kde_grid <- readRDS(matching_files[1])
    
    # Get all combinations for this pathway
    pathway_combos <- pathway_summary %>%
      filter(PATHWAY == pathway)
    
    # Process each combination
    for (i in 1:nrow(pathway_combos)) {
      combo <- pathway_combos[i,]
      
      # Get the risk value for this combination
      risk_value <- combo$total_mean_kg
      
      # Create shorter name with abbreviated components
      pathway_abbr <- substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 3)
      container_abbr <- substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3)
      prod_path_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3)
      prod_subpath_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3)
      
      short_combo_name <- paste(pathway_abbr, container_abbr, prod_path_abbr, prod_subpath_abbr, i, sep = "_")
      
      # Apply the risk to the KDE grid
      distributed_kde <- kde_grid %>%
        mutate(risk_value = .data$kde_value * risk_value)
      
      # Create full combo name
      full_combo_name <- paste(
        gsub("[^[:alnum:]]", "_", combo$PATHWAY),
        gsub("[^[:alnum:]]", "_", combo$container),
        gsub("[^[:alnum:]]", "_", combo$product_pathway),
        gsub("[^[:alnum:]]", "_", combo$product_subpathway),
        sep = "_"
      )
      
      # Store in layers list
      distributed_layers[[full_combo_name]] <- distributed_kde
      
      # Create a new row for this combination and append to mapping_df
      new_mapping <- data.frame(
        short_name = short_combo_name,
        full_name = full_combo_name,
        pathway = combo$PATHWAY,
        container = combo$container,
        product_pathway = combo$product_pathway,
        product_subpathway = combo$product_subpathway,
        stringsAsFactors = FALSE
      )
      
      # Append to the master mapping dataframe
      mapping_df <- rbind(mapping_df, new_mapping)
      
      # Create output directory if it doesn't exist
      output_dir <- here::here('output','spatial layers','risk at final destinations')
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Save to file with shorter name
      output_file <- file.path(output_dir, paste0("risk_", short_combo_name, ".gpkg"))
      
      # Write the file
      tryCatch({
        st_write(distributed_kde, output_file, delete_layer = TRUE)
        cat(sprintf("Processed %s → %s with risk %f\n", 
                    full_combo_name, short_combo_name, risk_value))
      }, error = function(e) {
        cat(sprintf("ERROR: Could not write %s - %s\n", short_combo_name, e$message))
      })
    }
  }
  
  # Save the name mapping file for reference
  write.csv(mapping_df, file.path(output_dir, "filename_mapping.csv"), row.names = FALSE)
  
  return(list(
    layers = distributed_layers,
    mapping = mapping_df
  ))
}

# Apply the function
distributed_risks <- distribute_pathway_risks(
  pathway_spatial_layers_summary,
  here::here("output", "proportional destination layers")
)

# Create a summary of the distributed risks: Extract risk summary from the already-created catalog
risk_summary <- layer_catalog %>%
  select(layer_name = full_name, total_distributed_risk = risk_value, cell_count)

# Print summary
cat("\nSummary of distributed risk layers:\n")
print(head(risk_summary, 10))

# Save summary
write.csv(risk_summary, here::here("output", "aqim total distributed risk", "distributed_risk_summary.csv"), row.names = FALSE)

# 1. Save the entire object as RDS for convenience
output_dir <- here::here('output','spatial layers','risk at final destinations')
saveRDS(distributed_risks, file.path(output_dir, "all_distributed_risks.rds"))

# 2. Save a catalog file that helps with loading specific layers
layer_catalog <- distributed_risks$mapping %>%
  mutate(
    file_path = file.path(output_dir, paste0("risk_", short_name, ".gpkg")),
    risk_value = sapply(distributed_risks$layers[full_name], function(x) sum(x$risk_value, na.rm=TRUE)),
    cell_count = sapply(distributed_risks$layers[full_name], nrow)
  ) %>%
  arrange(desc(risk_value))

# Save the catalog for easy reference
write.csv(layer_catalog, file.path(output_dir, "layer_catalog.csv"), row.names = FALSE)

cat("Saved distributed risk layers and catalog for future use\n")
```

