---
title: "aqim pathway volume per incident"
author: "Dalon White"
date: "2025-08-15"
output: html_document

params:
  risk_quantification: "0_1" # 0_1 or "range"
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  aqim_qmi_size_distribution_filename: "pathway_weight_distribution_models" # Name of the output file for the distribution models
---
Run in conjunction with aqim pathway passenger layers.Rmd and aqim pathway approach rates.Rmd

```{r setup, include=FALSE}
library(tidyverse)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      risk_quantification = "0_1", # 0_1 or "range"
      affected_countries = "current", # "current" or "mex_can" 
      years = 2024:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"),
      aqim_qmi_size_distribution_filename = "pathway_weight_distribution_models"
    )
  }
}
```

# Determine if for ASF countries or mex/can

```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
if(params$affected_countries == "current") {
  affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
} else if (params$affected_countries == "mex_can") {
  affected_countries <- c("MEX", "CAN")
} else if (params$affected_countries == "other") {
  affected_countries <- params$other_affected_countries
} else {
  stop("Invalid value for params$affected_countries")
}


# setwd(here::here("input","data","asf affected countries by year"))
# affected_countries <- read.csv("asf affected countries.csv") |> 
#     mutate(Year = as.numeric(Year)) |> 
#     filter(Year %in% params$years) |> 
#     filter(Disease == "ASF") |>
#     distinct(ISO3_CODE) |> 
#     pull(ISO3_CODE)

# affected_countries
```

# Read and prep data
Read prepared data
Isolate to ASF at risk commodities
Identify the distribution of sizes by pathway and subpathway, entry_type, and container

```{r}

dat <- read.csv("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv")

```

# identify which commodities are at risk for ASF

```{r assign risk}
  # filter to only those with IS_ASF_RISK == "YES"
# Assign risk
commodity_dat_tmp <- dat %>% mutate(IS_ASF_RISK = case_when(
  edge_origin %in% affected_countries | product_origin %in% affected_countries ~ IS_ASF_RISK,
  .default = "NO"
         )
)

  # filter to only those with IS_ASF_RISK == "YES"
# keep pathway, location, commodity, pathway, subpathway, entry_type, container, and TOTAL_KG
commodity_dat <- commodity_dat_tmp %>%
  filter(!IS_ASF_RISK == "NO") %>%
  dplyr::select(
    ID, 
    PATHWAY, 
    PORT_CD,
    CALENDAR_MONTH,
    CALENDAR_YEAR,
    INSPECTION_LOCATION_ID, 
    INSPECTION_LOCATION_NAME, 
    COMMODITY_DISPLAY_NAME, 
    entry_type, 
    container, 
    TOTAL_KG,
    product_pathway,
    product_subpathway
  )
print(paste(nrow(commodity_dat |> filter(is.na(TOTAL_KG))), "entries with NA TOTAL_KG. It is being removed."))


print(
    paste(nrow(commodity_dat_tmp %>%
  filter(!IS_ASF_RISK == "NO")),
  "records with ASF risk")
)

print("These numbers drop off significantly when you filter to only those countries with ASF")

View(dat |> filter(IS_ASF_RISK != "NO") |> count(edge_origin, product_origin) |> arrange(desc(n)))

print("Most items that could carry risk are coming from Mex, Can, or the US; in 4th and 5th, though, are ESP, BHS, PRI (Puerto Rico), with ITA, DEU, and PHL not far behind. These all actually carry risk")

View(commodity_dat_tmp |> filter(IS_ASF_RISK == "NO") |> count(COMMODITY_DISPLAY_NAME, sort=TRUE))
print("I double checked all these commodities to ensure none of these names carried risk, and that the reference tables are working appropriately. They are, and these commodities do not carry risk.")
```

# get the distribution of each

```{r fit distributions to item weights}
if (!requireNamespace("fitdistrplus", quietly = TRUE)) {
  install.packages("fitdistrplus")
}
library(fitdistrplus)
library(dplyr)

# Create a function to fit distributions to each group
fit_distributions <- function(data) {
  # Remove zeros and NAs for better fitting
  clean_data <- data$TOTAL_KG[data$TOTAL_KG > 0 & !is.na(data$TOTAL_KG)]
  
  if(length(clean_data) < 5) {
    return(list(distribution = "empirical", data = clean_data))
  }
  
  # Try to fit different distributions
  tryCatch({
    # Fit lognormal distribution (common for weights)
    fit_lnorm <- fitdist(clean_data, "lnorm")
    
    # Fit gamma distribution (also common for positive continuous data)
    fit_gamma <- fitdist(clean_data, "gamma")
    
    # Fit weibull distribution
    fit_weibull <- fitdist(clean_data, "weibull")
    
    # Compare AICs to find best fit
    aic_values <- c(fit_lnorm$aic, fit_gamma$aic, fit_weibull$aic)
    best_dist <- c("lnorm", "gamma", "weibull")[which.min(aic_values)]
    
    # Return the best distribution and its parameters
    if(best_dist == "lnorm") {
      return(list(distribution = "lnorm", 
                  params = list(meanlog = fit_lnorm$estimate["meanlog"], 
                                sdlog = fit_lnorm$estimate["sdlog"]),
                  fit_object = fit_lnorm))
    } else if(best_dist == "gamma") {
      return(list(distribution = "gamma", 
                  params = list(shape = fit_gamma$estimate["shape"], 
                                rate = fit_gamma$estimate["rate"]),
                  fit_object = fit_gamma))
    } else {
      return(list(distribution = "weibull", 
                  params = list(shape = fit_weibull$estimate["shape"], 
                                scale = fit_weibull$estimate["scale"]),
                  fit_object = fit_weibull))
    }
  }, error = function(e) {
    # If fitting fails, return empirical distribution (the actual data)
    return(list(distribution = "empirical", data = clean_data))
  })
}

# Apply to each group
pathway_distributions <- commodity_dat %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  nest() %>%
  mutate(distribution_model = map(data, fit_distributions))
  
  pathway_distributions  %>%
  dplyr::select(-data)
```
```{r test distribution fits}
library(stats)
test_all_distribution_fits <- function(distribution_models = pathway_distributions) {
  # Get all unique combinations
  combinations <- distribution_models %>%
    dplyr::select(PATHWAY, container, product_pathway, product_subpathway)
  
  # Create empty dataframe to store results
  results <- data.frame()
  
  # Loop through each combination
  for(i in 1:nrow(combinations)) {
    tryCatch({
      # Extract current combination
      current <- combinations[i, ]
      
      # Get distribution model
      dist_model <- distribution_models %>%
        filter(PATHWAY == current$PATHWAY,
               container == current$container,
               product_pathway == current$product_pathway,
               product_subpathway == current$product_subpathway) %>%
        pull(distribution_model) %>%
        .[[1]]
      
      # Skip empirical distributions (no formal tests available)
      if(dist_model$distribution == "empirical") {
        results <- rbind(results, data.frame(
          PATHWAY = current$PATHWAY,
          container = current$container,
          product_pathway = current$product_pathway,
          product_subpathway = current$product_subpathway,
          distribution = "empirical",
          ks_statistic = NA,
          p_value = NA,
          is_good_fit = NA,
          sample_size = length(dist_model$data)
        ))
        next
      }
      
      # Extract data and parameters
      data <- dist_model$fit_object$data
      dist_name <- dist_model$distribution
      
      # Run Kolmogorov-Smirnov test
      if(dist_name == "lnorm") {
        ks_result <- ks.test(data, "plnorm", 
                            meanlog = dist_model$params$meanlog,
                            sdlog = dist_model$params$sdlog)
      } else if(dist_name == "gamma") {
        ks_result <- ks.test(data, "pgamma",
                            shape = dist_model$params$shape,
                            rate = dist_model$params$rate)
      } else if(dist_name == "weibull") {
        ks_result <- ks.test(data, "pweibull",
                            shape = dist_model$params$shape,
                            scale = dist_model$params$scale)
      }
      
      # Add results to dataframe
      results <- rbind(results, data.frame(
        PATHWAY = current$PATHWAY,
        container = current$container,
        product_pathway = current$product_pathway,
        product_subpathway = current$product_subpathway,
        distribution = dist_name,
        ks_statistic = ks_result$statistic,
        p_value = ks_result$p.value,
        is_good_fit = ks_result$p.value > 0.05,
        sample_size = length(data)
      ))
    }, error = function(e) {
      # Handle errors gracefully
      results <- rbind(results, data.frame(
        PATHWAY = current$PATHWAY,
        container = current$container,
        product_pathway = current$product_pathway,
        product_subpathway = current$product_subpathway,
        distribution = "error",
        ks_statistic = NA,
        p_value = NA,
        is_good_fit = FALSE,
        sample_size = NA,
        error = as.character(e)
      ))
    })
    
    # Print progress
    if(i %% 10 == 0) cat("Processed", i, "of", nrow(combinations), "combinations\n")
  }
  
  return(results)
}

# Run the test on all combinations
fit_test_results <- test_all_distribution_fits()

# View summary of results
summary_results <- fit_test_results %>%
  group_by(distribution) %>%
  summarize(
    total = n(),
    good_fits = sum(is_good_fit, na.rm = TRUE),
    good_fit_percentage = mean(is_good_fit, na.rm = TRUE) * 100
  )

print(summary_results)

# Show the combinations with poor fit (p-value < 0.05)
poor_fits <- fit_test_results %>%
  filter(!is_good_fit & !is.na(is_good_fit)) %>%
  arrange(p_value)

print(poor_fits)
```

## save distributions
```{r save distributions}

setwd(here::here("output","item size sampling distributions","aqim"))
# Save the distribution model object
saveRDS(pathway_distributions, paste(params$aqim_qmi_size_distribution_filename, ".rds", sep = "")
)

```




# Example usage of sampling from the distributions

```{r}

# Create diagnostic plots to check the distribution fits
plot_distribution_fit <- function(pathway, container, prod_pathway, prod_subpathway,
                                 distribution_models = pathway_distributions) {
  dist_model <- distribution_models %>%
    filter(PATHWAY == pathway, 
           container == container,
           product_pathway == prod_pathway,
           product_subpathway == prod_subpathway) %>%
    pull(distribution_model) %>%
    .[[1]]
  
  if(dist_model$distribution == "empirical") {
    hist(dist_model$data, main = "Empirical data only (no fit)", xlab = "Weight (kg)")
  } else {
    plot(dist_model$fit_object)
  }
}

# Example usage:
# Check a specific distribution fit
plot_distribution_fit("Land Border - Truck - Conveyance - CBP NB", "baggage", "swine products/by-products", "personal use or consumption")

plot_distribution_fit("Express_courier", "baggage", "swine products/by-products", "smuggled commercial products")

print("warnings are based on low sample sizes. eg smuggled commercials only has 1 example in express courier. In such a case, the empirical distribution is used, and if n=1 it just re-samples that value.")
# Sample from a specific distribution
# sample_weights <- sample_weight(100, "Your_Pathway", "Your_Container", "Your_Prod_Pathway", "Your_Prod_Subpathway")
# hist(sample_weights, main = "Sampled weights", xlab = "Weight (kg)")
```
````


