---
title: "aqim pathway passenger volumes"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  wads_volume_begin_date: "2023-01-01" # Start date for pulling WADS volume data
  wads_volume_end_date: "2024-12-31" # End date for pulling WADS volume data
  wads_volume_reference_filename: "wads volume reference table.csv" # Name of the input reference file that houses which ACTIVITY_CODE is for which PATHWAY
  wads_volume_filename: wads_passenger_volume_by_port_pathway_month_year # Name of the output file for WADS volume data
---

Third Rmd

-   Separate the WADS reference table for a column that identifies if the line item is passenger volume or intercepted material; see input\>data\>wads codes\>get wads codes.R; process those codes (started this process with WADS Lookup Codes.csv but I think I need to include more WADS codes)

    -   WADS for passenger volumes at each port

    -   WADS interceptions at each port

-   Produce a table with each identified combination that calculates:

    -   Port Code

    -   ASF item volume (volume \* rates for each combination)

        -   This is a single calculation - passenger volume at port X the total pathway-combination's rate,

    -   Interceptions of the combination

    -   Entry (volume - interceptions) of the combination

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(here)
library(lubridate)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      wads_volume_begin_date = "2021-01-01",
      wads_volume_end_date = "2024-12-31",
      wads_volume_reference_filename = "wads volume reference table.csv",
      wads_volume_filename = "wads_passenger_volume_by_port_pathway_month_year"
    )
  }
}
```

## gather WADS codes
### Get WADS reference table for volume, inspections, and interceptions
```{r get wads codes}
setwd(here::here("input","data","wads codes"))
wads_ref_table <- read.csv(params$wads_volume_reference_filename) 

wads_vol_ref <- wads_ref_table |> 
  filter(quantification_purpose %in% c("arrival")) |>
  dplyr::select(CODE, PATHWAY)

wads_volume_codes <- wads_vol_ref |> pull(CODE)
# wads_ref_list <- split(wads_ref_table, wads_ref_table$quantification_purpose)
# View(wads_ref_list$volume)

```

### Connect to ARM
```{r set driven}

db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

wads_cols <- c(
  "LOCATION_ID", 
  "LOCATION_NAME", 
  "LOCATION_STATE_CODE", 
  "COUNT", 
  "MAX", 
  "MIN", 
  "ACTIVITY_CODE",
  "WORK_ACTIVITY_DATE"
  )


#-- Pull AQIM inspection records from ARM  ---- 
system.time(wads <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_WADS]")) |> 
              dplyr::select(all_of(wads_cols)) |> 
              #filter to begin and end date
              dplyr::filter(WORK_ACTIVITY_DATE >= params$wads_volume_begin_date & WORK_ACTIVITY_DATE <= params$wads_volume_end_date) |>
              dplyr::filter(ACTIVITY_CODE %in% wads_volume_codes) |>
              collect()
           )
 
```

## Summarize WADS volume by month, year, location, and pathway
```{r summarize wads volume}
# First, adjust count so that each count falls within the min/max range for that activity code
# If COUNT is below MIN, then COUNT = MIN; if COUNT is above MAX, then COUNT = MAX.
#If MIN is NULL, then do not adjust if COUNT is below MIN.
#If MAX is NULL, then do not adjust if COUNT is above MAX.
#If both MIN and MAX are NULL, then do not adjust COUNT.
wads <- wads |> 
  mutate(COUNT = case_when(
    !is.na(MIN) & COUNT < MIN ~ MIN,
    !is.na(MAX) & COUNT > MAX ~ MAX,
    TRUE ~ COUNT
  )) |> dplyr::select(-c(MIN, MAX))

wads_volume <- wads |> 
  left_join(wads_vol_ref, by = c("ACTIVITY_CODE" = "CODE")) |> 
  mutate(year = lubridate::year(WORK_ACTIVITY_DATE),
         month = lubridate::month(WORK_ACTIVITY_DATE)) |> 
         #reduce to only relevant pathways
  filter(!(is.na(PATHWAY) | PATHWAY == "")) |>
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, year, month) |> 
  summarize(COUNT = sum(COUNT, na.rm=TRUE),
            .groups = "drop")

```

```{r define border state codes}
NB_state_cd <- c("AK","ID","ME","MI","MN","MT","NH","NY","ND","OH","PA","VT","WA")
SB_state_cd <- c("CA","AZ","NM","TX")
```
```{r separate land border to NB-SB}
wads_volume <- wads_volume %>%
  mutate(
    PATHWAY = case_when(
      LOCATION_STATE_CODE %in% NB_state_cd & grepl("Land Border|Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " NB"),
      LOCATION_STATE_CODE %in% SB_state_cd & grepl("Land Border|Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " SB"),
      TRUE ~ PATHWAY
    )
  )
```
## Save summarized WADS volume
```{r save wads volume}
setwd(here::here("output","wads","arrivals_or_entries"))
saveRDS(wads_volume, paste(params$wads_volume_filename,".rds"))
```

