---
title: "aqim pathway risk by PoE"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
 
  wads_volume_filename: "wads_passenger_volume_by_port_pathway_month_year" # Name of the output file for WADS volume data
  glm_results_filename: "glmm_results_aqim fraction at risk by pathway and location"
  bootstrap_results_filename: "bootstrap_results_aqim fraction at risk by pathway and location"
  aqim_qmi_size_distribution_filename: "pathway_weight_distribution_models" # Name of the output file for the distribution models of the weight of items observed in the AQIM pathways
  ---

```{r setup, include=FALSE}
pacman::p_load(tidyverse)

```
```{r params for vscode}
if(!exists("params")) {
  params <- list(
    wads_volume_filename = "wads_passenger_volume_by_port_pathway_month_year",
    glm_results_filename = "glmm_results_aqim fraction at risk by pathway and location",
    bootstrap_results_filename = "bootstrap_results_aqim fraction at risk by pathway and location",
    aqim_qmi_size_distribution_filename = "pathway_weight_distribution_models" # Name of the output file for the distribution models
  )
}
```

# Read WADS volume data
```{r read wads volume data}
setwd(here::here("output","wads","arrivals_or_entries"))
wads_volume <- readRDS(paste(params$wads_volume_filename, ".rds"))
```
# read approach rates
```{r read approach rates}
setwd(here::here('output','aqim pathway approach rates')) 
location_probabilities_glmm <- readRDS(paste(params$glm_results_filename,".rds", sep = " "))
bootstrap_results <- readRDS(paste(params$bootstrap_results_filename,".rds", sep = " "))
```
# Join WADS volume and approach rates

```{r join wads volume and approach rates}
#for every entry in wads_volume where PATHWAY exists in location_probabilities_glmm, then all possible `group` values in location_probabilities_glmm of that PATHWAY can exist. 


# connect wads_volume and location_probabilities_glmm using PATHWAY and INSPECTION_LOCATION_NAME = LOCATION_NAME, and group
wads_volume_joined <- wads_volume |> 
  left_join(location_probabilities_glmm, by = c("PATHWAY" = 'pathway', "LOCATION_NAME" = "location"), relationship = "many-to-many")
View(wads_volume_joined)

back_join_test <- location_probabilities_glmm |> 
  left_join(wads_volume, by = c("pathway" = 'PATHWAY', "location" = "LOCATION_NAME"), relationship = "many-to-many")
View(back_join_test)

print("I've fixed everything so far except that the location_name is AQAS is not the same as that of ARM, so all that location-based predictions were for nothing...")
print("Think about whether location can just be done by state, maybe?")

#Split the group column into its component parts
test <- wads_volume_joined %>%
  separate(group, 
           into = c("PATHWAY_check", "container", "product_pathway", "product_subpathway"),
           sep = ";",
           remove = FALSE)
View(test)
test2 <- test %>%  # Keep original group column
  mutate(
    # Verify PATHWAY matches with split value
    pathway_match = PATHWAY == PATHWAY_check
  ) %>%
  dplyr::select(-PATHWAY_check)  # Remove the redundant PATHWAY column
View(test2)
# If a location_name is not in the location_probabilities_glmm, then if pathway is not NA and  the probability is not known (i.e., the LOCATION_NAME was not in the model parameters), then the probability should equivalent to the pathway when LOCATION_NAME == "Average"
#the significant column is whether it is different from the baseline, which is not relevant
wads_volume_joined <- wads_volume_joined |> rowwise() |>
  mutate(probability = ifelse(!is.na(PATHWAY) & is.na(probability), 
                              location_probabilities_glmm$probability[location_probabilities_glmm$location == "Average" & location_probabilities_glmm$pathway == PATHWAY], 
                              probability))
#View(wads_volume_joined)

# wads_volume_joined |> 
#   mutate(probability = ifelse(is.na(probability), 0, probability)) # replace NA probabilities with 0

```
# retrieve pathway weight distribution models
```{r retrieve pathway weight distribution models}
setwd(here::here("output","item size sampling distributions","aqim"))
pathway_distributions <- readRDS(paste(params$aqim_qmi_size_distribution_filename, ".rds", sep = ""))
```



# Sample weights from the distribution models
skip this one for the next section
```{r}
   #Create a function to sample from these distributions
sample_weight <- function(n, pathway, container, prod_pathway, prod_subpathway, 
                          distribution_models = pathway_distributions) {
  
  # Find the matching distribution
  dist_model <- distribution_models %>%
    filter(PATHWAY == pathway, 
           container == container,
           product_pathway == prod_pathway,
           product_subpathway == prod_subpathway) %>%
    pull(distribution_model) %>%
    .[[1]]
  
  if(is.null(dist_model)) {
    warning("No distribution found for the specified combination. Returning NA.")
    return(rep(NA, n))
  }
  
  # Sample from the distribution
  if(dist_model$distribution == "empirical") {
    # If empirical, sample from the actual data with replacement
    return(sample(dist_model$data, size = n, replace = TRUE))
  } else if(dist_model$distribution == "lnorm") {
    return(rlnorm(n, meanlog = dist_model$params$meanlog, sdlog = dist_model$params$sdlog))
  } else if(dist_model$distribution == "gamma") {
    return(rgamma(n, shape = dist_model$params$shape, rate = dist_model$params$rate))
  } else if(dist_model$distribution == "weibull") {
    return(rweibull(n, shape = dist_model$params$shape, scale = dist_model$params$scale))
  }
}

sample_weight(10, "Express_courier", "baggage", "swine products/by-products", "personal use or consumption")


```








```{r try again}
str(pathway_distributions, max.level = 2)

# Calculate total kg at risk for each row in wads_volume_joined
risk_weights_by_pathway_tmp <- wads_volume_joined %>%
  # Process each row individually
  rowwise() %>%
  mutate(
    # Calculate number of events with risk (rounded to integer)
    events_with_risk = round(probability * COUNT, 0)
  )

risk_weights_by_pathway <- risk_weights_by_pathway_tmp %>%
  rowwise() %>%
  mutate(
    # Use the sample_weight function with column values as arguments
    sampled_weights = if(!is.na(events_with_risk) && events_with_risk > 0 && !is.na(PATHWAY)) {
      # Sample weights based on the current row's values
      # Make sure these column names match exactly what's in your dataframe
      list(sample_weight(
        n = events_with_risk, 
        pathway = PATHWAY,
        container = container,
        prod_pathway = product_pathway,
        prod_subpathway = product_subpathway
      ))
    } else {
      list(numeric(0))  # Empty list if no events with risk
    },
    # Sum the sampled weights to get total kg at risk
    total_kg_at_risk = if(length(sampled_weights) > 0) sum(sampled_weights, na.rm = TRUE) else 0
  ) %>%
  ungroup()



```




```{r sample events from distriution models of weight per event}
risk_weights_by_pathway <- risk_weights_by_pathway_tmp %>%
  # Add mutate() here
  mutate(
    total_kg_at_risk = case_when(
      # Only process if we have events with risk and can find the pathway
      events_with_risk > 0 & !is.na(PATHWAY) ~ {
        # Rest of your case_when logic remains the same
        dist_model <- pathway_distributions %>%
          filter(PATHWAY == .data$PATHWAY,
          container == .data$container,
          product_pathway == .data$product_pathway,
          product_subpathway == .data$product_subpathway) %>%
          pull(distribution_model) %>%
          first()

#test
pathway_distributions %>%
          filter(PATHWAY == 'Express_courier',
          container == 'baggage',
          product_pathway == .data$'swine products/by-products',
          product_subpathway == 'personal use or consumption') %>%
          pull(distribution_model) %>%
          first()

#end test

        # If we found a distribution model, sample from it
        if(!is.null(dist_model) && length(dist_model) > 0) {
          # Sample weights based on the distribution model
          sampled_weights <- if(dist_model$distribution == "empirical") {
            sample(dist_model$data, size = events_with_risk, replace = TRUE)
          } else if(dist_model$distribution == "lnorm") {
            rlnorm(events_with_risk, 
                   meanlog = dist_model$params$meanlog, 
                   sdlog = dist_model$params$sdlog)
          } else if(dist_model$distribution == "gamma") {
            rgamma(events_with_risk, 
                   shape = dist_model$params$shape, 
                   rate = dist_model$params$rate)
          } else if(dist_model$distribution == "weibull") {
            rweibull(events_with_risk, 
                     shape = dist_model$params$shape, 
                     scale = dist_model$params$scale)
          } else {
            # Default case if distribution type unknown
            rep(NA_real_, events_with_risk)
          }
          
          # Sum the sampled weights
          sum(sampled_weights, na.rm = TRUE)
        } else {
          NA_real_  # No distribution model found
        }
      },
      TRUE ~ 0  # No events with risk
    )
  ) %>%
  ungroup()

# View the results
summary(risk_weights_by_pathway$total_kg_at_risk)

```