---
title: "aqim pathway volume per incident"
author: "Dalon White"
date: "2025-08-15"
output: html_document

params:
  risk_quantification: "0_1" # 0_1 or "range"
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  aqim_qmi_size_distribution_filename: "pathway_weight_distribution_models" # Name of the output file for the distribution models
---
Run in conjunction with aqim pathway passenger layers.Rmd and aqim pathway approach rates.Rmd

```{r setup, include=FALSE}
library(tidyverse)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      risk_quantification = "0_1", # 0_1 or "range"
      affected_countries = "current", # "current" or "mex_can" 
      years = 2024:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"),
      aqim_qmi_size_distribution_filename = "pathway_weight_distribution_models"
    )
  }
}
```

# Determine if for ASF countries or mex/can

```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
if(params$affected_countries == "current") {
  affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
} else if (params$affected_countries == "mex_can") {
  affected_countries <- c("MEX", "CAN")
} else if (params$affected_countries == "other") {
  affected_countries <- params$other_affected_countries
} else {
  stop("Invalid value for params$affected_countries")
}


# setwd(here::here("input","data","asf affected countries by year"))
# affected_countries <- read.csv("asf affected countries.csv") |> 
#     mutate(Year = as.numeric(Year)) |> 
#     filter(Year %in% params$years) |> 
#     filter(Disease == "ASF") |>
#     distinct(ISO3_CODE) |> 
#     pull(ISO3_CODE)

# affected_countries
```

# Read and prep data
Read prepared data
Isolate to ASF at risk commodities
Identify the distribution of sizes by pathway and subpathway, entry_type, and container

```{r}

dat <- read.csv("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv")

```

# identify which commodities are at risk for ASF

```{r assign risk}
  # filter to only those with IS_ASF_RISK == "YES"
# Assign risk
commodity_dat_tmp <- dat %>% mutate(IS_ASF_RISK = case_when(
  edge_origin %in% affected_countries | product_origin %in% affected_countries ~ IS_ASF_RISK,
  .default = "NO"
         )
)

  # filter to only those with IS_ASF_RISK == "YES"
# keep pathway, location, commodity, pathway, subpathway, entry_type, container, and TOTAL_KG
commodity_dat <- commodity_dat_tmp %>%
  filter(!IS_ASF_RISK == "NO") %>%
  dplyr::select(
    ID, 
    PATHWAY, 
    PORT_CD,
    CALENDAR_MONTH,
    CALENDAR_YEAR,
    INSPECTION_LOCATION_ID, 
    INSPECTION_LOCATION_NAME, 
    COMMODITY_DISPLAY_NAME, 
    entry_type, 
    container, 
    TOTAL_KG,
    product_pathway,
    product_subpathway
  )
print(paste(nrow(commodity_dat |> filter(is.na(TOTAL_KG))), "entries with NA TOTAL_KG. It is being removed."))


print(
    paste(nrow(commodity_dat_tmp %>%
  filter(!IS_ASF_RISK == "NO")),
  "records with ASF risk")
)

print("These numbers drop off significantly when you filter to only those countries with ASF")

View(dat |> filter(IS_ASF_RISK != "NO") |> count(edge_origin, product_origin) |> arrange(desc(n)))

print("Most items that could carry risk are coming from Mex, Can, or the US; in 4th and 5th, though, are ESP, BHS, PRI (Puerto Rico), with ITA, DEU, and PHL not far behind. These all actually carry risk")

write.csv(dat |> filter(IS_ASF_RISK != "NO") |> count(edge_origin, product_origin) |> arrange(desc(n)),
          here::here("output","report files","commodity origin counts for risk.csv"),
          row.names=FALSE)

View(commodity_dat_tmp |> filter(IS_ASF_RISK == "NO") |> count(COMMODITY_DISPLAY_NAME, sort=TRUE))
print("I double checked all these commodities to ensure none of these names carried risk, and that the reference tables are working appropriately. They are, and these commodities do not carry risk.")
```

# get the distribution of each

```{r fit distributions to item weights}
if (!requireNamespace("fitdistrplus", quietly = TRUE)) {
  install.packages("fitdistrplus")
}
library(fitdistrplus)
library(dplyr)

# Create a function to fit distributions to each group
fit_distributions <- function(data) {
  # Remove zeros and NAs for better fitting
  clean_data <- data$TOTAL_KG[data$TOTAL_KG > 0 & !is.na(data$TOTAL_KG)]
  
  if(length(clean_data) < 5) {
    return(list(distribution = "empirical", data = clean_data))
  }
  
  # Try to fit different distributions
  tryCatch({
    # Fit lognormal distribution (common for weights)
    fit_lnorm <- fitdist(clean_data, "lnorm")
    
    # Fit gamma distribution (also common for positive continuous data)
    fit_gamma <- fitdist(clean_data, "gamma")
    
    # Fit weibull distribution
    fit_weibull <- fitdist(clean_data, "weibull")
    
    # Compare AICs to find best fit
    aic_values <- c(fit_lnorm$aic, fit_gamma$aic, fit_weibull$aic)
    best_dist <- c("lnorm", "gamma", "weibull")[which.min(aic_values)]
    
    # Return the best distribution and its parameters
    if(best_dist == "lnorm") {
      return(list(distribution = "lnorm", 
                  params = list(meanlog = fit_lnorm$estimate["meanlog"], 
                                sdlog = fit_lnorm$estimate["sdlog"]),
                  fit_object = fit_lnorm))
    } else if(best_dist == "gamma") {
      return(list(distribution = "gamma", 
                  params = list(shape = fit_gamma$estimate["shape"], 
                                rate = fit_gamma$estimate["rate"]),
                  fit_object = fit_gamma))
    } else {
      return(list(distribution = "weibull", 
                  params = list(shape = fit_weibull$estimate["shape"], 
                                scale = fit_weibull$estimate["scale"]),
                  fit_object = fit_weibull))
    }
  }, error = function(e) {
    # If fitting fails, return empirical distribution (the actual data)
    return(list(distribution = "empirical", data = clean_data))
  })
}

# Apply to each group
pathway_distributions <- commodity_dat %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  nest() %>%
  mutate(distribution_model = map(data, fit_distributions))
  
  pathway_distributions  %>%
  dplyr::select(-data)
```
```{r test distribution fits}
library(stats)
test_all_distribution_fits <- function(distribution_models = pathway_distributions) {
  # Get all unique combinations
  combinations <- distribution_models %>%
    dplyr::select(PATHWAY, container, product_pathway, product_subpathway)
  
  # Create empty dataframe to store results
  results <- data.frame()
  
  # Loop through each combination
  for(i in 1:nrow(combinations)) {
    tryCatch({
      # Extract current combination
      current <- combinations[i, ]
      
      # Get distribution model
      dist_model <- distribution_models %>%
        filter(PATHWAY == current$PATHWAY,
               container == current$container,
               product_pathway == current$product_pathway,
               product_subpathway == current$product_subpathway) %>%
        pull(distribution_model) %>%
        .[[1]]
      
      # Skip empirical distributions (no formal tests available)
      if(dist_model$distribution == "empirical") {
        results <- rbind(results, data.frame(
          PATHWAY = current$PATHWAY,
          container = current$container,
          product_pathway = current$product_pathway,
          product_subpathway = current$product_subpathway,
          distribution = "empirical",
          ks_statistic = NA,
          p_value = NA,
          is_good_fit = NA,
          sample_size = length(dist_model$data)
        ))
        next
      }
      
      # Extract data and parameters
      data <- dist_model$fit_object$data
      dist_name <- dist_model$distribution
      
      # Run Kolmogorov-Smirnov test
      if(dist_name == "lnorm") {
        ks_result <- ks.test(data, "plnorm", 
                            meanlog = dist_model$params$meanlog,
                            sdlog = dist_model$params$sdlog)
      } else if(dist_name == "gamma") {
        ks_result <- ks.test(data, "pgamma",
                            shape = dist_model$params$shape,
                            rate = dist_model$params$rate)
      } else if(dist_name == "weibull") {
        ks_result <- ks.test(data, "pweibull",
                            shape = dist_model$params$shape,
                            scale = dist_model$params$scale)
      }
      
      # Add results to dataframe
      results <- rbind(results, data.frame(
        PATHWAY = current$PATHWAY,
        container = current$container,
        product_pathway = current$product_pathway,
        product_subpathway = current$product_subpathway,
        distribution = dist_name,
        ks_statistic = ks_result$statistic,
        p_value = ks_result$p.value,
        is_good_fit = ks_result$p.value > 0.05,
        sample_size = length(data)
      ))
    }, error = function(e) {
      # Handle errors gracefully
      results <- rbind(results, data.frame(
        PATHWAY = current$PATHWAY,
        container = current$container,
        product_pathway = current$product_pathway,
        product_subpathway = current$product_subpathway,
        distribution = "error",
        ks_statistic = NA,
        p_value = NA,
        is_good_fit = FALSE,
        sample_size = NA,
        error = as.character(e)
      ))
    })
    
    # Print progress
    if(i %% 10 == 0) cat("Processed", i, "of", nrow(combinations), "combinations\n")
  }
  
  return(results)
}

# Run the test on all combinations
fit_test_results <- test_all_distribution_fits()

# View summary of results
summary_results <- fit_test_results %>%
  group_by(distribution) %>%
  summarize(
    total = n(),
    good_fits = sum(is_good_fit, na.rm = TRUE),
    good_fit_percentage = mean(is_good_fit, na.rm = TRUE) * 100
  )

print(summary_results)

# Show the combinations with poor fit (p-value < 0.05)
poor_fits <- fit_test_results %>%
  filter(!is_good_fit & !is.na(is_good_fit)) %>%
  arrange(p_value)

print(poor_fits)
```

## save distributions
```{r save distributions}

setwd(here::here("output","item size sampling distributions","aqim"))
# Save the distribution model object
saveRDS(pathway_distributions, paste(params$aqim_qmi_size_distribution_filename, ".rds", sep = ""))
```

# Summarize and visualize the pathway distributions

```{r summarize_distribution_models}
# Create a function to extract key information from each distribution model
extract_distribution_summary <- function(dist_model, data) {
  # Initialize result with NA values
  result <- list(
    distribution = NA_character_,
    mean = NA_real_, 
    median = NA_real_,
    sd = NA_real_,
    min = NA_real_,
    max = NA_real_,
    q25 = NA_real_,
    q75 = NA_real_,
    n_samples = NA_integer_,
    goodness_of_fit = NA_character_,
    aic = NA_real_,
    p_value = NA_real_
  )
  
  # Extract the original data
  original_data <- data$TOTAL_KG[!is.na(data$TOTAL_KG) & data$TOTAL_KG > 0]
  
  # Set basic summary statistics from the original data
  result$n_samples <- length(original_data)
  if(result$n_samples > 0) {
    result$min <- min(original_data)
    result$max <- max(original_data)
    result$q25 <- quantile(original_data, 0.25)
    result$q75 <- quantile(original_data, 0.75)
  }
  
  # If the model is empirical, just return summary statistics
  if(dist_model$distribution == "empirical") {
    result$distribution <- "empirical"
    if(result$n_samples > 0) {
      result$mean <- mean(original_data)
      result$median <- median(original_data)
      result$sd <- sd(original_data)
    }
    result$goodness_of_fit <- "N/A (empirical)"
    return(result)
  }
  
  # Extract distribution type and parameters
  result$distribution <- dist_model$distribution
  
  # Calculate mean, median, and SD based on the distribution
  if(dist_model$distribution == "lnorm") {
    meanlog <- dist_model$params$meanlog
    sdlog <- dist_model$params$sdlog
    result$mean <- exp(meanlog + sdlog^2/2)
    result$median <- exp(meanlog)
    result$sd <- sqrt((exp(sdlog^2) - 1) * exp(2*meanlog + sdlog^2))
  } else if(dist_model$distribution == "gamma") {
    shape <- dist_model$params$shape
    rate <- dist_model$params$rate
    result$mean <- shape / rate
    result$median <- qgamma(0.5, shape = shape, rate = rate)
    result$sd <- sqrt(shape / rate^2)
  } else if(dist_model$distribution == "weibull") {
    shape <- dist_model$params$shape
    scale <- dist_model$params$scale
    result$mean <- scale * gamma(1 + 1/shape)
    result$median <- scale * (log(2))^(1/shape)
    result$sd <- scale * sqrt(gamma(1 + 2/shape) - (gamma(1 + 1/shape))^2)
  }
  
  # Extract goodness of fit information
  if(!is.null(dist_model$fit_object)) {
    result$aic <- dist_model$fit_object$aic
    
    # Perform KS test for goodness of fit
    if(dist_model$distribution == "lnorm") {
      ks_test <- ks.test(original_data, "plnorm", 
                          meanlog = dist_model$params$meanlog,
                          sdlog = dist_model$params$sdlog)
    } else if(dist_model$distribution == "gamma") {
      ks_test <- ks.test(original_data, "pgamma",
                          shape = dist_model$params$shape,
                          rate = dist_model$params$rate)
    } else if(dist_model$distribution == "weibull") {
      ks_test <- ks.test(original_data, "pweibull",
                          shape = dist_model$params$shape,
                          scale = dist_model$params$scale)
    }
    
    result$p_value <- ks_test$p.value
    result$goodness_of_fit <- ifelse(result$p_value > 0.05, "Good fit", "Poor fit")
  }
  
  return(result)
}

# Apply the function to each group
distribution_summary <- pathway_distributions %>%
  mutate(summary = pmap(list(distribution_model, data), extract_distribution_summary)) %>%
  unnest_wider(summary)

# Create a more readable summary table
readable_summary <- distribution_summary %>%
  dplyr::select(PATHWAY, container, product_pathway, product_subpathway, 
         distribution, mean, median, sd, min, max, n_samples, goodness_of_fit, p_value) %>%
  # Format numbers to be more readable
  mutate(across(c(mean, median, sd, min, max), ~round(., 2)),
         across(p_value, ~round(., 4)),
         across(n_samples, ~as.integer(.)))

# View the summary table
head(readable_summary, 10)

# Save the summary table
setwd(here::here("output", "item size sampling distributions", "aqim"))
write.csv(readable_summary, "pathway_distribution_summary.csv", row.names = FALSE)

# Print overall statistics
distribution_counts <- readable_summary %>%
  count(distribution) %>%
  mutate(percentage = n / sum(n) * 100)

goodness_counts <- readable_summary %>%
  filter(distribution != "empirical") %>%
  count(goodness_of_fit) %>%
  mutate(percentage = n / sum(n) * 100)

cat("Distribution Model Counts:\n")
print(distribution_counts)

cat("\nGoodness of Fit Summary (excluding empirical):\n")
print(goodness_counts)

cat("\nOverall Summary Statistics:\n")
print(summary(readable_summary[c("mean", "median", "sd", "n_samples")]))
```

```{r visualize_distribution_models}
# Create a function to generate a plot for a specific combination
create_distribution_plot <- function(pathway, container, prod_pathway, prod_subpathway,
                                    distribution_models = pathway_distributions) {
  # Extract the specific model
  row_data <- distribution_models %>%
    filter(PATHWAY == pathway, 
           container == container,
           product_pathway == prod_pathway,
           product_subpathway == prod_subpathway)
  
  if(nrow(row_data) == 0) {
    return(ggplot() + 
             annotate("text", x = 0, y = 0, label = "No data found for this combination") +
             theme_minimal())
  }
  
  dist_model <- row_data$distribution_model[[1]]
  data <- row_data$data[[1]]
  original_data <- data$TOTAL_KG[!is.na(data$TOTAL_KG) & data$TOTAL_KG > 0]
  
  if(length(original_data) == 0) {
    return(ggplot() + 
             annotate("text", x = 0, y = 0, label = "No valid data points") +
             theme_minimal())
  }
  
  # Prepare the plot data
  max_x <- max(original_data) * 1.5
  x_vals <- seq(0, max_x, length.out = 1000)
  
  # Create data frame for histogram
  hist_data <- data.frame(x = original_data)
  
  # Initialize plot
  p <- ggplot(hist_data, aes(x = x)) +
    geom_histogram(aes(y = ..density..), binwidth = max_x/30, fill = "lightblue", color = "black", alpha = 0.7) +
    labs(title = paste("Distribution for:", pathway),
         subtitle = paste("Container:", container, "| Pathway:", prod_pathway, "| Subpathway:", prod_subpathway),
         x = "Weight (kg)",
         y = "Density") +
    theme_minimal() +
    theme(plot.title = element_text(size = 10),
          plot.subtitle = element_text(size = 8))
  
  # Add the fitted distribution curve if available
  if(dist_model$distribution != "empirical") {
    # Create a data frame for the fitted line
    if(dist_model$distribution == "lnorm") {
      y_vals <- dlnorm(x_vals, meanlog = dist_model$params$meanlog, sdlog = dist_model$params$sdlog)
      fit_label <- sprintf("Lognormal (μ=%.2f, σ=%.2f)", dist_model$params$meanlog, dist_model$params$sdlog)
    } else if(dist_model$distribution == "gamma") {
      y_vals <- dgamma(x_vals, shape = dist_model$params$shape, rate = dist_model$params$rate)
      fit_label <- sprintf("Gamma (shape=%.2f, rate=%.2f)", dist_model$params$shape, dist_model$params$rate)
    } else if(dist_model$distribution == "weibull") {
      y_vals <- dweibull(x_vals, shape = dist_model$params$shape, scale = dist_model$params$scale)
      fit_label <- sprintf("Weibull (shape=%.2f, scale=%.2f)", dist_model$params$shape, dist_model$params$scale)
    }
    
    line_data <- data.frame(x = x_vals, y = y_vals)
    
    # Add the line to the plot
    p <- p + 
      geom_line(data = line_data, aes(x = x, y = y), color = "red", size = 1) +
      annotate("text", x = max_x * 0.7, y = max(y_vals) * 0.8, 
               label = fit_label, hjust = 0, size = 3, color = "red")  } else {
    # Handle case where there's only 1 data point
    if(length(original_data) == 1) {
      p <- p + 
        annotate("text", x = max_x * 0.7, y = 0.5, 
                 label = "Empirical distribution (single point)", hjust = 0, size = 3, color = "red")
    } else {
      p <- p + 
        annotate("text", x = max_x * 0.7, y = 0.9 * max(density(original_data)$y), 
                 label = "Empirical distribution (no fit)", hjust = 0, size = 3, color = "red")
    }
  }
  
  # Add summary statistics
  summary_stats <- extract_distribution_summary(dist_model, data)
  stats_text <- sprintf(
    "n=%d | Mean=%.2f | Median=%.2f | SD=%.2f", 
    summary_stats$n_samples, summary_stats$mean, summary_stats$median, summary_stats$sd
  )
  
  if(summary_stats$distribution != "empirical") {
    stats_text <- paste0(stats_text, sprintf(" | p-value=%.4f", summary_stats$p_value))
  }
  
  p <- p + 
    annotate("text", x = max_x * 0.5, y = 0, 
             label = stats_text, hjust = 0.5, vjust = 0, size = 3)
  
  return(p)
}

# Identify the top combinations by sample size
top_combinations <- readable_summary %>%
  arrange(desc(n_samples)) %>%
  head(12)

# Create a directory for plots if it doesn't exist
plot_dir <- here::here("output", "item size sampling distributions", "aqim", "plots")
if(!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)

# Create plots for the top combinations
library(gridExtra)
plot_list <- list()

for(i in 1:nrow(top_combinations)) {
  row <- top_combinations[i, ]
  plot_list[[i]] <- create_distribution_plot(
    pathway = row$PATHWAY,
    container = row$container,
    prod_pathway = row$product_pathway,
    prod_subpathway = row$product_subpathway
  )
}

# Create a multi-panel plot for the top combinations
combined_plot <- gridExtra::grid.arrange(grobs = plot_list, ncol = 3)

# Save the combined plot
ggsave(file.path(plot_dir, "top_combinations_plot.png"), 
       combined_plot, width = 15, height = 10, dpi = 300)

# Create a function to generate a report for each model type
create_distribution_report <- function(dist_type) {
  # Filter combinations by distribution type
  filtered_combs <- readable_summary %>%
    filter(distribution == dist_type) %>%
    arrange(desc(n_samples))
  
  if(nrow(filtered_combs) == 0) {
    return(NULL)
  }
  
  # Take the top 6 by sample size
  top_combs <- head(filtered_combs, 6)
  
  # Create individual plots
  plot_list <- list()
  for(i in 1:nrow(top_combs)) {
    row <- top_combs[i, ]
    plot_list[[i]] <- create_distribution_plot(
      pathway = row$PATHWAY,
      container = row$container,
      prod_pathway = row$product_pathway,
      prod_subpathway = row$product_subpathway
    )
  }
  
  # Combine plots
  if(length(plot_list) > 0) {
    combined_plot <- gridExtra::grid.arrange(grobs = plot_list, ncol = 2)
    
    # Save the combined plot
    filename <- paste0("distribution_type_", dist_type, "_examples.png")
    ggsave(file.path(plot_dir, filename), combined_plot, width = 10, height = 8, dpi = 300)
  }
}

# Create reports for each distribution type
dist_types <- unique(readable_summary$distribution)
for(dist_type in dist_types) {
  create_distribution_report(dist_type)
}

# Create a summary of mean weights by pathway and container
pathway_container_summary <- pathway_distributions %>%
  unnest(data) %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    mean_weight = mean(TOTAL_KG, na.rm = TRUE),
    median_weight = median(TOTAL_KG, na.rm = TRUE),
    total_samples = n(),
    min_weight = min(TOTAL_KG, na.rm = TRUE),
    max_weight = max(TOTAL_KG, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_samples))

# Save this summary
write.csv(pathway_container_summary, 
          file.path(here::here("output", "item size sampling distributions", "aqim"), 
                   "pathway_container_weight_summary.csv"), 
          row.names = FALSE)

# Plot summary by pathway with both mean and median
pathway_summary <- pathway_container_summary %>%
  group_by(PATHWAY) %>%
  summarize(
    mean_weight = weighted.mean(mean_weight, total_samples, na.rm = TRUE),
    median_weight = median(median_weight, na.rm = TRUE),
    total_samples = sum(total_samples),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_weight))

# Create a combined plot showing both mean and median
mean_median_plot <- pathway_summary %>%
  pivot_longer(
    cols = c(mean_weight, median_weight),
    names_to = "statistic",
    values_to = "weight"
  ) %>%
  mutate(
    statistic = ifelse(statistic == "mean_weight", "Mean", "Median"),
    PATHWAY = factor(PATHWAY, levels = pathway_summary$PATHWAY)
  ) %>%
  ggplot(aes(x = PATHWAY, y = log1p(weight), fill = statistic)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(aes(label = sprintf("%.1f kg", weight)), 
            position = position_dodge(width = 0.9), 
            hjust = -0.1, 
            size = 3, 
            angle = 90) +
  coord_flip() +
  labs(title = "Average Weight of ASF-Risk Commodities by Pathway",
       subtitle = "Comparison of mean and median weights",
       x = "Pathway",
       y = "Weight (kg)",
       fill = "Statistic") +
  theme_minimal()

ggsave(file.path(plot_dir, "average_weight_by_pathway_mean_median.png"), 
       mean_median_plot, width = 12, height = 8, dpi = 300)


```