---
title: "AQIM Volume Spatial Layers"
author: "Dalon White"
date: "2025-08-28"
output: html_document

params:
  # Set number of simulations
  n_sims_for_volume: 100
  pathway_volumes_filename: "volume_of_risk_by_pathway" # Name of the output file for the pathway risk summary results
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: !r c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify countries
  # Years to consider for affected countries
  years: !r 2023:2024
---

<!--
This script creates spatial layers for ASF risk assessment based on AQIM data.
It processes the input data to:
1. Load pathway risk summary data
2. Add geolocation information from various sources
3. Convert to spatial objects
4. Create individual spatial layers by pathway combination
5. Generate final destination risk layers
6. Save outputs as GeoPackages (.gpkg) for Shiny app compatibility
-->

```{r setup, include=FALSE}
# Load required packages
library(tidyverse) # For dplyr, ggplot2, etc.
library(magrittr)  # For %>% pipe
library(lubridate) # For date handling
library(readr)     # For data reading
library(stringdist) # For string matching
library(purrr)     # For functional programming
library(DBI)       # For database connections
library(odbc)      # For ODBC database connections
library(sf)        # For spatial data handling
library(tigris)    # For geographic boundaries
library(SpatialKDE) # For kernel density estimation
library(here)      # For project-relative paths
# Conditionally load tidygeocoder if we need it
if(requireNamespace("tidygeocoder", quietly = TRUE)) {
  library(tidygeocoder) # For geocoding
}
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      n_sims_for_volume = 100,
      affected_countries = "current", 
      other_affected_countries = c("Country1 ISO-3 Code", "Country2 ISO-3 Code"),
      pathway_volumes_filename = "volume_of_risk_by_pathway"
    )
  }
}

setwd(here::here('output','aqim based volume of at risk product approaching'))
# Load pathway risk summary data
pathway_risk_summary <- readRDS(paste(params$pathway_volumes_filename, min(params$years), "to", max(params$years), params$affected_countries, params$n_sims_for_volume, "sims",".rds", sep = "_"))

```

# 1. Gather Geolocation Data

```{r geolocation-database-connect}
# Connect to ARM and GeoLocations databases
db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=PPQ_AQI_ARMDMV2;
                      trusted_connection=yes")

geo_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=GeoLocations;
                      trusted_connection=yes")

# Define columns to extract from each table
brg_loc_cols <- c("ID", "NAME", "SITE", "PHYSICAL_ADDRESS_ID", 
                 "MAIL_ADDRESS_ID", 'CBP_PORT_NUMBER', "IATA_CODE")
cbp_poe_cols <- c("CBP_NAME", "CBP_CODE", "LAT", "LON", "Port of Entry Name")
ppq_facilities_cols <- c("LocationName", "PPQCode", "City", "State", "Longitude", "Latitude")
```

# 2. Gather Coordinates From Different Sources

```{r gather-coordinates}
# Get location data from BRG2_LOCATION table
brg_loc <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_LOCATION]")) %>% 
           dplyr::select(all_of(brg_loc_cols)) %>% 
           collect()

# Get CBP Port of Entry data with coordinates
cbp_poe <- tbl(geo_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_CBP_POEs]")) %>% 
           dplyr::select(all_of(cbp_poe_cols)) %>% 
           collect()

# Get PPQ facilities data with coordinates
ppq_facilities <- tbl(db_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_PPQ_Facilities]")) %>% 
                 dplyr::select(all_of(ppq_facilities_cols)) %>% 
                 collect()

# Join pathway risk data with location reference
pathway_risk_brg_loc <- pathway_risk_summary %>% 
                       left_join(brg_loc, by = c("LOCATION_ID" = "ID"))

# Step 1: Join with CBP POE data first (preferred source)
pathway_risk_geo <- pathway_risk_brg_loc %>%
  left_join(cbp_poe, by = c("CBP_PORT_NUMBER" = "CBP_CODE")) %>%
  rename(
    CBP_LAT = LAT,
    CBP_LON = LON,
    CBP_PORT_NAME = `Port of Entry Name`
  )

# Step 2: Join with PPQ facilities for rows still missing coordinates
pathway_risk_geo <- pathway_risk_geo %>%
  left_join(
    ppq_facilities %>% 
      dplyr::select(LocationName, Latitude, Longitude) %>% 
      rename(
        PPQ_LAT = Latitude,
        PPQ_LON = Longitude
      ),
    by = c("LOCATION_NAME" = "LocationName")
  )

# Step 3: Create combined coordinate columns, prioritizing CBP data
pathway_risk_geo <- pathway_risk_geo %>%
  mutate(
    final_LAT = coalesce(CBP_LAT, PPQ_LAT),
    final_LON = coalesce(CBP_LON, PPQ_LON)
  )

# Step 4: For remaining missing coordinates, use IATA code if available
missing_coords <- pathway_risk_geo %>%
  filter(is.na(final_LAT) | is.na(final_LON)) %>%
  filter(!is.na(IATA_CODE))

# Use tidygeocoder to look up IATA airport codes if available
if(nrow(missing_coords) > 0 && requireNamespace("tidygeocoder", quietly = TRUE)) {
  iata_coords <- missing_coords %>%
    dplyr::select(IATA_CODE) %>%
    distinct() %>%
    mutate(query = paste("airport", IATA_CODE)) %>%
    tidygeocoder::geocode(address = query, method = "osm")
  
  # Join the results back
  pathway_risk_geo <- pathway_risk_geo %>%
    left_join(
      iata_coords %>% 
        dplyr::select(IATA_CODE, lat, long) %>% 
        rename(IATA_LAT = lat, IATA_LON = long),
      by = "IATA_CODE"
    ) %>%
    mutate(
      final_LAT = coalesce(final_LAT, IATA_LAT),
      final_LON = coalesce(final_LON, IATA_LON)
    )
} else if(nrow(missing_coords) > 0) {
  cat("Found", nrow(missing_coords), "rows with IATA codes but no coordinates.\n")
  cat("Consider installing the 'tidygeocoder' package to look up these codes.\n")
  print(missing_coords %>% dplyr::select(IATA_CODE) %>% distinct())
}

# Close database connections when done
dbDisconnect(db_conn)
dbDisconnect(geo_conn)
```

# 3. Evaluate Coordinate Coverage

```{r coordinate_coverage}
# Calculate coordinate coverage statistics
coords_summary <- pathway_risk_geo %>%
  summarize(
    total_rows = n(),
    rows_with_coords = sum(!is.na(final_LAT) & !is.na(final_LON)),
    percent_coverage = round(100 * rows_with_coords / total_rows, 1)
  )
print(coords_summary)

# Calculate risk coverage statistics
risk_summary <- pathway_risk_geo %>%
  summarize(
    total_risk = sum(mean_kg, na.rm = TRUE),
    risk_with_coords = sum(mean_kg[!is.na(final_LAT) & !is.na(final_LON)], na.rm = TRUE),
    percent_risk = round(100 * risk_with_coords / total_risk, 1)
  )
print(risk_summary)

# Analyze coordinate sources
source_summary <- pathway_risk_geo %>%
  mutate(
    coord_source = case_when(
      !is.na(CBP_LAT) & !is.na(CBP_LON) ~ "CBP",
      !is.na(PPQ_LAT) & !is.na(PPQ_LON) ~ "PPQ",
      !is.na(IATA_LAT) & !is.na(IATA_LON) ~ "IATA",
      TRUE ~ "Missing"
    )
  ) %>%
  count(coord_source) %>%
  mutate(percent = round(100 * n / sum(n), 1))
print(source_summary)

# Check missing locations
if(coords_summary$rows_with_coords < coords_summary$total_rows) {
  cat("\nSample of locations still missing coordinates:\n")
  print(pathway_risk_geo %>% 
        filter(is.na(final_LAT) | is.na(final_LON)) %>% 
        dplyr::select(LOCATION_NAME, PATHWAY) %>% 
        distinct())
}
```

# 4. Convert to Spatial Objects

```{r convert-to-spatial}
# Filter out rows without coordinates and convert to sf object
pathway_risk_geo_sf <- pathway_risk_geo %>% 
  filter(!is.na(final_LAT) & !is.na(final_LON)) %>% 
  sf::st_as_sf(coords = c("final_LON", "final_LAT"), crs = 4326) %>%
  # Transform to NAD83/Albers (EPSG:5070) for analysis
  sf::st_transform(crs = 5070)

# Save spatial object for future use
saveRDS(pathway_risk_geo_sf, here::here("output", "aqim based volume of at risk product approaching", 
                                       "pathway_risk_with_geolocation.rds"))

                                       
pathway_risk_geo_sf <- readRDS(here::here("output", "aqim based volume of at risk product approaching", 
                   "pathway_risk_with_geolocation.rds"))

View(pathway_risk_geo_sf )

```

# 5. Filter to US Boundaries

```{r us-boundaries}
# Load US boundaries 
us_bounds_sf <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_bounds.rds"))
us_grid <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_grid.rds"))

# Filter spatial data to US boundaries
pathway_risk_geo_sf_filtered <- pathway_risk_geo_sf %>%
  st_join(us_bounds_sf, left = TRUE) 

# Report filtering statistics
total_points <- nrow(pathway_risk_geo_sf)
kept_points <- nrow(pathway_risk_geo_sf_filtered)
removed_points <- total_points - kept_points

cat(sprintf("Filtered to %d points within US boundaries (%d removed)\n", 
          kept_points, removed_points))
```

# 6. Summarize Data for Spatial Layers

```{r summarize-for-spatial-layers}
# Group by location and pathway combinations and summarize
# This doesn't do anything if year and month are taken out of the summary in 01_06 aqim pathway risk_PoE.Rmd
spatial_layers_summary <- pathway_risk_geo_sf_filtered %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, 
           container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(mean_kg, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Filter out rows that dont' have any underlying risk from the AQIM sampling (i.e., no approach rate)
  filter(!is.na(container) & !is.na(product_pathway) & !is.na(product_subpathway))
```
```{r}
spatial_layers_summary |> 
  group_by(PATHWAY, container, product_pathway, product_subpathway) |> 
  summarize(total_kg = sum(total_mean_kg, na.rm = TRUE))

```


# 7. Create and Save Spatial Layers Function

```{r test chunk}
# spatial_layers_summary_test <- spatial_layers_summary %>%
#   filter(grepl('Airport', PATHWAY) & grepl('CBP', PATHWAY)  & grepl("consumption", product_subpathway))

spatial_layers_summary_test2 <- spatial_layers_summary %>%
  filter(grepl('Airport', PATHWAY)  & grepl("consumption", product_subpathway))
```
- there was an error where shortening the string in the function below (so the filename isn't too long when it is saved) caused these 2 to overlap in their name. Probably not an issue in the future but leaving it in here for awareness.


```{r create-spatial-layers-function}
# Function to create and save individual spatial layers for each pathway combination
create_spatial_layers <- function(summarized_data, save_files = FALSE, output_dir = NULL) {
#  Split data by pathway combinations
  split_data <- summarized_data %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    group_split()
  
  cat(sprintf("Created %d spatial layers for different pathway combinations\n", length(split_data)))
  
  # Create list to store the layers
  spatial_layers <- list()
  
  # Create scenario string based on affected_countries parameter
  scenario_string <- dplyr::case_when(
    params$affected_countries == "current" ~ "current",
    params$affected_countries == "mex_can" ~ "mex_can",
    params$affected_countries == "other" ~ "custom",
    TRUE ~ "unspecified"
  )
  
  # Process each split
  for (i in seq_along(split_data)) {
    current_split <- split_data[[i]]
    
    # Get combination values for naming
    combo <- current_split %>%
      dplyr::select(PATHWAY, container, product_pathway, product_subpathway) %>%
      slice(1) %>%
      as.list()
    
    # Create clean name for layer using abbreviations
    layer_name <- paste(
      substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 27),
      substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3),
      sep = "_"
    )
    
    # Store in list with clean name
    spatial_layers[[layer_name]] <- current_split
    
    # Save to file if requested
    if (save_files && !is.null(output_dir)) {
      # Ensure output directory exists
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Convert to POINT geometry (required for GeoPackage)
      export_layer <- current_split %>%
        sf::st_cast("POINT")
      
      # Include scenario in filename - use GeoPackage format for Shiny compatibility
      file_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".gpkg"))
      
      # Try to save the file
      tryCatch({
        sf::st_write(export_layer, file_path, delete_layer = TRUE)
        cat(sprintf("Saved layer %d/%d: %s\n", i, length(split_data), file_path))
      }, error = function(e) {
        cat(sprintf("ERROR saving layer %d/%d: %s - %s\n", 
                   i, length(split_data), layer_name, e$message))
      })
    }
  }
    # Calculate totals directly from the original summarized data to ensure consistency
  # This approach guarantees the same results as the manual grouping and summarization
  layer_summary <- summarized_data %>%
    sf::st_drop_geometry() %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    summarize(
      total_risk = sum(total_mean_kg, na.rm = TRUE),
      point_count = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      layer_name = paste(
        substr(gsub("[^[:alnum:]]", "", PATHWAY), 1, 27),
        substr(gsub("[^[:alnum:]]", "", container), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_pathway), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_subpathway), 1, 3),
        sep = "_"
      ),
      scenario = scenario_string
    ) %>%
    dplyr::select(layer_name, scenario, point_count, total_risk) %>%
    arrange(desc(total_risk))
  
  # Print summary
  cat(paste0("\nSummary of spatial layers for scenario '", scenario_string, "' (top 10 by risk):\n"))
  print(head(layer_summary, 10))
  
  # Also save the summary table as CSV for reference
  if (save_files && !is.null(output_dir)) {
    summary_path <- file.path(output_dir, paste0("layer_summary_", scenario_string, ".csv"))
    write.csv(layer_summary, summary_path, row.names = FALSE)
    cat(sprintf("\nSaved summary table to: %s\n", summary_path))
  }
  
  return(list(
    layers = spatial_layers,
    summary = layer_summary,
    scenario = scenario_string
  ))
}
```

# 8. Create Port Risk Spatial Layers

```{r create-port-risk-layers}
# Set output directory for port risk layers
port_risk_output_dir <- here::here("output", "spatial layers", "risk at ports")

# Create the spatial layers
spatial_layers <- create_spatial_layers(
  summarized_data = spatial_layers_summary,
  save_files = TRUE,
  output_dir = port_risk_output_dir
)
```

# 9. Prepare Data for Final Destination Risk Layers

```{r prepare-final-destination-data}
# Summarize by pathway combination (dropping geometry)
pathway_spatial_layers_summary <- spatial_layers_summary %>% 
  sf::st_drop_geometry() %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(total_mean_kg, na.rm = TRUE),
    .groups = 'drop'
  )

setwd(here::here('output','report files'))
write.csv(pathway_spatial_layers_summary, paste("aqim pathway summary", min(params$years), "to", max(params$years), params$affected_countries, params$n_sims_for_volume, "sims",".csv", sep = "_"), row.names = FALSE)
```

# 10. Create Final Destination Risk Layers Function

```{r distribute-pathway-risks-function}
# Function to distribute pathway risks to final destination spatial grids
distribute_pathway_risks <- function(pathway_summary, kde_folder_path) {
  # List all available KDE files (kernel density estimation grids)
  kde_files <- list.files(path = kde_folder_path, pattern = "kde_scaled_.*\\.rds", full.names = TRUE)
  
  # Create output directory for final destination risk layers
  output_dir <- here::here('output', 'spatial layers', 'risk at final destinations')
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Initialize data structures
  distributed_layers <- list()
  mapping_df <- data.frame(
    short_name = character(),
    full_name = character(),
    pathway = character(),
    container = character(),
    product_pathway = character(),
    product_subpathway = character(),
    stringsAsFactors = FALSE
  )
  
  # Process each unique pathway
  for (pathway in unique(pathway_summary$PATHWAY)) {
    # Find the matching KDE file for this pathway
    kde_filename_pattern <- paste0("kde_scaled_", pathway, ".rds")
    matching_files <- grep(kde_filename_pattern, kde_files, value = TRUE, fixed = TRUE)
    
    if (length(matching_files) == 0) {
      cat(sprintf("No KDE file found for pathway: %s\n", pathway))
      next
    }
    
    # Load the KDE grid for this pathway
    kde_grid <- readRDS(matching_files[1])
    
    # Get all combinations for this pathway
    pathway_combos <- pathway_summary %>% filter(PATHWAY == pathway)
    
    # Process each combination
    for (i in seq_len(nrow(pathway_combos))) {
      combo <- pathway_combos[i,]
      risk_value <- combo$total_mean_kg
      
      # Create abbreviated name components for file naming
      pathway_abbr <- substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 27)
      container_abbr <- substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3)
      prod_path_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3)
      prod_subpath_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3)
      
      # Create short name for file (with unique index)
      short_combo_name <- paste(pathway_abbr, container_abbr, prod_path_abbr, prod_subpath_abbr, i, sep = "_")
      
      # Create full descriptive name for reference
      full_combo_name <- paste(
        gsub("[^[:alnum:]]", "_", combo$PATHWAY),
        gsub("[^[:alnum:]]", "_", combo$container),
        gsub("[^[:alnum:]]", "_", combo$product_pathway),
        gsub("[^[:alnum:]]", "_", combo$product_subpathway),
        sep = "_"
      )
      
      # Apply the risk value to the KDE grid to create distributed risk
      distributed_kde <- kde_grid %>%
        dplyr::mutate(risk_value = kde_value * risk_value)
      
      # Store in layers list
      distributed_layers[[full_combo_name]] <- distributed_kde
      
      # Add mapping entry
      new_mapping <- data.frame(
        short_name = short_combo_name,
        full_name = full_combo_name,
        pathway = combo$PATHWAY,
        container = combo$container,
        product_pathway = combo$product_pathway,
        product_subpathway = combo$product_subpathway,
        stringsAsFactors = FALSE
      )
      mapping_df <- rbind(mapping_df, new_mapping)
      
      # Save as GeoPackage for Shiny compatibility
      output_file <- file.path(output_dir, paste0("risk_", short_combo_name, ".gpkg"))
      
      tryCatch({
        sf::st_write(distributed_kde, output_file, delete_layer = TRUE)
        cat(sprintf("Processed %s → %s with risk %f\n", full_combo_name, short_combo_name, risk_value))
      }, error = function(e) {
        cat(sprintf("ERROR: Could not write %s - %s\n", short_combo_name, e$message))
      })
    }
  }
  
  # Save the name mapping file for reference
  write.csv(mapping_df, file.path(output_dir, "filename_mapping.csv"), row.names = FALSE)
  
  # Create a layer catalog with additional metadata
  layer_catalog <- mapping_df %>%
    mutate(
      file_path = file.path(output_dir, paste0("risk_", short_name, ".gpkg")),
      risk_value = sapply(distributed_layers[full_name], function(x) sum(x$risk_value, na.rm=TRUE)),
      cell_count = sapply(distributed_layers[full_name], nrow)
    ) %>%
    arrange(desc(risk_value))
  
  # Save the catalog
  write.csv(layer_catalog, file.path(output_dir, "layer_catalog.csv"), row.names = FALSE)
  
  return(list(
    layers = distributed_layers,
    mapping = mapping_df,
    catalog = layer_catalog
  ))
}
```

# 11. Create Final Destination Risk Layers

```{r create-final-destination-layers}
# Apply the function to distribute risks to final destinations
distributed_risks <- distribute_pathway_risks(
  pathway_summary = pathway_spatial_layers_summary,
  kde_folder_path = here::here("output", "proportional destination layers")
)

# Save a compact version of the catalog for Shiny app use
risk_summary <- distributed_risks$catalog %>%
  dplyr::select(
    layer_name = full_name, 
    short_name,
    pathway,
    container, 
    product_pathway,
    product_subpathway,
    total_risk = risk_value, 
    cell_count
  )

# Save the summary for reference
write.csv(risk_summary, 
          here::here("output", "aqim total distributed risk", paste0("distributed_risk_summary_", params$affected_countries,".csv")), 
          row.names = FALSE)

cat("Spatial risk layers created and ready for use in Shiny app\n")
```

```{r test open} 
library(sf)

# Open a GeoPackage file
gpkg_data <- st_read(here::here("output", "spatial layers", "risk at final destinations", "risk_Air_bag_swi_per_2.gpkg"))

# View the data
head(gpkg_data)

# If you want to see the structure
str(gpkg_data)

# Plot the spatial data
plot(gpkg_data$geom)  # Basic plot
plot(st_geometry(gpkg_data))  # Alternative syntax


valid_geoms <- st_is_valid(gpkg_data)
if(all(valid_geoms)) {
  plot(gpkg_data)
} else {
  # Try to fix invalid geometries
  fixed_gpkg <- st_make_valid(gpkg_data)
  plot(fixed_gpkg)
}

```