---
title: "aqim pathway passenger volumes"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  wads_volume_begin_date: "2023-01-01" # Start date for pulling WADS volume data
  wads_volume_end_date: "2024-12-31" # End date for pulling WADS volume data
  wads_volume_reference_filename: "wads volume reference table.csv" # Name of the input reference file that houses which ACTIVITY_CODE is for which PATHWAY
  wads_volume_filename: wads_passenger_volume_by_port_pathway_month_year # Name of the output file for WADS volume data
  wads_ECO_volume: (mean(1066000000+1400000000)/12) # Total express courier volume to supplement WADS data if needed (2023 + 2024)
---

Third Rmd

-   Separate the WADS reference table for a column that identifies if the line item is passenger volume or intercepted material; see input\>data\>wads codes\>get wads codes.R; process those codes (started this process with WADS Lookup Codes.csv but I think I need to include more WADS codes)

    -   WADS for passenger volumes at each port

    -   WADS interceptions at each port

-   Produce a table with each identified combination that calculates:

    -   Port Code

    -   ASF item volume (volume \* rates for each combination)

        -   This is a single calculation - passenger volume at port X the total pathway-combination's rate,

    -   Interceptions of the combination

    -   Entry (volume - interceptions) of the combination

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(here)
library(lubridate)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      wads_volume_begin_date = "2021-01-01",
      wads_volume_end_date = "2024-12-31",
      wads_volume_reference_filename = "wads volume reference table.csv",
      wads_volume_filename = "wads_passenger_volume_by_port_pathway_month_year",
      wads_ECO_volume = (mean(1066000000+1400000000)/12) # Total express courier volume to supplement WADS data if needed (2023 + 2024)
    )
  }
}
```

## gather WADS codes
### Get WADS reference table for volume, inspections, and interceptions
```{r get wads codes}
setwd(here::here("input","data","wads codes"))
wads_ref_table <- read.csv(params$wads_volume_reference_filename) 

wads_vol_ref <- wads_ref_table |> 
  filter(quantification_purpose %in% c("arrival")) |>
  dplyr::select(CODE, PATHWAY)

wads_volume_codes <- wads_vol_ref |> pull(CODE)
# wads_ref_list <- split(wads_ref_table, wads_ref_table$quantification_purpose)
# View(wads_ref_list$volume)

```

### Connect to ARM
```{r set driven}

db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

wads_cols <- c(
  "LOCATION_ID", 
  "LOCATION_NAME", 
  "LOCATION_STATE_CODE", 
  "COUNT", 
  "MAX", 
  "MIN", 
  "ACTIVITY_CODE",
  "WORK_ACTIVITY_DATE"
  )


#-- Pull AQIM inspection records from ARM  ---- 
system.time(wads <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_WADS]")) |> 
              dplyr::select(all_of(wads_cols)) |> 
              #filter to begin and end date
              dplyr::filter(WORK_ACTIVITY_DATE >= params$wads_volume_begin_date & WORK_ACTIVITY_DATE <= params$wads_volume_end_date) |>
              dplyr::filter(ACTIVITY_CODE %in% wads_volume_codes) |>
              collect()
           )
 
```

## Summarize WADS volume by month, year, location, and pathway
```{r summarize wads volume}
# First, adjust count so that each count falls within the min/max range for that activity code
# If COUNT is below MIN, then COUNT = MIN; if COUNT is above MAX, then COUNT = MAX.
#If MIN is NULL, then do not adjust if COUNT is below MIN.
#If MAX is NULL, then do not adjust if COUNT is above MAX.
#If both MIN and MAX are NULL, then do not adjust COUNT.
wads <- wads |> 
  mutate(COUNT = case_when(
    !is.na(MIN) & COUNT < MIN ~ MIN,
    !is.na(MAX) & COUNT > MAX ~ MAX,
    .default = COUNT
  )) |> dplyr::select(-c(MIN, MAX))

wads_volume <- wads |> 
  left_join(wads_vol_ref, by = c("ACTIVITY_CODE" = "CODE")) |> 
  mutate(year = lubridate::year(WORK_ACTIVITY_DATE),
         month = lubridate::month(WORK_ACTIVITY_DATE)) |> 
         #reduce to only relevant pathways
  filter(!(is.na(PATHWAY) | PATHWAY == "")) |>
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, year, month) |> 
  summarize(COUNT = sum(COUNT, na.rm=TRUE),
            .groups = "drop")

```

```{r}
#Verify its working correctly
test = wads_volume |> filter(grepl('Airport - Aircraft - Person', PATHWAY)) |> group_by(LOCATION_NAME, year) |> summarize(total_count = sum(COUNT, na.rm = TRUE))
View(test)
sum(test$total_count)
print(paste("the total air travelers is", format(sum(test$total_count), big.mark=","), "over", params$wads_volume_begin_date, "to", params$wads_volume_end_date))

test2 <- wads_volume |> filter(grepl('Airport - Aircraft - Person', PATHWAY)) |>
group_by(LOCATION_NAME, PATHWAY) |> summarize(total_count = sum(COUNT, na.rm = TRUE))
View(test2)
```


# Supplement WADS ECO volume
There isa WADS volume for Express_courier, but it sums to a few thousand whereas we know that ~ 1.066bn packages entered based on CBP reporting.
So we'll use that number to adjust the COUNT in the WADS volum; I'll write it flexibly so that it's a controlled parameter

```{r supplement WADS express courier data}
# Check if ECO pathway exists in WADS data
if(!"Express_courier" %in% wads_volume$PATHWAY) {
  # If not, create a new entry for ECO with the total volume
  # Check if parameter is defined, if not use default
  eco_volume <- if(!is.null(params$wads_ECO_volume)) params$wads_ECO_volume else (mean(1066000000+1400000000)/12) # 1.066bn from 2023 + 1.4bn from 2024
  message("Using Express_courier volume: ", eco_volume)
  
  eco_entry <- data.frame(
    LOCATION_ID = 0,
    LOCATION_NAME = "All Locations",
    LOCATION_STATE_CODE = "ALL",
    PATHWAY = "Express_courier",
    year = NA,
    month = NA,
    COUNT = eco_volume
  )
  # Append ECO entries to WADS volume data
  wads_volume <- bind_rows(wads_volume, eco_entry)
} else {
  # Check if parameter is defined, if not use default
  eco_volume <- if(!is.null(params$wads_ECO_volume)) params$wads_ECO_volume else (mean(1066000000+1400000000)/12) # 1.066bn from 2023 + 1.4bn from 2024
  message("Using Express_courier volume: ", eco_volume)
  
  #scale the existing ECO volume to match the known total. Use the COUNT to get the proportion for each location, and then scale to the known total
  existing_eco_volume <- sum(wads_volume$COUNT[wads_volume$PATHWAY == "Express_courier"], na.rm = TRUE)
  if(existing_eco_volume > 0) {
    scaling_factor <- eco_volume / existing_eco_volume
    wads_volume <- wads_volume %>%
      mutate(COUNT = ifelse(PATHWAY == "Express_courier", round(COUNT * scaling_factor, 0), COUNT))
    message("Scaled Express_courier volume by factor: ", scaling_factor)
  } else {
    warning("Existing Express_courier volume is 0 or NA, cannot scale properly")
  }
}

```

```{r define border state codes}
NB_state_cd <- c("AK","ID","ME","MI","MN","MT","NH","NY","ND","OH","PA","VT","WA")
SB_state_cd <- c("CA","AZ","NM","TX")
```
```{r separate land border to NB-SB}
wads_volume <- wads_volume %>%
  mutate(
    PATHWAY = case_when(
      LOCATION_STATE_CODE %in% NB_state_cd & grepl("Land Border|Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " NB"),
      LOCATION_STATE_CODE %in% SB_state_cd & grepl("Land Border|Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " SB"),
      .default = PATHWAY
    )
  )
```

# drop month and year

- 9/25 - something is not working correctly. I am going to try to drop month and year and see if that helps
```{r drop month year}
wads_volume <- wads_volume %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY) %>%
  summarize(COUNT = sum(COUNT, na.rm = TRUE), .groups = "drop")
```

## Save summarized WADS volume
```{r save wads volume}
setwd(here::here("output","wads","arrivals_or_entries"))
write.csv(wads_volume, paste(params$wads_volume_filename, ".csv") )
saveRDS(wads_volume, paste(params$wads_volume_filename, ".rds"))
```

