---
title: "aqim pathway approach rates"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  begin_date: 2023-01-01
  end_date: 2024-12-31
  affected_country_file: "asf affected countries.csv"
  affected_product_file: "asf affected product name reference table.csv"
  pathway_volume_smuggling_reference_table_filename: "product pathway by volume reference table"
  quantity_units_to_kg: "quantity units to kilograms reference table.csv"
---

-   \% estimate of each pathway type for each AQIM-conveyance/pathway

<!-- -->

-   Only use 2022-2024 data

    -   Use the reference table (at risk product names.csv) to assign every commodity intercepted to a pathway & subpathway

    -   Further modify the reference table that identifies every product with additional columns for PoE, Conveyance, Subconveyance, Edge as possible

    -   For each PoE, Conveyance, Subconveyance, Edge, Subpathway, Pathway, estimate the rate of entry

# setup

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(countrycode)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      begin_date = "2021-01-01",
      end_date = "2024-12-31",
      affected_product_file = "asf affected product name reference table.csv",
      affected_country_file = "asf affected countries.csv",
      quantity_units_to_kg = "quantity units to kilograms reference table.csv",
      pathway_volume_smuggling_reference_table_filename = "product pathway by volume reference table"
    )
  }
}
```

```{r define border state codes}
NB_state_cd <- c("AK","ID","ME","MI","MN","MT","NH","NY","ND","OH","PA","VT","WA")
SB_state_cd <- c("CA","AZ","NM","TX")
```

## Filtering by dates within the dplyr package

Can't just filter by the datetime because it needs lubridate() to manipulate the datetime. Instead, can filter by fiscal month and year, which are reported, but the needed datetimes need to be converted first

```{r convert calendar dates to fiscal}
# Load necessary libraries
library(lubridate)
library(dplyr)

fiscal_date_to_calendar_month <- function(fiscal_date) {
  tryCatch({
    if(is.na(fiscal_date)) return(NA)
    
    # Ensure we're working with a Date object
    if(!inherits(fiscal_date, c("Date", "POSIXct", "POSIXlt"))) {
      fiscal_date <- as.Date(fiscal_date)
    }
    
    as.character((as.integer(format(fiscal_date, "%m")) - 10) %% 12 + 1)
  }, error = function(e) {
    NA  # Return NA if conversion fails
  })
}

fiscal_date_to_calendar_year <- function(fiscal_date) {
  tryCatch({
    if(is.na(fiscal_date)) return(NA)
    
    # Ensure we're working with a Date object
    if(!inherits(fiscal_date, c("Date", "POSIXct", "POSIXlt"))) {
      fiscal_date <- as.Date(fiscal_date)
    }
    
    as.character(
      ifelse(as.integer(format(fiscal_date, "%m")) >= 10,
             as.integer(format(fiscal_date, "%Y")) + 1,
             as.integer(format(fiscal_date, "%Y")))
    )
  }, error = function(e) {
    NA  # Return NA if conversion fails
  })
}

# Assume your data frame is called df and the date column is 'date'
begin_date <- as.Date(params$begin_date)  # Ensure it's a Date type

# Fiscal year: If month >= 10, fiscal year is year + 1, else year
begin_year <- fiscal_date_to_calendar_year(begin_date)

# Fiscal month: October is 1, November is 2, ..., September is 12
begin_month <- fiscal_date_to_calendar_month(begin_date)

# Assume your data frame is called df and the date column is 'date'
end_date <- as.Date(params$end_date)  # Ensure it's a Date type

# Fiscal year: If month >= 10, fiscal year is year + 1, else year
end_year <- fiscal_date_to_calendar_year(end_date)

# Fiscal month: October is 1, November is 2, ..., September is 12
end_month <- fiscal_date_to_calendar_month(end_date)

```

```{r create environmental params}

# Load ASF affected country name reference table
setwd(here::here('input', 'data', 'asf affected countries by year'))
asf_country_ref <- read.csv(params$affected_country_file, stringsAsFactors = FALSE)


# Load ASF affected product name reference table
setwd(here::here('input', 'data', 'asf affected product names'))
asf_product_ref <- read.csv(params$affected_product_file, stringsAsFactors = FALSE)
quantity_units_to_kg <- read.csv(params$quantity_units_to_kg, stringsAsFactors = FALSE)

```

## ARM AQIM data

get arm data

```{r ARM AQIM data}

db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

inspection_cols <- c(
  "ID", 
  "INSPECTION_NUMBER", 
  "CATEGORY", 
  "SUBCATEGORY", 
  "IS_AQIM", 
  "PATHWAY_ID", 
  "PATHWAY",
  "PORT_OF_ENTRY_ID",
  
  "INSPECTION_DATETIME", 
  "INSPECTION_DATETIME_FISCAL_YEAR", 
  "INSPECTION_DATETIME_FISCAL_MONTH",
  "INSPECTION_LOCATION_ID", 
  "INSPECTION_LOCATION_NAME", 
  "INSPECTION_LOCATION_STATE_CODE", 
  "COUNTRY_OF_ORIGIN_NAME", 
 # "COUNTRY_OF_RESIDENCE_NAME", 
  
 # "FINAL_DESTINATION_STATE_NAME", 
 # "FINAL_DESTINATION_STATE_ID", 
 # "FINAL_DESTINATION_CITY", 
 # "PASSENGER_DISTANCE_DESTINATION_NAME",
  "NUMBER_OF_PASSENGERS")


commodity_cols <- c(
  "ID",
  "INSPECTION_ID",
  "IS_SMUGGLED",
  "IS_CONTAMINANT_FOUND",
  "COMMODITY_CLASSIFICATION",
  "COUNTRY_OF_ORIGIN_NAME", 
  "REF_COMMODITY_ID",
  "COMMODITY_DISPLAY_NAME",
  "MISC_ANIMAL_COMMODITY_NAME",
  "QUANTITY",
  "QUANTITY_UNITS_NAME"
)


#-- Pull AQIM inspection records from ARM  ---- 
system.time(arm_aqim <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_INSPECTION]")) |> 
              dplyr::filter(CATEGORY == "AQIM") |>
              dplyr::select(inspection_cols) |> 
              #rename to fit AQAS data structure
              rename(edge_origin = COUNTRY_OF_ORIGIN_NAME,
                      group_size = NUMBER_OF_PASSENGERS
           )
)
 
#-- Load Bridge Commodity dataset ----
system.time(commodity <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |> 
              dplyr::select(commodity_cols) |> 
              dplyr::rename(COMMODITY_ID = ID,
              #rename to match AQAS data structure
                            product_origin = COUNTRY_OF_ORIGIN_NAME)
            )

arm_aqim_records <- arm_aqim |> left_join(commodity, by = c("ID" = "INSPECTION_ID"), relationship = "many-to-many")

# #filter to pathway of interest
# system.time(arm_aqim_records <- arm_aqim_records |> dplyr::filter(PATHWAY %in% params$pathway)
# )

#filter to on or after begin date
system.time(arm_aqim_records <- arm_aqim_records  |>
  filter(INSPECTION_DATETIME_FISCAL_YEAR > begin_year | (INSPECTION_DATETIME_FISCAL_YEAR == begin_year & INSPECTION_DATETIME_FISCAL_MONTH >= begin_month)
  )
)

#filter to at or before end date
system.time(arm_aqim_records <- arm_aqim_records |> 
  filter(INSPECTION_DATETIME_FISCAL_YEAR < end_year | (INSPECTION_DATETIME_FISCAL_YEAR == end_year & INSPECTION_DATETIME_FISCAL_MONTH <= end_month)
  )
)


#collect records
system.time(arm_aqim_records <- arm_aqim_records |> collect()
            )
```

```{r get month and year from datetime}

system.time(arm_aqim_records <- arm_aqim_records %>%
  mutate(
    CALENDAR_MONTH = lubridate::month(INSPECTION_DATETIME),
    CALENDAR_YEAR = lubridate::year(INSPECTION_DATETIME )
  )
)
```

```{r separate ARM land border data to NB-SB}
arm_aqim_records <- arm_aqim_records %>%
  mutate(
    PATHWAY = case_when(
      INSPECTION_LOCATION_STATE_CODE %in% NB_state_cd & grepl("Land Border", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " NB"),
      INSPECTION_LOCATION_STATE_CODE %in% SB_state_cd & grepl("Land Border", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " SB"),
      TRUE ~ PATHWAY
    )
  )
```

```{r convert arm countries to iso codes}
arm_aqim_records <- arm_aqim_records %>% 
  mutate(
    edge_origin = countrycode(edge_origin, 
                                  origin = "country.name", 
                                  destination = "iso3c"),
    product_origin = countrycode(product_origin, 
                                    origin = "country.name", 
                                    destination = "iso3c")
  )

```


# AQAS AQIM data

Will have to get old AQAS data as well, and make sure it matches naming structures

```{r connect to AQAS framework}
db_conn = DBI::dbConnect(
  odbc::odbc(),
  .connection_string =
    "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_AQAS_DW;
                        trusted_connection=yes"
  )


```

# Function to collect AQAS data for each pathway

This function will collect the AQAS data for each pathway, and then join it with the commodity data. It will also rename the columns to match the ARM AQIM data structure. It will also filter the data to only include the relevant columns and add a PATHWAY column to identify the pathway type. It will also convert the quantity units to kilograms if they are labeled "Each" and calculate the total kilograms for each commodity on the assumption that "Each" weights 0.5kg (the instructions given for labeling kg if a fruit).

```{r AQAS data collection function}

aqas_aqim_func <- function(insp_data, commod_data, pathway_name, 
                           month, year, group_size, action_count, qmi_count, clean_treat_count,
                           passenger_origin, item_origin) {
  # First check if the group_size column exists in the data
  avail_cols <- colnames(insp_data)
  
  # Create a select_cols vector with columns that definitely exist
  select_cols <- c("AQIM_FORM_ID", "LOCATION", "PORT_CD", "STATE_CD", month, year, action_count)
  
  # Conditionally add columns that may or may not exist
  if(group_size %in% avail_cols) {
    select_cols <- c(select_cols, group_size)
  }
  if("LOCATION_ID" %in% avail_cols) {
    select_cols <- c(select_cols, "LOCATION_ID")
  }
  if(qmi_count %in% avail_cols) {
    select_cols <- c(select_cols, qmi_count)
  }
  if(clean_treat_count %in% avail_cols) {
    select_cols <- c(select_cols, clean_treat_count)
  }
  if(passenger_origin %in% avail_cols) {
    select_cols <- c(select_cols, passenger_origin)
  }
  if(item_origin %in% avail_cols) {
    select_cols <- c(select_cols, item_origin)
  }

  
  # Now select only the columns that actually exist
  insp_df <- insp_data %>% 
    dplyr::select(all_of(select_cols)) %>%
    mutate(PATHWAY = pathway_name) %>%
    relocate(PATHWAY, .after = AQIM_FORM_ID)
  
  # Create a list of rename mappings, including only those that exist
  rename_list <- list(
    "CALENDAR_MONTH" = month,
    "CALENDAR_YEAR" = year,
    "INSPECTION_LOCATION_NAME" = "LOCATION",
    "INSPECTION_LOCATION_STATE_CODE" = "STATE_CD"
  )
  
  # Add conditional renames
  if(group_size %in% avail_cols) {
    rename_list[["group_size"]] <- group_size
  }
  if("LOCATION_ID" %in% avail_cols) {
    rename_list[["INSPECTION_LOCATION_ID"]] <- "LOCATION_ID"
  }
  if(action_count %in% avail_cols) {
    rename_list[["actions"]] <- action_count
  }
  if(qmi_count %in% avail_cols) {
    rename_list[["qmis"]] <- qmi_count
  }
  if(clean_treat_count %in% avail_cols) {
    rename_list[["ct_count"]] <- clean_treat_count
  }
  if(passenger_origin %in% avail_cols) {
    rename_list[["edge_origin"]] <- passenger_origin
  }
  if(item_origin %in% avail_cols) {
    rename_list[["product_origin"]] <- item_origin
  }
  
  # Apply the renames that are applicable
  insp_df <- insp_df %>% rename(!!!rename_list)
  
  # Process commodity data as before
  commod_df <- commod_data %>% 
    dplyr::select(AQIM_FORM_ID, ITEM, QUANTITY, UNIT, ACTN_TAKEN) %>%
    filter(!ACTN_TAKEN == "Inspected and Released")
  
  # Join the data as before
  df <- insp_df %>%
    left_join(commod_df, by = "AQIM_FORM_ID", relationship = "many-to-many") %>%
    rename("ID" = "AQIM_FORM_ID")
  
  return(df)
}


```


```{r AQAS pedestrian - separated by NB-SB aqim}

insp_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_PEDESTRIAN_MV]'))

commod_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_PEDESTRIAN_ITEM_MV]'))

ped <- aqas_aqim_func(insp_data,
                      commod_data,
                      pathway_name = "Pedestrian",
                      month = "MON",
                      year = "CALENDAR_YR",
                      group_size = "NUM_GRP",
                      action_count = "ACTION_CNT",
                      qmi_count = "QMI_CNT",
                      clean_treat_count = "CT_CNT",
                      passenger_origin = "PAX_ORIGIN_CD",
                      item_origin = "ITEM_ORIGIN_CD")

ped <- collect(ped)

##NB pedestrian does not exist
#View(ped %>% count(INSPECTION_LOCATION_STATE_CODE))
```

```{r AQAS express carrier}

insp_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_EXP_CARRIER_MV]'))
                     
commod_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_EXP_CARRIER_ITEM_MV]'))

eco <- aqas_aqim_func(insp_data,
                      commod_data,
                      pathway_name = "Express_courier",
                      month = "MON",
                      year = "CALENDAR_YR",
                      group_size = "NUM_GRP",
                      action_count = "ACTION_CNT",
                      qmi_count = "QMI_CNT",
                      clean_treat_count = "CT_CNT",
                      passenger_origin = "COUNTRY_CD",
                      item_origin = "PARCEL_ORIGIN_CD")

eco <- collect(eco)

```

```{r AQAS ship}
insp_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_SHIP_MV]'))
                     
commod_data <- tbl(db_conn,sql('SELECT
                      *
                     FROM [PPQ_AQI_AQAS_DW].[DW_AQAS].[AQIM_SHIP_ITEM_MV]'))

ship <- aqas_aqim_func(insp_data,
                      commod_data,
                      pathway_name = "Cruise_ship",
                      month = "MON",
                      year = "CALENDAR_YR",
                      group_size = "NUM_PAX",
                      action_count = "ACTION_CNT",
                      qmi_count = "QMI_CNT",
                      clean_treat_count = "CT_CNT",
                      passenger_origin = "COUNTRY_CD",
                      item_origin = "PARCEL_ORIGIN_CD")

ship <- collect(ship)

```

## Bind aqas aqim data together

```{r combined aqas}
aqas_aqim <- dplyr::bind_rows(ped, eco, ship)

# change ITEM name to fit ARM data structure
aqas_aqim <- aqas_aqim %>%
  rename(COMMODITY_DISPLAY_NAME = ITEM)

```

## Correct location names to be compatible with ARM WADS location names
```{r correct location names}
setwd(here::here('input', 'data', 'aqas to arm port location name crosswalk'))
aqas_location_ref <- read.csv("aqas to arm port location name crosswalk.csv", stringsAsFactors = FALSE)
aqas_aqim <- aqas_aqim %>%
  left_join(aqas_location_ref, by = c("INSPECTION_LOCATION_NAME" = "AQAS_location_name"), relationship = "many-to-one") 

aqas_aqim <- aqas_aqim %>%
  mutate(INSPECTION_LOCATION_NAME = coalesce(ARM_location_name, INSPECTION_LOCATION_NAME)) %>%
  dplyr::select(-ARM_location_name, port_type)

#View(aqas_aqim)
```
```{r filter low observation years}
# Count observations by year and pathway, then filter out low-count combinations
low_data_combinations <- aqas_aqim %>%
  count(PATHWAY, CALENDAR_YEAR) %>%
  filter(n < 500)

# Apply the filter to remove low-observation year-pathway combinations
aqas_aqim_filtered <- aqas_aqim %>%
  anti_join(low_data_combinations, by = c("PATHWAY", "CALENDAR_YEAR"))

# Quick summary of records removed
records_before <- nrow(aqas_aqim)
records_after <- nrow(aqas_aqim_filtered)
percent_kept <- round(records_after / records_before * 100, 1)

cat(sprintf("Filtered out %d observations (%.1f%% data retained)\n", 
            records_before - records_after,
            percent_kept))

aqas_aqim <- aqas_aqim_filtered
```

## Separate land border to NB-SB

This will separate the land border pathway into NB and SB pathways based on the STATE_CD of the inspection location. It will append "NB" or "SB" to the PATHWAY name.

```{r separate AQAS land border data to NB-SB}
aqas_aqim <- aqas_aqim %>%
  mutate(
    PATHWAY = case_when(
      INSPECTION_LOCATION_STATE_CODE %in% NB_state_cd & grepl("Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " NB"),
      INSPECTION_LOCATION_STATE_CODE %in% SB_state_cd & grepl("Pedestrian", PATHWAY, ignore.case = TRUE) ~ paste0(PATHWAY, " SB"),
      TRUE ~ PATHWAY
    )
  )
  
```

## standardize unit column name
```{r}
aqas_aqim <- aqas_aqim %>%
  rename(QUANTITY_UNITS_NAME = UNIT)
```

# Combine AQAS and AQIM data
```{r combine aqas and arm aqim data}
# Find matching columns between the two dataframes
matching_columns <- intersect(names(aqas_aqim), names(arm_aqim_records))

# See which columns are unique to each dataframe
aqas_only_columns <- setdiff(names(aqas_aqim), names(arm_aqim_records))
arm_only_columns <- setdiff(names(arm_aqim_records), names(aqas_aqim))

# Print the results for inspection
print("Common columns between both datasets:")
print(matching_columns)

print("Columns only in AQAS data:")
print(aqas_only_columns)

print("Columns only in ARM data:")
print(arm_only_columns)

column_types <- data.frame(
  column_name = matching_columns,
  aqas_type = sapply(matching_columns, function(col) class(aqas_aqim[[col]])[1]),
  arm_type = sapply(matching_columns, function(col) class(arm_aqim_records[[col]])[1]),
  stringsAsFactors = FALSE
)

# Identify columns with mismatched types
mismatched_columns <- column_types[column_types$aqas_type != column_types$arm_type, ]
print("Columns with mismatched types:")
print(mismatched_columns)

# Fix mismatched columns
for (i in 1:nrow(mismatched_columns)) {
  col <- mismatched_columns$column_name[i]
  aqas_type <- mismatched_columns$aqas_type[i]
  arm_type <- mismatched_columns$arm_type[i]
  
  # Choose the most appropriate type based on data needs
  if (aqas_type == "character" && arm_type == "integer64") {
    arm_aqim_records[[col]] <- as.character(arm_aqim_records[[col]])
  } else if (aqas_type == "factor" && arm_type == "character") {
    aqas_aqim[[col]] <- as.character(aqas_aqim[[col]])
  } else if (aqas_type == "numeric" && arm_type == "integer64") {
    arm_aqim_records[[col]] <- as.numeric(arm_aqim_records[[col]])
  } else if (aqas_type == "integer" && arm_type == "numeric") {
    aqas_aqim[[col]] <- as.integer(aqas_aqim[[col]])
  } else if (aqas_type == "character" && arm_type == "double") {
    aqas_aqim[[col]] <- as.character(aqas_aqim[[col]])
  } else if (inherits(aqas_aqim[[col]], "Date") && !inherits(arm_aqim_records[[col]], "Date")) {
    arm_aqim_records[[col]] <- as.Date(arm_aqim_records[[col]])
  } else if (!inherits(aqas_aqim[[col]], "Date") && inherits(arm_aqim_records[[col]], "Date")) {
    aqas_aqim[[col]] <- as.Date(aqas_aqim[[col]])
  }
  # Add more conversions as needed
}

aqim_data <- dplyr::bind_rows(aqas_aqim, arm_aqim_records)
```


# Convert non-kg quantities to kg units

```{r correct quantity}
# Function to convert various unit formats to standardized kg values
convert_to_kg <- function(data, reference_table) {
  # Ensure reference table unit names are lowercase for case-insensitive matching
  reference_table <- reference_table %>%
    mutate(unit_lower = tolower(unit))
  
  # First, standardize the data's unit names by removing plurals and converting to lowercase
  data <- data %>%
    mutate(
      # Make a lowercase version of units for matching
      QUANTITY_UNITS_LOWER = tolower(QUANTITY_UNITS_NAME),
      # Remove trailing 's' from units (e.g., "Kilograms" -> "Kilogram")
      QUANTITY_UNITS_LOWER = gsub("s$", "", QUANTITY_UNITS_LOWER),
      # Ensure quantity is positive
      QUANTITY = abs(QUANTITY)
    )
  
  # Join with reference table using the lowercase, singular version
  data <- data %>%
    left_join(
      reference_table %>% 
        mutate(unit_lower = gsub("s$", "", unit_lower)),
      by = c("QUANTITY_UNITS_LOWER" = "unit_lower"),
      relationship = "many-to-many"
    )

  # Multiply the quantity observed and the est_kilo per unit to get total kg
  data <- data %>%
    mutate(est_kilograms = as.numeric(est_kilograms))
  
  # Apply conversion and handle missing values with common-sense defaults
  data <- data %>%
    mutate(
      TOTAL_KG = case_when(
        !is.na(est_kilograms) ~ QUANTITY * est_kilograms,
        # Fallback rules for missing conversions
        grepl("gram", QUANTITY_UNITS_LOWER) & !grepl("kilo", QUANTITY_UNITS_LOWER) ~ QUANTITY / 1000,
        grepl("kilo", QUANTITY_UNITS_LOWER) ~ QUANTITY,
        grepl("pound", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.453592,
        grepl("ounce", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.0283495,
        grepl("each", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.5, # Assuming 0.5kg per item
        grepl("piece", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.5,
        grepl("ton", QUANTITY_UNITS_LOWER) & !grepl("metric", QUANTITY_UNITS_LOWER) ~ QUANTITY * 907.185,
        grepl("ton", QUANTITY_UNITS_LOWER) & grepl("metric", QUANTITY_UNITS_LOWER) ~ QUANTITY * 1000,
        TRUE ~ NA_real_
      )
    ) %>%
    # Clean up temporary columns
    dplyr::select(-QUANTITY_UNITS_LOWER)
  
  return(data)
}

# Apply the conversion function
aqim_data <- convert_to_kg(aqim_data, quantity_units_to_kg)

# Check how many records have missing TOTAL_KG values
missing_kg_count <- sum(is.na((aqim_data |> filter(QUANTITY > 0))$TOTAL_KG))
total_records <- nrow(aqim_data |> filter(QUANTITY > 0))
missing_percent <- round(missing_kg_count / total_records * 100, 2)

# Print summary of conversion results
cat(sprintf("Conversion complete: %d out of %d records (%.2f%%) have valid kg values.\n", 
            total_records - missing_kg_count, total_records, 100 - missing_percent))

# If there are missing values, show the top unit types that couldn't be converted
if(missing_kg_count > 0) {
  cat("Top unit types without kg conversion:\n")
  aqim_data |> filter(QUANTITY > 0) %>% 
    filter(is.na(TOTAL_KG)) %>%
    count(QUANTITY_UNITS_NAME, sort = TRUE) %>%
    head(8) %>%
    print()
}
```



# Create a reference table that identifies what entry type each PATHWAY is associated with

```{r}

setwd(here::here('input', 'data', 'PATHWAY name reference'))
# Load the pathway reference table
pathway_ref <- read.csv("pathway to entry type reference table.csv", stringsAsFactors = FALSE)
aqim_data <- aqim_data %>%
  left_join(pathway_ref, by = "PATHWAY", relationship = "many-to-one")

get_entry_type <- function(pathway_str) {
  for (i in 1:nrow(pathway_ref)) {
    ref_pathway <- pathway_ref$PATHWAY[i]
    if (grepl(ref_pathway, pathway_str, fixed = TRUE)) {
      return(pathway_ref$entry_type[i])
    }
  }
  return(NA_character_)
}

# Apply the function to each pathway in arm_aqim_records
system.time(aqim_data <- aqim_data %>%
  rowwise() %>%
  mutate(entry_type = get_entry_type(PATHWAY)) %>%
  ungroup()
)

```

# Assign location 
```{r}
# db_conn <- dbConnect(odbc::odbc(),.connection_string = 
#                       "Driver=SQL Server;
#                         Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
#                         Database=PPQ_AQI_ARMDMV2;
#                         trusted_connection=yes")

# inspection_cols <- c(
#     "ID", 
#     "CATEGORY",
#     "NAME", 
#     "PARENT_LOCATION_ID", 
#     "SITE", 
#     "LOCATION_STATE_CD", 
#     "PATHWAY_ID", 
#     "PATHWAY"
#     )


# #-- Pull AQIM inspection records from ARM  ---- 
# system.time(arm_aqim <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_LOCATION]")) |> 
#               dplyr::filter(CATEGORY == "AQIM") |>
#               dplyr::select(inspection_cols) |> 
#               #rename to fit AQAS data structure
#               rename(edge_origin = COUNTRY_OF_ORIGIN_NAME,
#                       group_size = NUMBER_OF_PASSENGERS
#            )
# )
 

```

```{r }
# Function to match commodities with ASF reference table
match_asf_risk <- function(data, reference_table) {
  # Try matching on COMMODITY_DISPLAY_NAME
  asf_product_match <- data %>%
    inner_join(reference_table, 
               by = c("COMMODITY_DISPLAY_NAME" = "ITEM", "entry_type" = "entry_type"), 
               relationship = "many-to-one")
  
  # Combine matches and remove duplicates
  asf_matches <- asf_product_match %>%
    distinct(ID, COMMODITY_ID, .keep_all = TRUE)
  
  # Return the matches for further processing
  return(asf_matches)
}

# Apply to both datasets
asf_matches_aqim <- match_asf_risk(aqim_data, asf_product_ref)
# asf_matches_arm <- match_asf_risk(arm_aqim_records, asf_product_ref)

# # Combine all matches from both sources
# asf_matches <- bind_rows(asf_matches_aqim, asf_matches_arm) %>%
#   distinct(ID, COMMODITY_ID, .keep_all = TRUE)

# Apply risk classification to the full dataset
dat <- aqim_data %>% 
  left_join(asf_matches_aqim, relationship = "many-to-many") %>%
  mutate(IS_ASF_RISK = ifelse(is.na(IS_ASF_RISK), "NO", IS_ASF_RISK))

# # Apply risk classification to ARM records separately
# dat_arm <- arm_aqim_records %>% 
#   left_join(asf_matches, by = c("ID", "COMMODITY_ID"), relationship = "many-to-many") %>% 
#   mutate(IS_ASF_RISK = ifelse(is.na(IS_ASF_RISK), "NO", IS_ASF_RISK))

```
## Convert the entry_type to commercial if size is large

```{r if large interceptions then convert product_subpathway to commercial}
#Use the reference table to convert entry_type to commercial if the size is large
setwd(here::here('input', 'data', 'product pathway by volume key'))
size_ref <- read.csv(paste0(params$pathway_volume_smuggling_reference_table_filename,".csv"), stringsAsFactors = FALSE)

size_ref_func <- function(data, reference_table){
    data %>%
  left_join(reference_table, by = c("entry_type" = "entry_type_ref"), relationship = "many-to-one") %>%
  mutate(
    product_subpathway = ifelse(IS_ASF_RISK != "NO" & !is.na(size_threshold_kg
) & !is.na(TOTAL_KG) & TOTAL_KG >= size_threshold_kg
, subpathway, product_subpathway)
  ) }

dat <- size_ref_func(data = dat, reference_table = size_ref)
View(dat |> filter(grepl("smuggled", product_subpathway)))
```

# # trouble shooting
#  ```{r trouble shooting}
# #Only one passenger in SAN FRANCISCO with a large quantity of pork that would be considered smuggled
# View(aqim_data |> filter(grepl("SAN FRANCISCO",INSPECTION_LOCATION_NAME)) |> count(PATHWAY, entry_type, TOTAL_KG))
# # so, the 0.03 approach rate estimate is not correct
# dat |> filter(grepl("SAN FRANCISCO",INSPECTION_LOCATION_NAME)) |> View()


```{r save aqim data}

write.csv(dat, "C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv", row.names = FALSE)

```
