---
title: "aqim pathway risk by PoE"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  wads_volume_filename: "wads_passenger_volume_by_port_pathway_month_year" # Name of the output file for WADS volume data
  glm_results_filename: "glmm_results_aqim fraction at risk by pathway and location"
  bootstrap_results_filename: "bootstrap_results_aqim fraction at risk by pathway and location"
  aqim_qmi_size_distribution_filename: "pathway_weight_distribution_models" # Name of the output file for the distribution models of the weight of items observed in the AQIM pathways
  group_size_filename: "aqim_mean_pathway_group_size"
  # Set number of simulations
  n_sims_for_volume: 100
  pathway_volumes_filename: "volume_of_risk_by_pathway" # Name of the output file for the pathway risk summary results
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      wads_volume_filename = "wads_passenger_volume_by_port_pathway_month_year",
      glm_results_filename = "glmm_results_aqim fraction at risk by pathway and location",
      bootstrap_results_filename = "bootstrap_results_aqim fraction at risk by pathway and location",
      aqim_qmi_size_distribution_filename = "pathway_weight_distribution_models",
      group_size_filename = "aqim_mean_pathway_group_size",
      n_sims_for_volume = 100,
      pathway_volumes_filename = "volume_of_risk_by_pathway"
    )
  }
}
```

# Read WADS volume data
```{r read wads volume data}
setwd(here::here("output","wads","arrivals_or_entries"))
wads_volume <- readRDS(paste(params$wads_volume_filename, ".rds"))
```

# read approach rates
```{r read approach rates}
setwd(here::here('output','aqim pathway approach rates')) 
location_probabilities_glmm <- readRDS(paste0(params$glm_results_filename))
```
# Join WADS volume and approach rates

```{r join wads volume and approach rates}
#for every entry in wads_volume where PATHWAY exists in location_probabilities_glmm, then all possible `group` values in location_probabilities_glmm of that PATHWAY can exist. 


# Step 1: Create a unique identifier for each row in wads_volume
wads_volume <- wads_volume %>%
  mutate(row_id = row_number())

# Step 2: Get all group combinations for each pathway regardless of location
pathway_group_combinations <-
    location_probabilities_glmm |> dplyr::select(pathway, group) |> unique()

wads_volume <- wads_volume |> left_join(pathway_group_combinations, by = c("PATHWAY" = "pathway"), relationship = "many-to-many")

# Step 3: Match estimated probabilities of product type to any pathway, group,  and location matches
wads_volume_joined <- wads_volume %>%
  left_join(
    location_probabilities_glmm, 
    by = c("PATHWAY" = "pathway", "LOCATION_NAME" = "location", "group" = "group"),
    relationship = "many-to-many"
  )


#Split the group column into its component parts
wads_volume_joined <- wads_volume_joined %>%
  separate(group, 
           into = c("PATHWAY_check", "container", "product_pathway", "product_subpathway"),
           sep = ";",
           remove = FALSE)

wads_volume_joined <- wads_volume_joined %>%  # Keep original group column
  mutate(
    # Verify PATHWAY matches with split value
    pathway_match = PATHWAY == PATHWAY_check
  ) %>%
  dplyr::select(-PATHWAY_check)  # Remove the redundant PATHWAY column

# # If a location_name is not in the location_probabilities_glmm, then if pathway is not NA and  the probability is not known (i.e., the LOCATION_NAME was not in the model parameters), then the probability should equivalent to the pathway when LOCATION_NAME == "Average"
# #the significant column is whether it is different from the baseline, which is not relevant
# wads_volume_joined <- wads_volume_joined |> rowwise() |>
#   mutate(probability = ifelse(!is.na(PATHWAY) & is.na(probability), 
#                               location_probabilities_glmm$probability[location_probabilities_glmm$location == "Average" & location_probabilities_glmm$pathway == PATHWAY], 
#                               probability))
# #View(wads_volume_joined)

# wads_volume_joined |> 
#   mutate(probability = ifelse(is.na(probability), 0, probability)) # replace NA probabilities with 0


# Pre-compute average probabilities for all pathways once
average_probs <- location_probabilities_glmm %>%
  filter(location == "Average") %>%
  dplyr::select(pathway, group, probability) %>%
  rename(avg_probability = probability)

# Join this to the main dataset in one vectorized operation
wads_volume_joined <- wads_volume_joined %>% 
  left_join(average_probs, by = c("PATHWAY" = "pathway", "group" = "group")) %>%
  mutate(
    # Apply the decision rules using vectorized operations
    probability = case_when(
      # If pathway is NA, keep original probability
      is.na(PATHWAY) ~ probability,
      
      # If original probability is NA, use average
      is.na(probability) ~ avg_probability,
      
      # If original probability is 0, take the average
      !is.na(probability) & probability < 1e-7 & !is.na(avg_probability) ~ avg_probability,

      # # I was originally taking the average probability if the average was within the CI range (see below lines), but CI is now 0-1 for everything, so that is useless. Instead, I changed it to take the average if its really low
      # # If average is WITHIN the confidence interval, use average
      # !is.na(avg_probability) & !is.na(ci_lower) & !is.na(ci_upper) & 
      #   avg_probability >= ci_lower & avg_probability <= ci_upper ~ avg_probability,
      
      # Otherwise keep the original probability (when average is outside CI)
      TRUE ~ probability
    )
  )

```

# retrieve pathway weight distribution models
```{r load pathway weight distribution models}
setwd(here::here("output","item size sampling distributions","aqim"))
pathway_distributions <- readRDS(paste(params$aqim_qmi_size_distribution_filename, ".rds", sep = ""))
```

## remove non-existing combinations of pathway, container, product_pathway, product_subpathway



```{r create weight sampling function}
# Create a more robust sample_weight function that handles missing combinations
sample_weight <- function(n, pathway, container, prod_pathway, prod_subpathway, 
                         distribution_models = pathway_distributions) {
  
  # Check if the combination exists in the distribution models
  matching_dist <- distribution_models %>%
    filter(PATHWAY == pathway, 
           container == container,
           product_pathway == prod_pathway,
           product_subpathway == prod_subpathway)
  
  # If combination doesn't exist, return zeros (effectively zeroing out)
  if(nrow(matching_dist) == 0) {
    return(rep(0, n))  # Return zeros instead of NA to ensure zero total weight
  }
  
  # Otherwise proceed with sampling as before
  dist_model <- matching_dist %>% pull(distribution_model) %>% .[[1]]
  
  # Sample from the distribution
  if(dist_model$distribution == "empirical") {
    return(sample(dist_model$data, size = n, replace = TRUE))
  } else if(dist_model$distribution == "lnorm") {
    return(rlnorm(n, meanlog = dist_model$params$meanlog, sdlog = dist_model$params$sdlog))
  } else if(dist_model$distribution == "gamma") {
    return(rgamma(n, shape = dist_model$params$shape, rate = dist_model$params$rate))
  } else if(dist_model$distribution == "weibull") {
    return(rweibull(n, shape = dist_model$params$shape, scale = dist_model$params$scale))
  } else {
    return(rep(0, n))  # Default to zeros for unknown distributions
  }
}
```

## Modify the count based on aqim data and average group sizes
This is really only relevant for passenger pathways, the others will end up with a group size of 1
```{r modify count based on aqim data and average group sizes}
# Join average group sizes to wads_volume_joined
setwd(here::here('output','aqim pathway approach rates'))
group_size_summary <- readRDS(paste(params$group_size_filename,".rds", sep=""))
wads_volume_joined <- wads_volume_joined %>%
  left_join(group_size_summary, by = c("PATHWAY" = "PATHWAY")) %>%
  mutate(
    # If avg_group_size is NA (no data), default to 1
    avg_group_size = ifelse(is.na(avg_group_size) | avg_group_size < 1, 1, avg_group_size),
    # Adjust COUNT based on mean group size
    adjusted_COUNT = round(COUNT / avg_group_size, 0)
  ) %>%
  dplyr::select(-avg_group_size)  # Remove avg_group_size if not needed further
```



```{r}
# Create storage for simulation results
simulation_results <- list()

# Run simulations
for(i in 1:params$n_sims_for_volume) {
  # Run a complete simulation for this iteration
  sim_result <- wads_volume_joined %>%
    rowwise() %>%
    mutate(
      # Simulate number of positive events
      positive_count_sim = ifelse(
        is.na(probability) | is.na(adjusted_COUNT) | probability == 0 | adjusted_COUNT == 0, 
        0,
        rbinom(n = 1, size = adjusted_COUNT, prob = probability)
      ),
      
      # Calculate total kg at risk for this iteration - safely handle missing combinations
      total_kg_sim = tryCatch({
        if(positive_count_sim > 0 && !is.na(PATHWAY)) {
          # Sample weights based on the distribution - this will return zeros for missing combinations
          weights <- sample_weight(
            n = positive_count_sim,
            pathway = PATHWAY,
            container = container,
            prod_pathway = product_pathway,
            prod_subpathway = product_subpathway
          )
          sum(weights, na.rm = TRUE)
        } else {
          0
        }
      }, error = function(e) {
        warning(paste("Error in row with PATHWAY:", PATHWAY, 
                     "container:", container, 
                     "product_pathway:", product_pathway, 
                     "product_subpathway:", product_subpathway,
                     "- Error:", e$message))
        0  # Return 0 on error
      }),
      
      # Add simulation ID
      sim_id = i
    ) %>%
    ungroup()
  
  # Store this simulation's results
  simulation_results[[i]] <- sim_result %>%
    dplyr::select(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, container, product_pathway, product_subpathway, year, month, probability, COUNT, adjusted_COUNT, positive_count_sim, total_kg_sim, sim_id)
}


# Combine all simulation results
all_simulations <- bind_rows(simulation_results)

# Calculate summary statistics by pathway/container/product group
pathway_risk_summary <- all_simulations %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, container, product_pathway, product_subpathway, year, probability, month, sim_id) %>%
  # Sum within each pathway group for each simulation
  summarize(
    total_positives = sum(positive_count_sim, na.rm = TRUE),
    total_kg = sum(total_kg_sim, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Now calculate statistics across simulations
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, container, product_pathway, product_subpathway, year, probability) %>%
  summarize(
    mean_positives = mean(total_positives),
    lower_ci_positives = quantile(total_positives, 0.05),
    upper_ci_positives = quantile(total_positives, 0.95),
    mean_kg = mean(total_kg),
    lower_ci_kg = quantile(total_kg, 0.05),
    upper_ci_kg = quantile(total_kg, 0.95),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_kg))

print("Do pathways with at risk items having been found have 0 probability? For instance, there are none found in the pedestrian SB pathway.")
print("Ah - yes, if running with current data, pedestrian SB has no positive events in AQIM data, so probability is 0. That is correct.")

#write results immediately because it takes a while to run
setwd(here::here("output","aqim based volume of at risk product approaching"))
saveRDS(pathway_risk_summary, paste(params$pathway_volumes_filename, params$n_sims_for_volume, "sims",".rds", sep = ""))

pathway_risk_summary <- readRDS(paste(params$pathway_volumes_filename, params$n_sims_for_volume, "sims",".rds", sep = ""))
```




## Inspect why smuggled commercial products pathway has such a high probability
```{r}
# The highest estimate is SF int'l airport, air passenger, commercial smuggled products, baggage, 2024; 734,424 people (adjusted to 521708).  probability is 0.0311. Mean_positives is 16218, mean kg per distirbution is 32kg;
#This leads to a total estimate of ~500k kg of at risk product approaching in this month from this one WADS entry alone
mean(sample_weight(n=100, pathway = "Airport - Aircraft - Person - CBP", container = "baggage", prod_pathway = "swine products/by-products", prod_subpathway = "smuggled commercial products"))

# 3 in 100 people with smuggled products seems to be a ludicrous Check that this smuggled commercial products pathway is supposed to be such a high probability
# Check how many actual positive events were found in this pathway in AQIM data
dat <- read.csv("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv")

smuggled_prods_air <- dat %>%
  filter(PATHWAY == "Airport - Aircraft - Person - CBP",
         product_pathway == "swine products/by-products",
         product_subpathway == "smuggled commercial products") %>%
  count(CALENDAR_YEAR, IS_ASF_RISK) %>%
  arrange(CALENDAR_YEAR, desc(IS_ASF_RISK))
sum(smuggled_prods_air$n)

smuggled_prods_model <- location_probabilities_glmm %>%
  filter(pathway == "Airport - Aircraft - Person - CBP",
         grepl("smuggled commercial products", group))

# Looking at 01_03, I think this is due to a naming convention - the code does not specify that the product_pathway / subpathway when it converts to smuggled commercial products, so I think somewhere along the code that it is getting added

# Probably just need to fix the lookup table to correct the product pathway/subpathway for smuggled commercial products in `data_for_risk`

# 8/28 - Yes,  I think I fixed it. I added an IS_ASF_RISK != "NO" condition so that it wouldn't assign smuggling subpathway to thing sthat weren't of risk. I think it was keeping this subpathway throughout even though it shouldn't have been. WIll have to re-calculate approach rates when I leave today.
```








```{r sample events from distriution models of weight per event}
# risk_weights_by_pathway <- risk_weights_by_pathway_tmp %>%
#   # Add mutate() here
#   mutate(
#     total_kg_at_risk = case_when(
#       # Only process if we have events with risk and can find the pathway
#       events_with_risk > 0 & !is.na(PATHWAY) ~ {
#         # Rest of your case_when logic remains the same
#         dist_model <- pathway_distributions %>%
#           filter(PATHWAY == .data$PATHWAY,
#           container == .data$container,
#           product_pathway == .data$product_pathway,
#           product_subpathway == .data$product_subpathway) %>%
#           pull(distribution_model) %>%
#           first()

# #test
# pathway_distributions %>%
#           filter(PATHWAY == 'Express_courier',
#           container == 'baggage',
#           product_pathway == .data$'swine products/by-products',
#           product_subpathway == 'personal use or consumption') %>%
#           pull(distribution_model) %>%
#           first()

# #end test

#         # If we found a distribution model, sample from it
#         if(!is.null(dist_model) && length(dist_model) > 0) {
#           # Sample weights based on the distribution model
#           sampled_weights <- if(dist_model$distribution == "empirical") {
#             sample(dist_model$data, size = events_with_risk, replace = TRUE)
#           } else if(dist_model$distribution == "lnorm") {
#             rlnorm(events_with_risk, 
#                    meanlog = dist_model$params$meanlog, 
#                    sdlog = dist_model$params$sdlog)
#           } else if(dist_model$distribution == "gamma") {
#             rgamma(events_with_risk, 
#                    shape = dist_model$params$shape, 
#                    rate = dist_model$params$rate)
#           } else if(dist_model$distribution == "weibull") {
#             rweibull(events_with_risk, 
#                      shape = dist_model$params$shape, 
#                      scale = dist_model$params$scale)
#           } else {
#             # Default case if distribution type unknown
#             rep(NA_real_, events_with_risk)
#           }
          
#           # Sum the sampled weights
#           sum(sampled_weights, na.rm = TRUE)
#         } else {
#           NA_real_  # No distribution model found
#         }
#       },
#       TRUE ~ 0  # No events with risk
#     )
#   ) %>%
#   ungroup()

# # View the results
# summary(risk_weights_by_pathway$total_kg_at_risk)
```