---
title: "EAN data"
author: "Dalon White"
date: "2025-09-12"
output: 


params:
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  affected_product_file: "asf affected product name reference table.csv"
  affected_country_file: "asf affected countries.csv"
  pathway_volume_smuggling_reference_table_filename: "product pathway by volume reference table"
  quantity_units_to_kg: "quantity units to kilograms reference table.csv"
---

-   prepped data already has the pathway, entry type, etc., and risk as no/low/med/high
-   need to use country of origin and product origin to determine if the inspection has any asf risk
-   then model the kg of product with asf risk by pathway, entry type, location, country of origin, product pathway, product subpathway

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(lme4)
library(ggplot2)
library(broom)
library(multcomp)
library(emmeans)
library(sjPlot)
library(broom.mixed)
library(boot)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      affected_countries = "current", # "current" or "mex_can" or "other" 
      years = 2023:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"),
      affected_product_file = "asf affected product name reference table.csv",
      affected_country_file = "asf affected countries.csv",
      pathway_volume_smuggling_reference_table_filename = "product pathway by volume reference table",
      quantity_units_to_kg = "quantity units to kilograms reference table.csv"
    )
  }
}
```

```{r}

#open connection
db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

```

```{r ARM AQIM data}
ean_cols <- c(
  "ID",
  "COMMODITY_ID",
   "CONVEYANCE_ID",
  "ISSUED_DATETIME",
  "ISSUED_DATETIME_FISCAL_YEAR",
  "ISSUED_DATETIME_FISCAL_MONTH",
  "CATEGORY",
  "SUBCATEGORY",
  "PATHWAY_ID", #PATHWAY ID can distinguish commerce and express carrier; probably need a reference table because the naming conventions aren't consistent
  "PATHWAY",
   "REMARKS",
   "DESTINATION_OF_ARTICLES_ID",
     "REGULATORY_ACTION_TYPE"
)

conveyance_cols <- c(
  "ID",
  "INSPECTION_ID",
  "CONVEYANCE_TYPE",
  "AIRCRAFT_TYPE",
  "VEHICLE_TYPE" #if !NULL, then it's a vehicle. Need to explore if it includes cargo trucks
)

inspection_cols <- c(
  "ID",

  "INSPECTION_NUMBER", 
  "IS_AQIM", 
  "PORT_OF_ENTRY_ID",
  
  "INSPECTION_DATETIME", 
  "INSPECTION_DATETIME_FISCAL_YEAR", 
  "INSPECTION_DATETIME_FISCAL_MONTH",
  "INSPECTION_LOCATION_ID", 
  "INSPECTION_LOCATION_NAME", 
  "INSPECTION_LOCATION_STATE_CODE", 
  "COUNTRY_OF_ORIGIN_NAME", 
 # "COUNTRY_OF_RESIDENCE_NAME", 
  
# All of these destination fields are empty for EANs; I checked.
 # "FINAL_DESTINATION_STATE_NAME", 
 # "FINAL_DESTINATION_STATE_ID", 
 # "FINAL_DESTINATION_CITY", 
 # "PASSENGER_DISTANCE_DESTINATION_NAME",
  # "CARGO_DESTINATION_ZIP_CODE",
  # "POSTMARK_COUNTRY_NAME",
  # "STATE_OF_ORIGIN_NAME",
  # "SUBSITE_LOCATION_NAME",
 
  "NUMBER_OF_PASSENGERS"
  )

commodity_cols <- c(
  "ID",
  "INSPECTION_ID",
  "IS_SMUGGLED",
  "IS_CONTAMINANT_FOUND",
  "COMMODITY_CLASSIFICATION",
  "COUNTRY_OF_ORIGIN_NAME", 
  "REF_COMMODITY_ID",
  "COMMODITY_DISPLAY_NAME",
  "MISC_ANIMAL_COMMODITY_NAME",
  "QUANTITY",
  "QUANTITY_UNITS_NAME"
)

#columns from the brg_pathway table that have more colloquial explanations to PATHWAY
pathway_cols <- c(
  "PATHWAY",
  "DISPLAY_NAME",
  "PORT_TYPE",
  "INSPECTION_TARGET",
  "MODE_OF_TRANSPORT",
  "DEFAULT_EXCLUSION_OPTIONS"
)
#-- Pull AQIM inspection records from ARM  ---- 
ean <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_EAN]")) |> 
              dplyr::select(all_of(ean_cols)) |> 
              #rename to fit AQAS data structure
              rename(EAN_ID = ID)

inspections <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_INSPECTION]")) |> 
              dplyr::filter(CATEGORY != "AQIM") |>
              dplyr::select(all_of(inspection_cols)) |>
              rename(INSPECTION_ID = ID,
                    #rename to match AQAS data structure
                     product_origin = COUNTRY_OF_ORIGIN_NAME,
                     LOCATION_ID = "INSPECTION_LOCATION_ID", 
                     LOCATION_NAME = "INSPECTION_LOCATION_NAME", 
                     LOCATION_STATE_CODE = "INSPECTION_LOCATION_STATE_CODE"
)

#-- Load Bridge Commodity dataset ----
commodity <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |> 
              dplyr::select(all_of(commodity_cols)) |>
              rename(COMMODITY_ID = ID)

conveyance <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_CONVEYANCE]")) |>
              dplyr::select(all_of(conveyance_cols)) |>
              rename(CONVEYANCE_ID = ID)

#-- Load Bridge Pathway dataset ----
pathway_tbl <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_PATHWAY]")) |> 
              dplyr::select(all_of(pathway_cols))

# Join ean records with commodity table info
## Identify columns to join by
intersect_cols <- intersect(colnames(ean), colnames(commodity))
ean_records_tmp1 <- ean |>
  left_join(commodity, by = intersect_cols, relationship = "many-to-many")

# Join ean and commodity with inspection details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records_tmp1), colnames(inspections))
ean_records_tmp2 <- ean_records_tmp1 |>
  left_join(inspections, by = intersect_cols, relationship = "many-to-many") 

# Join ean, commodity, and inspection details with conveyance details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records_tmp2), colnames(conveyance))
ean_records_tmp3 <- ean_records_tmp2 |>
  left_join(conveyance, by = intersect_cols, relationship = "many-to-many")

# Join ean, commodity, inspection, and conveyance details with pathway details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records_tmp3), colnames(pathway_tbl))
ean_records <- ean_records_tmp3 |>
  left_join(pathway_tbl, by = intersect_cols, relationship = "many-to-many")

#using intersect columns above will ensure that multiple overlapping columns won't create a .x and .y version of the same column

```

# Collect data
I have compared filtering in the database vs. collecting and filtering in R. Collecting is only a couple of seconds, so it's fast enough and filtering is much easier for some operations that require complex manipulations or functions not supported by the SQL. The filtering by datetime is particularly easier when done in R due to the need for lubridate functions and having to otherwise manipulate the datetime fiscal year & month columns separately.

```{r}
# Collect data into R
system.time(ean_records <- ean_records |> collect())

# Disconnect from the database
dbDisconnect(db_conn)
```


# filter to relevant records
## filter by years

### see 00_03 to find why filter by INSPECTION_DATETIME then ISSUED_DATETIME

In most cases, it looks like inspection datetime and issued datetime are pretty much the same, with 5% delayed EAN-issuance dates >7 days later than the inspection date
So, results suggest filter by INSPECTION_DATETIME first then by ISSUED_DATETIME if needed

Additionally: Filter using the params$years:
  Will call the years parameter here. The years parameter is used for the WADS codes when calculating the volume for AQIM-based pathways. By calling it here, this will ensure that we are comparing predicted volumes through AQIM+WADS for a given time to the volume of EAN-related products for the same time frame.

### Filter by INSPECTION_DATETIME then ISSUED_DATETIME
```{r}
# Filter to records within specified years
filter_years <- function(data) {
  # First ensure we have at least one datetime field
  filtered_data <- data %>% 
    filter(!is.na(INSPECTION_DATETIME) | !is.na(ISSUED_DATETIME))
  
  # Extract years from datetime fields
  filtered_data <- filtered_data %>%
    mutate(
      inspection_year = year(INSPECTION_DATETIME),
      issued_year = year(ISSUED_DATETIME)
    )
  
  # Filter by year using logical OR approach
  filtered_data <- filtered_data %>%
    filter(
      # Keep if INSPECTION_DATETIME year is in range
      (!is.na(inspection_year) & inspection_year %in% params$years) |
      # OR if INSPECTION_DATETIME is NA but ISSUED_DATETIME year is in range
      (is.na(inspection_year) & !is.na(issued_year) & issued_year %in% params$years)
    )
  
  # Remove the temporary year columns
  filtered_data <- filtered_data %>%
    dplyr::select(-inspection_year, -issued_year)
  
  return(filtered_data)
}

# Apply the function
ean_records <- filter_years(ean_records)
```

## separate the data into layers

This is to separate the data into layers that can be used to model risk. The layers are provided i nthe entry type reference table. This converts the PATHWAY to specific modes of entry
 - passenger
 - eco
 - mail
 - commercial
 - unknown
"unknown" need to be defined. There's 'Ambiguous', and some non-port/non-transport that end in EPP, DPP, and EC. Not sure if any are relevant to this analysis. For now they are coded just as unknown in the reference file.

Load pathway to entry type reference table:

```{r load pathway to entry type reference table}
setwd(here::here('input', 'data', 'PATHWAY name reference'))
# Load the pathway reference table
pathway_ref <- read.csv("pathway to entry type reference table.csv", stringsAsFactors = FALSE)
ean_records <- ean_records %>%
  left_join(pathway_ref, by = "PATHWAY", relationship = "many-to-one")

get_entry_type <- function(pathway_str) {
  for (i in 1:nrow(pathway_ref)) {
    ref_pathway <- pathway_ref$PATHWAY[i]
    if (grepl(ref_pathway, pathway_str, fixed = TRUE)) {
      return(pathway_ref$entry_type[i])
    }
  }
  return(NA_character_)
}

# Apply the function to each pathway in ean_records
ean_records <- ean_records %>%
  rowwise() %>%
  mutate(entry_type = get_entry_type(PATHWAY)) %>%
  ungroup()
```

### use brg_pathway to separate

Conveyance table can separate out private and commercial aircraft
- this doesn't work within the 2023-2024 time frame; leaving it because it might work for some other time frames, like in the future. Not sure. See 00_04 for more details.
```{r}
ean_records <- ean_records |> mutate(MODE_OF_TRANSPORT = 
case_when(
  CONVEYANCE_TYPE == "AIRCRAFT" & !is.na(AIRCRAFT_TYPE) ~ paste(MODE_OF_TRANSPORT, AIRCRAFT_TYPE, sep = " - "),
  .default = MODE_OF_TRANSPORT
  ))
```


# Determine if there is risk

Next step is to determine if there is any risk associated with the EAN record. This is through both the country of origin and the commodity type.

## Evaluate country
 - this only makes sense to use the current list of asf affected countries; otherwise, EANs wouldn't exist.
load country reference list
```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
```

filter to  eans that are from that country
```{r isolate records from risky countries}
ean_records <- ean_records |> mutate(country_code = countrycode::countrycode(COUNTRY_OF_ORIGIN_NAME, 
                                  origin = "country.name", 
                                  destination = "iso3c"),
                                  product_origin_code = countrycode::countrycode(product_origin, 
                                  origin = "country.name",
                                  destination = "iso3c")
                                  )

ean_records <- ean_records |> filter(country_code %in% affected_countries |
                                    product_origin_code %in% affected_countries)
```


## Evaluate product
The type of product will be stored in COMMODITY_DISPLAY_NAME, just like in AQIM data.

```{r what kind of products descriptions for eans}
 View(ean_records |> filter(grepl("swine|porcine|pork|hooves|meat|ASF|CSF|blood|manure|dirt|feed",tolower(COMMODITY_DISPLAY_NAME))) |> count(REMARKS))

print("most dont have a descriptive COMMODITY_DISPLAY_NAME like AQIM data")

print("instead, they all use 'Ruminant or Swine...', so we'll have to use the REMARKS field to identify relevant products, where available")
```

```{r save for remarks analysis}
setwd(here::here("output","intermediate files"))
saveRDS(ean_records, "ean_records.rds")

print("Remarks analysis done outside of this analysis, using Excel, M365 Copilot, and Ernesto R.")
```

# Match with the product reference list, which incldues ARM's typical match for EANs, which begins with "Ruminant or Swine..."
load product reference list
```{r identify products of risk}
setwd(here::here("input","data","asf affected product names"))
asf_product_ref <- read.csv(params$affected_product_file)
match_asf_risk <- function(data, reference_table) {
  # Try matching on COMMODITY_DISPLAY_NAME
  asf_product_match <- data %>%
    inner_join(reference_table, 
               by = c("COMMODITY_DISPLAY_NAME" = "ITEM", "entry_type" = "entry_type"), 
               relationship = "many-to-many",
               copy = TRUE)
  
  # Combine matches and remove duplicates
  asf_matches <- asf_product_match %>%
    distinct(EAN_ID, COMMODITY_ID, .keep_all = TRUE)
  
  # Return the matches for further processing
  return(asf_matches)
}

# Apply to eans
at_risk_eans <- match_asf_risk(ean_records, asf_product_ref)
```


# Convert non-kg quantities to kg units

```{r correct quantity}
# read the reference table for unit conversions
setwd(here::here('input', 'data', 'asf affected product names'))
quantity_units_to_kg <- read.csv(params$quantity_units_to_kg, stringsAsFactors = FALSE)


# Function to convert various unit formats to standardized kg values
convert_to_kg <- function(data, reference_table) {
  # Ensure reference table unit names are lowercase for case-insensitive matching
  reference_table <- reference_table %>%
    mutate(unit_lower = tolower(unit))
  
  # First, standardize the data's unit names by removing plurals and converting to lowercase
  data <- data %>%
    mutate(
      # Make a lowercase version of units for matching
      QUANTITY_UNITS_LOWER = tolower(QUANTITY_UNITS_NAME),
      # Remove trailing 's' from units (e.g., "Kilograms" -> "Kilogram")
      QUANTITY_UNITS_LOWER = gsub("s$", "", QUANTITY_UNITS_LOWER),
      # Ensure quantity is positive
      QUANTITY = abs(as.numeric(QUANTITY))
    )
  
  # Join with reference table using the lowercase, singular version
  data <- data %>%
    left_join(
      reference_table %>% 
        mutate(unit_lower = gsub("s$", "", unit_lower)),
      by = c("QUANTITY_UNITS_LOWER" = "unit_lower"),
      relationship = "many-to-many"
    )

  # make sure estimated kilograms is read in as a number
  data <- data %>%
    mutate(est_kilograms = as.numeric(est_kilograms))
  
  # Apply conversion and handle missing values with common-sense defaults
  data <- data %>%
    mutate(
      TOTAL_KG = case_when(
        !is.na(est_kilograms) ~ QUANTITY * est_kilograms,
        # Fallback rules for missing conversions
        grepl("gram", QUANTITY_UNITS_LOWER) & !grepl("kilo", QUANTITY_UNITS_LOWER) ~ QUANTITY / 1000,
        grepl("kilo", QUANTITY_UNITS_LOWER) ~ QUANTITY,
        grepl("pound", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.453592,
        grepl("ounce", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.0283495,
        grepl("each", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.5, # Assuming 0.5kg per item
        grepl("piece", QUANTITY_UNITS_LOWER) ~ QUANTITY * 0.5,
        grepl("ton", QUANTITY_UNITS_LOWER) & !grepl("metric", QUANTITY_UNITS_LOWER) ~ QUANTITY * 907.185,
        grepl("ton", QUANTITY_UNITS_LOWER) & grepl("metric", QUANTITY_UNITS_LOWER) ~ QUANTITY * 1000,
        TRUE ~ NA_real_
      )
    ) %>%
    # Clean up temporary columns
    dplyr::select(-QUANTITY_UNITS_LOWER)
  
  return(data)
}

# Apply the conversion function
at_risk_eans <- convert_to_kg(at_risk_eans, quantity_units_to_kg)

# Check how many records have missing TOTAL_KG values
missing_kg_count <- sum(is.na((at_risk_eans |> filter(QUANTITY > 0))$TOTAL_KG))
total_records <- nrow(at_risk_eans |> filter(QUANTITY > 0))
missing_percent <- round(missing_kg_count / total_records * 100, 2)

# Print summary of conversion results
cat(sprintf("Conversion complete: %d out of %d records (%.2f%%) have valid kg values.\n", 
            total_records - missing_kg_count, total_records, 100 - missing_percent))

# If there are missing values, show the top unit types that couldn't be converted
if(missing_kg_count > 0) {
  cat("Top unit types without kg conversion:\n")
  at_risk_eans |> filter(QUANTITY > 0) %>% 
    filter(is.na(TOTAL_KG)) %>%
    count(QUANTITY_UNITS_NAME, sort = TRUE) %>%
    head(8) %>%
    print()
}
```

## Convert the entry_type to commercial if size is large

```{r if large interceptions then convert product_subpathway to commercial}
#Use the reference table to convert entry_type to commercial if the size is large
setwd(here::here('input', 'data', 'product pathway by volume key'))
size_ref <- read.csv(paste0(params$pathway_volume_smuggling_reference_table_filename,".csv"), stringsAsFactors = FALSE)

size_ref_func <- function(data, reference_table){
    data %>%
  left_join(reference_table, by = c("entry_type" = "entry_type_ref"), relationship = "many-to-one") %>%
  mutate(
    product_subpathway = ifelse(
      # IS_ASF_RISK != "NO" & # not used in EAN analysis in the same way as AQIM
      !is.na(size_threshold_kg
) & !is.na(TOTAL_KG) & TOTAL_KG >= size_threshold_kg
, subpathway, product_subpathway)
  ) }

at_risk_eans <- size_ref_func(data = at_risk_eans, reference_table = size_ref)

#Can take a look at the ones that were changed to smuggled commercial product
View(at_risk_eans |> filter(grepl("smuggled", product_subpathway)))
```


# Save the processed EAN records for further analysis
```{r save processed ean records}

write.csv(at_risk_eans, "C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/ean_records_processed.csv", row.names = FALSE)

saveRDS(at_risk_eans, "C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/ean_records_processed.csv")

print("Processed EAN records saved to large file data folder for further analysis.")
```
# Report out what unit types were considered to have ASF

```{r save unit types}
setwd(here::here('output','units reported in ARM'))
ean_units_summary <- at_risk_eans |> filter(TOTAL_KG > 0) |> count(QUANTITY_UNITS_NAME)
write.csv(ean_units_summary, paste0('units converted in EAN data_',params$affected_countries,'.csv'))
```
