---
title: "EAN Product Type Analysis"
author: "ASF Risk Framework"
date: "`r Sys.Date()`"
output: html_document
params:
  ean_data_path: "output/intermediate files/ean_records.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
# library(reticulate)
# library(jsonlite)

# # Check if huggingfaceR is installed, if not install it
# if (!requireNamespace("remotes", quietly = TRUE)) {
#   install.packages("remotes")
# }
# if (!requireNamespace("huggingfaceR", quietly = TRUE)) {
#   remotes::install_github("farach/huggingfaceR")
# }
# library(huggingfaceR)
```
```{r load params}
params <- list(
  ean_data_path = "output/intermediate files/ean_records.rds"
)

```



## Loading EAN Data

First, we'll load the EAN data that was processed in the previous script.

```{r load_data}

# Check if the file exists
ean_file_path <- here::here(params$ean_data_path)
if (!file.exists(ean_file_path)) {
  stop(paste("EAN data file not found at:", ean_file_path, 
             "Please run the '03_01 ean data.Rmd' script first."))
}

# Load the data
ean_records <- readRDS(ean_file_path)

# Print summary of the data
cat("Loaded", nrow(ean_records), "EAN records\n")
cat("Number of records with REMARKS:", sum(!is.na(ean_records$REMARKS)), "\n")

# Sample some remarks to see what we're working with
if (sum(!is.na(ean_records$REMARKS)) > 0) {
  set.seed(123)
  sample_remarks <- ean_records %>% 
    filter(!is.na(REMARKS)) %>% 
    sample_n(min(5, sum(!is.na(ean_records$REMARKS))))
  
  cat("\nSample REMARKS:\n")
  for (i in 1:nrow(sample_remarks)) {
    cat(i, ": ", sample_remarks$REMARKS[i], "\n\n")
  }
}
```

# Filter to remarks

```{r filter_remarks}
remarks <- ean_records %>%
  filter(!is.na(REMARKS) & REMARKS != "") |>

  filter(grepl("swine|pig|pork|sausage|bacon|ham|jerky|salami|charcuterie|lard|fat|offal|tripe|tallow|chitterlings|chitlins|headcheese|scrapple|blood pudding|blood sausage|hay|feed|fodder|forage|silage|alfalfa|clover|grass|legume|soy|corn|maize|wheat|barley|oat|rye|bran|mash|pellet|meal|cake|oilseed|oil cake|oil meal|sunflower|canola|rapeseed|cottonseed|flaxseed|linseed|sesame seed", 
               tolower(REMARKS))) |> pull(REMARKS)

write.csv(data.frame(REMARKS=remarks), file=here::here("output","ean data","ean remarks","ean remarks to be quantified.csv"), row.names=FALSE)

```
I have passed the file to Colin F. to try to process the remarks with m365 CoPilot
Colin returned the file with copilot having partially processed the remarks. I passed the remainder to Ernesto to complete.
Ernesto returned the completed file. I reviewed for accuracy and made minor edits. I handed it off to Ernesto to analyze outside of this process.