---
title: "EAN geolocation"
author: "Dalon White"
date: "2025-09-12"
output: 


params:
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  ean_geolocated_filename: "ean_records_geocoded"
---

```{r setup, include=FALSE}
# Load required packages
library(tidyverse)
library(magrittr)
library(ggplot2) 
library(lubridate)
library(readr)
library(stringdist)
library(purrr)
library(DBI)
library(odbc)
library(tidygeocoder)
library(sf)
library(tigris)
library(SpatialKDE)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      begin_date = "2000-01-01",
      end_date = "2024-12-31",
      US_grid_cell_size = 30000,
      ean_geolocated_filename = "ean_records_geocoded"
    )
  }
}
```



# Load the processed EAN records for further analysis
```{r read processed ean records}
at_risk_eans <- readRDS("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/ean_records_processed.csv")
```

# Get location

- If it is a SUBCATEGORY == 'CBP|PIS', then can use the port location data like in AQIM
- If it is a SUBCATEGORY == "SITC", then can drop " SITC" from INSPECTION_LOCATION_NAME and geocode that location with the INSPECTION_LOCATION_STATE_CODE

## Gather Geolocation Data
### 1. POEs

```{r geolocation-database-connect}
# Connect to ARM and GeoLocations databases
db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=PPQ_AQI_ARMDMV2;
                      trusted_connection=yes")

geo_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=GeoLocations;
                      trusted_connection=yes")

# Define columns to extract from each table
brg_loc_cols <- c("ID", "NAME", "SITE", "PHYSICAL_ADDRESS_ID", 
                 "MAIL_ADDRESS_ID", 'CBP_PORT_NUMBER', "IATA_CODE")
cbp_poe_cols <- c("CBP_NAME", "CBP_CODE", "LAT", "LON", "Port of Entry Name")
ppq_facilities_cols <- c("LocationName", "PPQCode", "City", "State", "Longitude", "Latitude")
```

Gather Coordinates From Different Sources

```{r gather-coordinates}
# Get location data from BRG2_LOCATION table
brg_loc <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_LOCATION]")) %>% 
           dplyr::select(all_of(brg_loc_cols)) %>% 
           collect()

# Get CBP Port of Entry data with coordinates
cbp_poe <- tbl(geo_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_CBP_POEs]")) %>% 
           dplyr::select(all_of(cbp_poe_cols)) %>% 
           collect()

# Get PPQ facilities data with coordinates
ppq_facilities <- tbl(db_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_PPQ_Facilities]")) %>% 
                 dplyr::select(all_of(ppq_facilities_cols)) %>% 
                 collect()

# Join pathway risk data with location reference
at_risk_eans_brg_loc <- at_risk_eans %>% 
                       left_join(brg_loc, by = c("LOCATION_ID" = "ID"))

# Step 1: Join with CBP POE data first (preferred source)
at_risk_eans_loc_and_poe <- at_risk_eans_brg_loc %>%
  left_join(cbp_poe, by = c("CBP_PORT_NUMBER" = "CBP_CODE"), relationship = "many-to-many") %>%
  rename(
    CBP_LAT = LAT,
    CBP_LON = LON,
    CBP_PORT_NAME = `Port of Entry Name`
  )

# Step 2: Join with PPQ facilities for rows still missing coordinates
at_risk_eans_geo <- at_risk_eans_loc_and_poe %>%
  left_join(
    ppq_facilities %>% 
      dplyr::select(LocationName, Latitude, Longitude) %>% 
      rename(
        PPQ_LAT = Latitude,
        PPQ_LON = Longitude
      ),
    by = c("LOCATION_NAME" = "LocationName")
  )

# Step 3: Create combined coordinate columns, prioritizing CBP data
at_risk_eans_geo <- at_risk_eans_geo %>%
  mutate(
    final_LAT = coalesce(CBP_LAT, PPQ_LAT),
    final_LON = coalesce(CBP_LON, PPQ_LON)
  )

# Step 4: For remaining missing coordinates, use IATA code if available
missing_coords <- at_risk_eans_geo %>%
  filter(is.na(final_LAT) | is.na(final_LON)) %>%
  filter(!is.na(IATA_CODE))

# Use tidygeocoder to look up IATA airport codes if available
if(nrow(missing_coords) > 0 && requireNamespace("tidygeocoder", quietly = TRUE)) {
  iata_coords <- missing_coords %>%
    dplyr::select(IATA_CODE) %>%
    distinct() %>%
    mutate(query = paste("airport", IATA_CODE)) %>%
    tidygeocoder::geocode(address = query, method = "osm")
  
  # Join the results back
  at_risk_eans_geo <- at_risk_eans_geo %>%
    left_join(
      iata_coords %>% 
        dplyr::select(IATA_CODE, lat, long) %>% 
        rename(IATA_LAT = lat, IATA_LON = long),
      by = "IATA_CODE"
    ) %>%
    mutate(
      final_LAT = coalesce(final_LAT, IATA_LAT),
      final_LON = coalesce(final_LON, IATA_LON)
    )
} else if(nrow(missing_coords) > 0) {
  cat("Found", nrow(missing_coords), "rows with IATA codes but no coordinates.\n")
  cat("Consider installing the 'tidygeocoder' package to look up these codes.\n")
  print(missing_coords %>% dplyr::select(IATA_CODE) %>% distinct())
}

# Close database connections when done
dbDisconnect(db_conn)
dbDisconnect(geo_conn)
```

### 2. SITC Locations

```{r sitc-geocode}
# Filter to SITC records
sitc_sites <- at_risk_eans %>%
  filter(SUBCATEGORY == "SITC" & !is.na(LOCATION_NAME) &LOCATION_NAME != "")
# Clean LOCATION_NAME by removing " SITC"
sitc_sites <- sitc_sites %>%
  mutate(cleaned_location = str_replace(LOCATION_NAME, " SITC", ""))
# Create a combined address for geocoding
sitc_sites <- sitc_sites %>%
  mutate(geocode_address = paste(cleaned_location, LOCATION_STATE_CODE, "USA", sep = ", "))

sitc_sites <- sitc_sites %>%
  dplyr::select(LOCATION_NAME, geocode_address) %>%
  distinct()
# Use tidygeocoder to geocode the addresses
if(requireNamespace("tidygeocoder", quietly = TRUE)) {
    sitc_sites <- sitc_sites %>%
        tidygeocoder::geocode(address = geocode_address, method = "osm") |>
        rename(SITC_LAT = lat, SITC_LON = long)
    
    # Check for any failed geocodes
    failed_geocodes <- sitc_sites %>%
        filter(is.na(SITC_LAT) | is.na(SITC_LON))
    
    if(nrow(failed_geocodes) > 0) {
        cat("Warning: Some SITC locations failed to geocode:\n")
        print(failed_geocodes %>% dplyr::select(LOCATION_NAME, LOCATION_STATE_CODE, geocode_address))
    }
    
    } else {
    cat("The 'tidygeocoder' package is not installed. Please install it to geocode SITC locations.\n")
    }
# Join geocoded coordinates back to the sitc_records
at_risk_eans_geo <- at_risk_eans_geo %>%
  left_join(sitc_sites, by = c("LOCATION_NAME")) |>
    mutate(
      final_LAT = coalesce(final_LAT, SITC_LAT),
      final_LON = coalesce(final_LON, SITC_LON)
    )
  
# summarize how many records still missing coordinates
at_risk_eans_geo %>%
  summarize(
    total_records = n(),
    missing_coords = sum(is.na(final_LAT) | is.na(final_LON))
  ) %>% print()
  # which category and subcategory are missing coordinates
at_risk_eans_geo %>%
  filter(is.na(final_LAT) | is.na(final_LON)) %>%  
    count(CATEGORY, SUBCATEGORY) %>% arrange(desc(n)) %>% print()
    # what are the unique inspection location names missing coordinates
at_risk_eans_geo %>%
  filter(is.na(final_LAT) | is.na(final_LON)) %>%  
    count(LOCATION_NAME) %>% arrange(desc(n)) %>% print()
# its just DHL in ATL missing coordinates

# Manually assign coordinates for DHL ATLANTA HUB (1791)
at_risk_eans_geo <- at_risk_eans_geo |> mutate(final_LAT = case_when(
    is.na(final_LAT) | is.na(final_LON) & LOCATION_NAME == "DHL ATLANTA HUB (1791)" ~ 33.6407,
    .default = final_LAT
)) |> mutate(final_LON = case_when(
    is.na(final_LON) & LOCATION_NAME == "DHL ATLANTA HUB (1791)" ~ -84.4277,
    .default = final_LON
))
```

# Save the results with geocoded coordinates
```{r save-geocoded-results}
saveRDS(at_risk_eans_geo, here::here("output", "intermediate files", params$ean_geolocated_filename))
cat("Geocoded EAN records saved to 'output/intermediate files/ean_records_geocoded.rds'\n")

write.csv(at_risk_eans_geo |> st_drop_geometry(), "C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/ean_records_processed_and_geolocated.csv", row.names = FALSE)

```



