---
title: "EAN geolocation"
author: "Dalon White"
date: "2025-09-12"
output: 


params:
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  ean_geolocated_filename: "ean_records_geocoded"
---


```{r setup, include=FALSE}
# Load required packages
library(tidyverse)
library(magrittr)
library(ggplot2) 
library(lubridate)
library(readr)
library(stringdist)
library(purrr)
library(DBI)
library(odbc)
library(tidygeocoder)
library(sf)
library(tigris)
library(SpatialKDE)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      begin_date = "2000-01-01",
      end_date = "2024-12-31",
      US_grid_cell_size = 30000,
      ean_geolocated_filename = "ean_records_geocoded"
    )
  }
}
```

# load data
```{r load ean dat}
at_risk_eans_geo <- readRDS(here::here("output", "intermediate files", params$ean_geolocated_filename))

```
# 1. Prepare Data for Geospatial Processing
## 1.1 Convert to Spatial Objects

```{r convert-to-spatial}
# Filter out rows without coordinates and convert to sf object
at_risk_eans_geo_sf <- at_risk_eans_geo %>% 
  filter(!is.na(final_LAT) & !is.na(final_LON)) %>% 
  sf::st_as_sf(coords = c("final_LON", "final_LAT"), crs = 4326) %>%
  # Transform to NAD83/Albers (EPSG:5070) for analysis
  sf::st_transform(crs = 5070)

# Save spatial object for future use
saveRDS(at_risk_eans_geo_sf, here::here("output", "ean data", 
                                       paste0(params$ean_geolocated_filename,"_sf")))
```

## 1.2 Filter to US Boundaries

```{r us-boundaries}
# Load US boundaries 
us_bounds_sf <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_bounds.rds"))
us_grid <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_grid.rds"))

# Filter spatial data to US boundaries
at_risk_eans_geo_sf_filtered <- at_risk_eans_geo_sf %>%
  st_join(us_bounds_sf, left = TRUE) 

# Report filtering statistics
total_points <- nrow(at_risk_eans_geo_sf)
kept_points <- nrow(at_risk_eans_geo_sf_filtered)
removed_points <- total_points - kept_points

cat(sprintf("Filtered to %d points within US boundaries (%d removed)\n", 
          kept_points, removed_points))
```

# 1.3 Summarize Data for Spatial Layers

```{r summarize-for-spatial-layers}
# Group by location and pathway combinations and summarize
spatial_layers_summary <- at_risk_eans_geo_sf_filtered %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, 
           container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(TOTAL_KG, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Filter out incomplete records
  filter(!is.na(container) & !is.na(product_pathway) & !is.na(product_subpathway))
```

# 7. Create and Save Spatial Layers Function

```{r create-spatial-layers-function}
# Function to create and save individual spatial layers for each pathway combination
create_spatial_layers <- function(summarized_data, save_files = FALSE, output_dir = NULL) {
  # Split data by pathway combinations
  split_data <- summarized_data %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    group_split()
  
  cat(sprintf("Created %d spatial layers for different pathway combinations\n", length(split_data)))
  
  # Create list to store the layers
  spatial_layers <- list()
  
  # Create scenario string that specifies EANs
  scenario_string <- "EAN"
  
  # Process each split
  for (i in seq_along(split_data)) {
    current_split <- split_data[[i]]
    
    # Get combination values for naming
    combo <- current_split %>%
      dplyr::select(PATHWAY, container, product_pathway, product_subpathway) %>%
      slice(1) %>%
      as.list()
    
    # Create clean name for layer using abbreviations
    layer_name <- paste(
      substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 27),
      substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3),
      sep = "_"
    )
    
    # Store in list with clean name
    spatial_layers[[layer_name]] <- current_split
    
    # Save to file if requested
    if (save_files && !is.null(output_dir)) {
      # Ensure output directory exists
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Convert to POINT geometry (required for GeoPackage)
      export_layer <- current_split %>%
        sf::st_cast("POINT")
      
      # Include scenario in filename - use GeoPackage format for Shiny compatibility
      file_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".gpkg"))
      
      # Try to save the file
      tryCatch({
        sf::st_write(export_layer, file_path, delete_layer = TRUE)
        cat(sprintf("Saved layer %d/%d: %s\n", i, length(split_data), file_path))
      }, error = function(e) {
        cat(sprintf("ERROR saving layer %d/%d: %s - %s\n", 
                   i, length(split_data), layer_name, e$message))
      })
    }
  }
    # Calculate totals directly from the original summarized data to ensure consistency
  # This approach guarantees the same results as the manual grouping and summarization
  layer_summary <- summarized_data %>%
    sf::st_drop_geometry() %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    summarize(
      total_risk = sum(total_mean_kg, na.rm = TRUE),
      point_count = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      layer_name = paste(
        substr(gsub("[^[:alnum:]]", "", PATHWAY), 1, 27),
        substr(gsub("[^[:alnum:]]", "", container), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_pathway), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_subpathway), 1, 3),
        sep = "_"
      ),
      scenario = scenario_string
    ) %>%
    dplyr::select(layer_name, scenario, point_count, total_risk) %>%
    arrange(desc(total_risk))
  
  # Print summary
  cat(paste0("\nSummary of spatial layers for scenario '", scenario_string, "' (top 10 by risk):\n"))
  print(head(layer_summary, 10))
  
  # Also save the summary table as CSV for reference
  if (save_files && !is.null(output_dir)) {
    summary_path <- file.path(output_dir, paste0("layer_summary_", scenario_string, ".csv"))
    write.csv(layer_summary, summary_path, row.names = FALSE)
    cat(sprintf("\nSaved summary table to: %s\n", summary_path))
  }
  
  return(list(
    layers = spatial_layers,
    summary = layer_summary,
    scenario = scenario_string
  ))
}
```

# 8. Create Port Risk Spatial Layers

 - I guess just leave SITC data in here for now. What else would we do with it? It's not really a port, but the risk is certainly present and coming through a port.
    - At least through this we are not removing this risk from the mapping

```{r create-port-risk-layers}
# Set output directory for port risk layers
port_risk_output_dir <- here::here("output", "spatial layers", "risk at ports")

# Create the spatial layers
spatial_layers <- create_spatial_layers(
  summarized_data = spatial_layers_summary,
  save_files = TRUE,
  output_dir = port_risk_output_dir
)
```





# create destination layers
- i dont yet know how to make destination layers because every destination I can find is empty.

10/9/25 - I need to create a destination layer for EANs. 
 - I will need to identify the appropriate destination points for these EANs and create a spatial layer that represents the risk associated with them.
    - I need to have their destination addresses. Once I do, I can use the some KDE process from before to create the destination layers
    - This will 