---
title: "EAN Product Type Analysis"
author: "ASF Risk Framework"
date: "`r Sys.Date()`"
output: html_document
params:
  ean_data_path: "output/intermediate files/ean_records.rds"
  huggingface_token: ""  # Optional - only needed for some models
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(reticulate)
library(jsonlite)

# Check if huggingfaceR is installed, if not install it
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
if (!requireNamespace("huggingfaceR", quietly = TRUE)) {
  remotes::install_github("farach/huggingfaceR")
}
library(huggingfaceR)
```

## Loading EAN Data

First, we'll load the EAN data that was processed in the previous script.

```{r load_data}
# Check if the file exists
ean_file_path <- here::here(params$ean_data_path)
if (!file.exists(ean_file_path)) {
  stop(paste("EAN data file not found at:", ean_file_path, 
             "Please run the '03_01 ean data.Rmd' script first."))
}

# Load the data
ean_records <- readRDS(ean_file_path)

# Print summary of the data
cat("Loaded", nrow(ean_records), "EAN records\n")
cat("Number of records with REMARKS:", sum(!is.na(ean_records$REMARKS)), "\n")

# Sample some remarks to see what we're working with
if (sum(!is.na(ean_records$REMARKS)) > 0) {
  set.seed(123)
  sample_remarks <- ean_records %>% 
    filter(!is.na(REMARKS)) %>% 
    sample_n(min(5, sum(!is.na(ean_records$REMARKS))))
  
  cat("\nSample REMARKS:\n")
  for (i in 1:nrow(sample_remarks)) {
    cat(i, ": ", sample_remarks$REMARKS[i], "\n\n")
  }
}
```

## Set Up HuggingFace Connection

We'll use a text-to-text model from HuggingFace that can process natural language and extract structured information.

```{r setup_huggingface}
# Set HuggingFace token if provided
if (params$huggingface_token != "") {
  Sys.setenv(HUGGINGFACEHUB_API_TOKEN = params$huggingface_token)
}

# Initialize the HuggingFace client for inference
# We'll use a model that doesn't require an API key for basic text processing
# FLAN-T5 is a good general-purpose model for text extraction tasks
model_id <- "google/flan-t5-base"

# Test if we can use the HuggingFace API without a token
# If this fails, we'll need to use a local transformer model
tryCatch({
  hf_client <- huggingfaceR::hf_client(model_id = model_id)
  test_prompt <- "Extract animal species: This remark mentions pork sausage."
  test_response <- hf_client$text_generation(test_prompt, max_length = 50)
  cat("HuggingFace API connection successful\n")
  use_api <- TRUE
}, error = function(e) {
  cat("Could not connect to HuggingFace API:", e$message, "\n")
  cat("Will use local transformers package instead\n")
  use_api <- FALSE
})
```

## Local Fallback Option Using Transformers Package

If the HuggingFace API isn't accessible, we'll set up a local alternative using the transformers package.

```{r setup_local_transformers, eval=!exists("use_api") || !use_api}
# Install and set up the transformers package via reticulate
py_install(packages = c("transformers", "torch"), pip = TRUE)

transformers <- reticulate::import("transformers")

# Load a local model for text extraction
# We'll use a smaller model for local processing
model_name <- "google/flan-t5-small"  # Smaller model for local use
tokenizer <- transformers$AutoTokenizer$from_pretrained(model_name)
model <- transformers$AutoModelForSeq2SeqLM$from_pretrained(model_name)

# Function to generate responses locally
local_generate <- function(prompt, max_length = 100) {
  input_ids <- tokenizer(prompt, return_tensors = "pt")
  outputs <- model$generate(
    input_ids$input_ids,
    max_length = as.integer(max_length),
    num_return_sequences = as.integer(1)
  )
  return(tokenizer$decode(outputs[[1]], skip_special_tokens = TRUE))
}

cat("Local transformers setup complete\n")
```

## Extraction Function

Now we'll create a function to extract the required information from the REMARKS field.

```{r extraction_function}
extract_product_info <- function(remark_text, use_api = TRUE) {
  if (is.na(remark_text) || remark_text == "") {
    return(list(
      animal_species = "none",
      product_type = "unknown",
      product_subtype = "unknown",
      quantity = NA,
      units = NA,
      remark = remark_text,
      json_response = "{}"
    ))
  }
  
  # Create a structured prompt to extract the information
  prompt <- paste0(
    "Extract the following information from this EAN REMARK:\n",
    "1. Animal species (must be one of: swine, other, none)\n",
    "2. Product type (must be one of: live animal, meat, fomite, other product, feed, unknown)\n",
    "3. Product subtype (e.g., processed-smoked, processed-cured, soil, boots, vaccine, semen)\n",
    "4. Quantity (numeric value only)\n",
    "5. Units (e.g., kg, g, lbs, pieces)\n\n",
    "REMARK: ", remark_text,
    "\n\nRespond in JSON format with these exact fields: animal_species, product_type, product_subtype, quantity, units"
  )
  
  # Generate response using either API or local model
  if (use_api && exists("hf_client")) {
    response_text <- hf_client$text_generation(prompt, max_length = 200)
  } else {
    response_text <- local_generate(prompt, max_length = 200)
  }
  
  # Try to parse the response as JSON
  json_result <- tryCatch({
    # Extract JSON from the response (it might include extra text)
    json_pattern <- "\\{.*\\}"
    json_text <- regmatches(response_text, regexpr(json_pattern, response_text))
    if (length(json_text) > 0) {
      fromJSON(json_text)
    } else {
      NULL
    }
  }, error = function(e) {
    NULL
  })
  
  # If JSON parsing failed, apply rule-based extraction as fallback
  if (is.null(json_result)) {
    # Rule-based fallback extraction
    remark_lower <- tolower(remark_text)
    
    # Animal species detection
    animal_species <- "none"
    if (grepl("swine|pork|pig|hog", remark_lower)) {
      animal_species <- "swine"
    } else if (grepl("animal|cattle|beef|chicken|poultry|sheep|goat|duck", remark_lower)) {
      animal_species <- "other"
    }
    
    # Product type detection
    product_type <- "unknown"
    if (grepl("live|living", remark_lower)) {
      product_type <- "live animal"
    } else if (grepl("meat|flesh|muscle|sausage|ham|salami", remark_lower)) {
      product_type <- "meat"
    } else if (grepl("soil|boot|shoe|clothes|clothing|dirt|manure", remark_lower)) {
      product_type <- "fomite"
    } else if (grepl("feed|fodder|silage", remark_lower)) {
      product_type <- "feed"
    } else if (grepl("vaccine|semen|embryo|blood|serum", remark_lower)) {
      product_type <- "other product"
    }
    
    # Product subtype detection
    product_subtype <- "unknown"
    if (product_type == "meat" && grepl("processed|smoked|dried|cured|cooked|fermented", remark_lower)) {
      subtype_patterns <- c("smoked", "dried", "cured", "cooked", "fermented")
      for (pattern in subtype_patterns) {
        if (grepl(pattern, remark_lower)) {
          product_subtype <- paste0("processed-", pattern)
          break
        }
      }
      if (product_subtype == "unknown") {
        product_subtype <- "processed-unknown"
      }
    } else if (product_type == "fomite") {
      fomite_patterns <- c("soil", "boot", "shoe", "clothes", "manure", "equipment")
      for (pattern in fomite_patterns) {
        if (grepl(pattern, remark_lower)) {
          product_subtype <- pattern
          break
        }
      }
    } else if (product_type == "other product") {
      other_patterns <- c("vaccine", "semen", "embryo", "blood", "serum")
      for (pattern in other_patterns) {
        if (grepl(pattern, remark_lower)) {
          product_subtype <- pattern
          break
        }
      }
    } else if (product_type == "live animal") {
      animal_patterns <- c("pig", "swine", "cattle", "chicken", "duck", "goat", "sheep")
      for (pattern in animal_patterns) {
        if (grepl(pattern, remark_lower)) {
          product_subtype <- pattern
          break
        }
      }
    }
    
    # Quantity and units detection
    quantity_pattern <- "([0-9]+\\.?[0-9]*)"
    units_pattern <- "(kg|g|gram|grams|lb|lbs|pound|pounds|oz|ounce|ounces|piece|pieces|unit|units)"
    
    quantity <- NA
    units <- NA
    
    quantity_match <- regexpr(quantity_pattern, remark_lower)
    if (quantity_match > 0) {
      quantity <- as.numeric(regmatches(remark_lower, quantity_match))
      
      # Look for units after the quantity
      remainder <- substr(remark_lower, quantity_match + attr(quantity_match, "match.length"), nchar(remark_lower))
      units_match <- regexpr(units_pattern, remainder)
      if (units_match > 0) {
        units <- regmatches(remainder, units_match)
      }
    }
    
    # Create a JSON-like structure for consistency
    json_result <- list(
      animal_species = animal_species,
      product_type = product_type,
      product_subtype = product_subtype,
      quantity = quantity,
      units = units
    )
  }
  
  # Ensure the result has all required fields with proper types
  if (!("animal_species" %in% names(json_result))) json_result$animal_species <- "none"
  if (!("product_type" %in% names(json_result))) json_result$product_type <- "unknown"
  if (!("product_subtype" %in% names(json_result))) json_result$product_subtype <- "unknown"
  if (!("quantity" %in% names(json_result)) || is.null(json_result$quantity)) json_result$quantity <- NA
  if (!("units" %in% names(json_result)) || is.null(json_result$units)) json_result$units <- NA
  
  # Standardize animal species
  if (!json_result$animal_species %in% c("swine", "other", "none")) {
    if (grepl("pig|pork|swine|hog", tolower(json_result$animal_species))) {
      json_result$animal_species <- "swine"
    } else if (json_result$animal_species != "") {
      json_result$animal_species <- "other"
    } else {
      json_result$animal_species <- "none"
    }
  }
  
  # Standardize product type
  valid_product_types <- c("live animal", "meat", "fomite", "other product", "feed", "unknown")
  if (!json_result$product_type %in% valid_product_types) {
    json_result$product_type <- "unknown"
  }
  
  # Add the original remark
  json_result$remark <- remark_text
  
  # Add the raw JSON response for debugging
  json_result$json_response <- toJSON(json_result, auto_unbox = TRUE)
  
  return(json_result)
}
```

## Process Sample Remarks

Let's test our extraction function on a sample of remarks to see how well it works.

```{r test_extraction}
# Only run this if we have remarks to process
if (sum(!is.na(ean_records$REMARKS)) > 0) {
  # Get a sample of remarks to test
  set.seed(123)
  test_remarks <- ean_records %>% 
    filter(!is.na(REMARKS)) %>% 
    sample_n(min(10, sum(!is.na(ean_records$REMARKS))))
  
  # Process each remark
  extracted_results <- list()
  for (i in 1:nrow(test_remarks)) {
    cat("Processing sample remark", i, "of", nrow(test_remarks), "\n")
    extracted_results[[i]] <- extract_product_info(
      test_remarks$REMARKS[i], 
      use_api = exists("use_api") && use_api
    )
  }
  
  # Convert results to a data frame
  results_df <- bind_rows(extracted_results)
  
  # Display the results
  knitr::kable(results_df %>% select(-json_response), 
               caption = "Sample Extraction Results")
}
```

## Process All Remarks

Now we'll process all the remarks in the dataset. This might take some time depending on the size of the dataset.

```{r process_all_remarks}
# Only process if we have remarks
if (sum(!is.na(ean_records$REMARKS)) > 0) {
  # Create a progress bar
  pb <- txtProgressBar(min = 0, max = sum(!is.na(ean_records$REMARKS)), style = 3)
  
  # Process each remark
  all_results <- list()
  counter <- 0
  
  for (i in 1:nrow(ean_records)) {
    if (!is.na(ean_records$REMARKS[i])) {
      counter <- counter + 1
      all_results[[i]] <- extract_product_info(
        ean_records$REMARKS[i], 
        use_api = exists("use_api") && use_api
      )
      setTxtProgressBar(pb, counter)
    } else {
      # For records with no remark, create a default entry
      all_results[[i]] <- list(
        animal_species = "none",
        product_type = "unknown",
        product_subtype = "unknown",
        quantity = NA,
        units = NA,
        remark = NA,
        json_response = "{}"
      )
    }
  }
  close(pb)
  
  # Convert to a data frame
  extracted_data <- bind_rows(all_results)
  
  # Add to the original data
  ean_records_with_extraction <- bind_cols(
    ean_records,
    extracted_data %>% select(-remark)
  )
  
  # Summary of extracted data
  cat("\nExtraction Summary:\n")
  
  cat("\nAnimal Species Distribution:\n")
  print(table(extracted_data$animal_species, useNA = "ifany"))
  
  cat("\nProduct Type Distribution:\n")
  print(table(extracted_data$product_type, useNA = "ifany"))
  
  cat("\nProduct Subtype Distribution (top 10):\n")
  subtype_counts <- sort(table(extracted_data$product_subtype, useNA = "ifany"), decreasing = TRUE)
  print(head(subtype_counts, 10))
  
  cat("\nQuantity Units Distribution:\n")
  print(table(extracted_data$units, useNA = "ifany"))
}
```

## Save the Results

Finally, we'll save the processed data for use in subsequent analyses.

```{r save_results}
# Save the enhanced data
if (exists("ean_records_with_extraction")) {
  output_path <- here::here("output", "intermediate files", "ean_records_with_product_types.rds")
  saveRDS(ean_records_with_extraction, output_path)
  cat("Results saved to:", output_path, "\n")
  
  # Also save a CSV for easy viewing
  csv_path <- here::here("output", "intermediate files", "ean_product_types_summary.csv")
  extracted_data %>%
    group_by(animal_species, product_type, product_subtype) %>%
    summarize(
      count = n(),
      avg_quantity = mean(quantity, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(count)) %>%
    write.csv(csv_path, row.names = FALSE)
  
  cat("Summary saved to:", csv_path, "\n")
}
```

## Visualization of Results

Let's create some visualizations to better understand the extracted data.

```{r visualize_results, fig.width=10, fig.height=8}
if (exists("extracted_data")) {
  # Animal species distribution
  ggplot(extracted_data %>% filter(!is.na(animal_species)), 
         aes(x = animal_species, fill = animal_species)) +
    geom_bar() +
    labs(title = "Distribution of Animal Species", 
         x = "Animal Species",
         y = "Count") +
    theme_minimal() +
    theme(legend.position = "none")
  
  # Product type distribution
  ggplot(extracted_data %>% filter(product_type != "unknown"), 
         aes(x = product_type, fill = product_type)) +
    geom_bar() +
    labs(title = "Distribution of Product Types", 
         x = "Product Type",
         y = "Count") +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Top product subtypes
  top_subtypes <- extracted_data %>%
    filter(product_subtype != "unknown") %>%
    count(product_subtype) %>%
    top_n(10, n) %>%
    arrange(desc(n))
  
  ggplot(extracted_data %>%
           filter(product_subtype %in% top_subtypes$product_subtype), 
         aes(x = reorder(product_subtype, table(product_subtype)[product_subtype]), 
             fill = product_type)) +
    geom_bar() +
    labs(title = "Top 10 Product Subtypes", 
         x = "Product Subtype",
         y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Quantity distribution (for non-NA quantities)
  if (sum(!is.na(extracted_data$quantity)) > 0) {
    ggplot(extracted_data %>% filter(!is.na(quantity)), 
           aes(x = quantity)) +
      geom_histogram(bins = 30, fill = "steelblue") +
      labs(title = "Distribution of Quantities", 
           x = "Quantity",
           y = "Count") +
      theme_minimal()
  }
  
  # Heatmap of animal species vs product type
  ggplot(extracted_data %>% 
           filter(animal_species != "none" & product_type != "unknown") %>%
           count(animal_species, product_type), 
         aes(x = product_type, y = animal_species, fill = n)) +
    geom_tile() +
    scale_fill_viridis_c() +
    labs(title = "Heatmap of Animal Species by Product Type", 
         x = "Product Type",
         y = "Animal Species",
         fill = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Conclusion

This analysis has processed the REMARKS field from EAN records to extract structured information about animal species, product types, processing methods, quantities, and units. The results have been saved for further analysis and integration with the ASF risk framework.

Next steps might include:
1. Refining the extraction model with more training data
2. Integrating this data with other risk factors
3. Creating more detailed visualizations by location, pathway, or other factors