---
title: "EAN data"
author: "Dalon White"
date: "2025-09-12"
output: 


params:
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  affected_product_file: "asf affected product name reference table.csv"
  affected_country_file: "asf affected countries.csv"
---

-   prepped data already has the pathway, entry type, etc., and risk as no/low/med/high
-   need to use country of origin and product origin to determine if the inspection has any asf risk
-   then model the kg of product with asf risk by pathway, entry type, location, country of origin, product pathway, product subpathway

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(lme4)
library(ggplot2)
library(broom)
library(multcomp)
library(emmeans)
library(sjPlot)
library(broom.mixed)
library(boot)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      affected_countries = "current", # "current" or "mex_can" or "other" 
      years = 2023:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"),
      affected_product_file = "asf affected product name reference table.csv",
      affected_country_file = "asf affected countries.csv",
    )
  }
}
```


```{r ARM AQIM data}
ean_cols <- c(
  "ID",
  "COMMODITY_ID",
   "CONVEYANCE_ID",
  "ISSUED_DATETIME",
  "ISSUED_DATETIME_FISCAL_YEAR",
  "ISSUED_DATETIME_FISCAL_MONTH",
  "CATEGORY",
  "SUBCATEGORY",
  "PATHWAY_ID", #PATHWAY ID can distinguish commerce and express carrier; probably need a reference table because the naming conventions aren't consistent
  "PATHWAY",
   "REMARKS",
   "DESTINATION_OF_ARTICLES_ID",
     "REGULATORY_ACTION_TYPE"
)

conveyance_cols <- c(
  "ID",
  "INSPECTION_ID",
  "CONVEYANCE_TYPE",
  "AIRCRAFT_TYPE",
  "VEHICLE_TYPE" #if !NULL, then it's a vehicle. Need to explore if it includes cargo trucks
)

inspection_cols <- c(
  "ID",

  "INSPECTION_NUMBER", 
  "IS_AQIM", 
  "PORT_OF_ENTRY_ID",
  
  "INSPECTION_DATETIME", 
  "INSPECTION_DATETIME_FISCAL_YEAR", 
  "INSPECTION_DATETIME_FISCAL_MONTH",
  "INSPECTION_LOCATION_ID", 
  "INSPECTION_LOCATION_NAME", 
  "INSPECTION_LOCATION_STATE_CODE", 
  "COUNTRY_OF_ORIGIN_NAME", 
 # "COUNTRY_OF_RESIDENCE_NAME", 
  
 # "FINAL_DESTINATION_STATE_NAME", 
 # "FINAL_DESTINATION_STATE_ID", 
 # "FINAL_DESTINATION_CITY", 
 # "PASSENGER_DISTANCE_DESTINATION_NAME",
  "NUMBER_OF_PASSENGERS")


commodity_cols <- c(
  "ID",
  "INSPECTION_ID",
  "IS_SMUGGLED",
  "IS_CONTAMINANT_FOUND",
  "COMMODITY_CLASSIFICATION",
  "COUNTRY_OF_ORIGIN_NAME", 
  "REF_COMMODITY_ID",
  "COMMODITY_DISPLAY_NAME",
  "MISC_ANIMAL_COMMODITY_NAME",
  "QUANTITY",
  "QUANTITY_UNITS_NAME"
)

#columns from the brg_pathway table that have more colloquial explanations to PATHWAY
pathway_cols <- c(
  "PATHWAY",
  "DISPLAY_NAME",
  "PORT_TYPE",
  "INSPECTION_TARGET",
  "MODE_OF_TRANSPORT",
  "DEFAULT_EXCLUSION_OPTIONS"
)
#-- Pull AQIM inspection records from ARM  ---- 
ean <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_EAN]")) |> 
              dplyr::select(all_of(ean_cols)) |> 
              #rename to fit AQAS data structure
              rename(EAN_ID = ID)

inspections <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_FACT_INSPECTION]")) |> 
              dplyr::filter(CATEGORY != "AQIM") |>
              dplyr::select(all_of(inspection_cols)) |>
              rename(INSPECTION_ID = ID,
                    #rename to match AQAS data structure
                     product_origin = COUNTRY_OF_ORIGIN_NAME)
 
#-- Load Bridge Commodity dataset ----
commodity <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |> 
              dplyr::select(all_of(commodity_cols)) |>
              rename(COMMODITY_ID = ID)

conveyance <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_CONVEYANCE]")) |>
              dplyr::select(all_of(conveyance_cols)) |>
              rename(CONVEYANCE_ID = ID)

#-- Load Bridge Pathway dataset ----
pathway_ref <- tbl(db_conn, sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_PATHWAY]")) |> 
              dplyr::select(all_of(pathway_cols))

# Join ean records with commodity table info
## Identify columns to join by
intersect_cols <- intersect(colnames(ean), colnames(commodity))
ean_records_tmp <- ean |>
  left_join(commodity, by = intersect_cols, relationship = "many-to-many")

# Join ean and commodity with inspection details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records_tmp), colnames(inspections))
ean_records_tmp <-ean_records_tmp |>
  left_join(inspections, by = intersect_cols, relationship = "many-to-many") 

# Join ean, commodity, and inspection details with conveyance details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records_tmp), colnames(conveyance))
ean_records <- ean_records_tmp |>
  left_join(conveyance, by = intersect_cols, relationship = "many-to-many")

# Join ean, commodity, inspection, and conveyance details with pathway details
## identify columns to join by
intersect_cols <- intersect(colnames(ean_records), colnames(pathway_ref))
ean_records <- ean_records |>
  left_join(pathway_ref, by = intersect_cols, relationship = "many-to-many")

#using intersect columns above will ensure that multiple overlapping columns won't create a .x and .y version of the same column

```

# Collect data
I have compared filtering in the database vs. collecting and filtering in R. Collecting is only a couple of seconds, so it's fast enough and filtering is much easier for some operations that require complex manipulations or functions not supported by the SQL. The filtering by datetime is particularly easier when done in R due to the need for lubridate functions and having to otherwise manipulate the datetime fiscal year & month columns separately.

```{r}

#open connection
db_conn <- dbConnect(odbc::odbc(),.connection_string = 
                      "Driver=SQL Server;
                        Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
                        trusted_connection=yes")

ean_records <- ean_records |> collect()

# Disconnect from the database
dbDisconnect(db_conn)
```


# filter to relevant records
## filter by years

### Use the ISSUED_DATETIME or the INSPECTION_DATETIME ?
```{r}
datetime_comparison <- ean_records %>%
  summarize(
    total_records = n(),
    issued_na_count = sum(is.na(ISSUED_DATETIME)),
    inspection_na_count = sum(is.na(INSPECTION_DATETIME)),
    both_na_count = sum(is.na(ISSUED_DATETIME) & is.na(INSPECTION_DATETIME)),
    only_issued_na = sum(is.na(ISSUED_DATETIME) & !is.na(INSPECTION_DATETIME)),
    only_inspection_na = sum(!is.na(ISSUED_DATETIME) & is.na(INSPECTION_DATETIME)),
    both_present = sum(!is.na(ISSUED_DATETIME) & !is.na(INSPECTION_DATETIME))
  ) %>%
  mutate(
    issued_na_percent = round(issued_na_count / total_records * 100, 2),
    inspection_na_percent = round(inspection_na_count / total_records * 100, 2),
    both_na_percent = round(both_na_count / total_records * 100, 2),
    both_present_percent = round(both_present / total_records * 100, 2)
  )

# Create a visual representation
na_comparison <- data.frame(
  datetime_field = c("ISSUED_DATETIME", "INSPECTION_DATETIME", "Both NA", "Both Present"),
  na_count = c(datetime_comparison$issued_na_count, 
               datetime_comparison$inspection_na_count,
               datetime_comparison$both_na_count,
               datetime_comparison$both_present),
  percentage = c(datetime_comparison$issued_na_percent,
                 datetime_comparison$inspection_na_percent,
                 datetime_comparison$both_na_percent,
                 datetime_comparison$both_present_percent)
)

# Print the detailed summary
print(na_comparison)


# Calculate time difference between ISSUED_DATETIME and INSPECTION_DATETIME
datetime_diff <- head(ean_records, 10000) %>%
  # Filter to records where both datetime fields are present
  filter(!is.na(ISSUED_DATETIME) & !is.na(INSPECTION_DATETIME)) %>%
  # Calculate the difference in hours
  mutate(
    time_diff_hours = as.numeric(difftime(ISSUED_DATETIME, INSPECTION_DATETIME, units = "hours")),
    time_diff_days = as.numeric(difftime(ISSUED_DATETIME, INSPECTION_DATETIME, units = "days")),
    # Create categories for analysis
    time_diff_category = case_when(
      time_diff_hours < 0 ~ "Issued before inspection (error?)",
      time_diff_hours == 0 ~ "Same time",
      time_diff_hours > 0 & time_diff_hours <= 1 ~ "Within 1 hour",
      time_diff_hours > 1 & time_diff_hours <= 24 ~ "1-24 hours",
      time_diff_hours > 24 & time_diff_hours <= 168 ~ "1-7 days",
      time_diff_hours > 168 ~ "Over 7 days"
    )
  )

  # Distribution by category
time_diff_distribution <- datetime_diff %>%
  group_by(time_diff_category) %>%
  summarize(
    count = n(),
    percentage = n() / nrow(datetime_diff) * 100
  ) %>%
  arrange(factor(time_diff_category, levels = c(
    "Issued before inspection (error?)", 
    "Same time", 
    "Within 1 hour", 
    "1-24 hours", 
    "1-7 days", 
    "Over 7 days"
  )))


# Summary statistics
time_diff_summary <- datetime_diff %>%
  summarize(
    n_records = n(),
    min_diff_hours = min(time_diff_hours, na.rm = TRUE),
    max_diff_hours = max(time_diff_hours, na.rm = TRUE),
    median_diff_hours = median(time_diff_hours, na.rm = TRUE),
    mean_diff_hours = mean(time_diff_hours, na.rm = TRUE),
    sd_diff_hours = sd(time_diff_hours, na.rm = TRUE),
    q25_diff_hours = quantile(time_diff_hours, 0.25, na.rm = TRUE),
    q75_diff_hours = quantile(time_diff_hours, 0.75, na.rm = TRUE)
  )


print("Time difference summary statistics (in hours):")
print(time_diff_summary)

print("Distribution by time difference category:")
print(time_diff_distribution)
```

In most cases, it looks like inspection datetime and issued datetime are pretty much the same, with 5% delayed issuance dates >7 days
So, results suggest filter by INSPECTION_DATETIME first then by ISSUED_DATETIME if needed

Filter using the params$years:
  Will call the years parameter here. The years parameter is used for the WADS codes when calculating the volume for AQIM-based pathways. By calling it here, this will ensure that we are comparing predicted volumes through AQIM+WADS for a given time to the volume of EAN-related products for the same time frame.

### Filter by INSPECTION_DATETIME then ISSUED_DATETIME
```{r}
# Filter to records within specified years
filter_years <- function(data) {
  # First ensure we have at least one datetime field
  filtered_data <- data %>% 
    filter(!is.na(INSPECTION_DATETIME) | !is.na(ISSUED_DATETIME))
  
  # Extract years from datetime fields
  filtered_data <- filtered_data %>%
    mutate(
      inspection_year = year(INSPECTION_DATETIME),
      issued_year = year(ISSUED_DATETIME)
    )
  
  # Filter by year using logical OR approach
  filtered_data <- filtered_data %>%
    filter(
      # Keep if INSPECTION_DATETIME year is in range
      (!is.na(inspection_year) & inspection_year %in% params$years) |
      # OR if INSPECTION_DATETIME is NA but ISSUED_DATETIME year is in range
      (is.na(inspection_year) & !is.na(issued_year) & issued_year %in% params$years)
    )
  
  # Remove the temporary year columns
  filtered_data <- filtered_data %>%
    dplyr::select(-inspection_year, -issued_year)
  
  return(filtered_data)
}

# Apply the function
ean_records <- filter_years(ean_records)
```

## separate the data into layers

This is to separate the data into layers that can be used to model risk. The layers are provided i nthe entry type reference table. This converts the PATHWAY to specific modes of entry
 - passenger
 - eco
 - mail
 - commercial
 - unknown
"unknown" need to be defined. There's 'Ambiguous', and some non-port/non-transport that end in EPP, DPP, and EC. Not sure if any are relevant to this analysis. For now they are coded just as unknown in the reference file.

Load pathway to entry type reference table:

```{r load pathway to entry type reference table}
setwd(here::here('input', 'data', 'PATHWAY name reference'))
# Load the pathway reference table
pathway_ref <- read.csv("pathway to entry type reference table.csv", stringsAsFactors = FALSE)
ean_records <- ean_records %>%
  left_join(pathway_ref, by = "PATHWAY", relationship = "many-to-one")

get_entry_type <- function(pathway_str) {
  for (i in 1:nrow(pathway_ref)) {
    ref_pathway <- pathway_ref$PATHWAY[i]
    if (grepl(ref_pathway, pathway_str, fixed = TRUE)) {
      return(pathway_ref$entry_type[i])
    }
  }
  return(NA_character_)
}

# Apply the function to each pathway in ean_records
ean_records <- ean_records %>%
  rowwise() %>%
  mutate(entry_type = get_entry_type(PATHWAY)) %>%
  ungroup()
```

### use brg_pathway to separate

### Conveyance table can separate out private and commercial aircraft
```{r}
ean_records <- ean_records |> mutate(MODE_OF_TRANSPORT = 
case_when(
  CONVEYANCE_TYPE == "AIRCRAFT" & !is.na(AIRCRAFT_TYPE) ~ paste(MODE_OF_TRANSPORT, AIRCRAFT_TYPE, sep = " - "),
  .default = MODE_OF_TRANSPORT
  ))
```


# Determine if there is risk

Next step is to determine if there is any risk associated with the EAN record. This is through both the country of origin and the commodity type.

## Evaluate country

load country reference list
```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
if(params$affected_countries == "current") {
  affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
} else if (params$affected_countries == "mex_can") {
  affected_countries <- c("MEX", "CAN")
} else if (params$affected_countries == "other") {
  affected_countries <- params$other_affected_countries
} else {
  stop("Invalid value for params$affected_countries")
}
```

filter to  eans that are from that country
```{r isolate records from risky countries}
ean_records <- ean_records |> mutate(country_code = countrycode::countrycode(COUNTRY_OF_ORIGIN_NAME, 
                                  origin = "country.name", 
                                  destination = "iso3c")
                                  )

ean_records <- ean_records |> filter(country_code %in% affected_countries)
```


## Evaluate product
The type of product will be stored in COMMODITY_DISPLAY_NAME, just like in AQIM data.

```{r what kind of products descriptions for eans}
 View(ean_records |> filter(grepl("swine|porcine|pork|hooves|meat|ASF|CSF|blood|manure|dirt|feed",tolower(COMMODITY_DISPLAY_NAME))) |> count(REMARKS))

print("most dont have a descriptive COMMODITY_DISPLAY_NAME like AQIM data")
```

```{r save for remarks analysis}
setwd(here::here("output","intermediate files"))
saveRDS(ean_records, "ean_records.rds")

```



load product reference list
```{r identify products of risk}
setwd(here::here("input","data","asf affected product names"))
asf_product_ref <- read.csv(params$affected_product_file)
match_asf_risk <- function(data, reference_table) {
  # Try matching on COMMODITY_DISPLAY_NAME
  asf_product_match <- data %>%
    inner_join(reference_table, 
               by = c("COMMODITY_DISPLAY_NAME" = "ITEM", "entry_type" = "entry_type"), 
               relationship = "many-to-many",
               copy = TRUE)
  
  # Combine matches and remove duplicates
  asf_matches <- asf_product_match %>%
    distinct(ID, COMMODITY_ID, .keep_all = TRUE)
  
  # Return the matches for further processing
  return(asf_matches)
}

# Apply to eans
at_risk_eans <- match_asf_risk(ean_records, asf_product_ref)
```
filter to eans that are that product
```{r isolate records from risky products}
ean_records <- ean_records |> mutate(commodity_name = toupper(COMMODITY_DISPLAY_NAME))
ean_records <- ean_records |> filter(commodity_name %in% affected_products)
```




# separate the data into layers

## pathway can separate the conveyances



## Conveyance table can separate out private and commercial aircraft
```{r}

```






# Get address

## what are the address columns

## Where is destination address stored?

#Noah Stuart will get back to me on the addresses for EANs (Teams chat, 9/15/25)
#Until then, I can still pull data, filter by commodity, etc. Getting a location would be last step, anyways.
