---
title: "IBL geolocations and layers for port and destination"
author: "Dalon White"
date: "2025-09-30"
output: 


params:
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  ibl_geolocated_filename: "ibl_records_geocoded"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(here)
library(odbc)
library(DBI)
```

# Parameters

```{r parameters, include=FALSE}
# Load parameters
source(file.path(here::here(), "scripts", "initialize_params.R"))
```

# Access lat lon data for the correct years

```{r load-ibl-data}
ibl_data_filtered <- readRDS(here::here("output", "intermediate files", params$ibl_geolocated_filename))

## convert to spatial with the same CRS
ibl_data_sf <- ibl_data_filtered |>
  filter(!is.na(lon) & !is.na(lat)) |> 
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  # Transform to NAD83/Albers (EPSG:5070) for analysis
  sf::st_transform(crs = 5070)

# Save spatial object for future use
saveRDS(ibl_data_sf, here::here("output", "intermediate files", paste0(params$ibl_geolocated_filename,"_sf")))

```

## 1.2 Filter to US Boundaries

```{r us-boundaries}
# Load US boundaries 
us_bounds_sf <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_bounds.rds"))
us_grid <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_grid.rds"))

# Filter spatial data to US boundaries
ibl_data_sf_filtered <- ibl_data_sf %>%
  st_join(us_bounds_sf, left = TRUE) 

# Report filtering statistics
total_points <- nrow(ibl_data_sf)
kept_points <- nrow(ibl_data_sf_filtered)
removed_points <- total_points - kept_points

cat(sprintf("Filtered to %d points within US boundaries (%d removed)\n", 
          kept_points, removed_points))
```

# Next Steps

9/30/25
Need to convert the IBLs to closest port locatoin
 - this would require import PoE locations and assigning closest to locations
    - then summarize by PoE, pathway, container, product_pathway, product_subpathway

```{r summarize-for-spatial-layers}

geo_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=GeoLocations;
                      trusted_connection=yes")


cbp_poe_cols <- c("CBP_NAME", "CBP_CODE", "LAT", "LON", "Port of Entry Name")


# Get CBP Port of Entry data with coordinates
cbp_poe <- tbl(geo_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_CBP_POEs]")) %>% 
           dplyr::select(all_of(cbp_poe_cols)) %>% 
           collect()


```

```{r assign closest POE by lat lon}
# Convert CBP POE to sf object
cbp_poe_sf <- cbp_poe  %>%
  filter(!is.na(LON) & !is.na(LAT)) %>%
  sf::st_as_sf(coords = c("LON", "LAT"), crs = 4326) %>%
  sf::st_transform(crs = 5070)
# Method 1: Using st_nearest_feature with direct indexing (cleaner approach)
# Find nearest POE for each IBL point and join
ibl_data_with_poe <- sf::st_join(
  ibl_data_sf_filtered,
  cbp_poe_sf %>% dplyr::select(CBP_NAME, CBP_CODE, `Port of Entry Name`),
  join = sf::st_nearest_feature
)
```


# 1.3 Summarize Data for Spatial Layers

```{r summarize-for-spatial-layers}
# Group by location and pathway combinations and summarize
spatial_layers_summary <- ibl_data_with_poe %>% rename(LOCATION_ID = CBP_CODE,
LOCATION_NAME = CBP_NAME) |> 
  mutate(LOCATION_STATE_CODE = (
    #get state code from any letters after ',' in LOCATION_NAME
    str_extract(`Port of Entry Name`, "(?<=, )\\w{2}$")
  )) |>
  mutate(LOCATION_NAME = paste0(LOCATION_NAME, " (", LOCATION_ID, ")")) |>
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, 
           container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(Animal.kg, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Filter out incomplete records
  filter(!is.na(container) & !is.na(product_pathway) & !is.na(product_subpathway))
```



# 7. Create and Save Spatial Layers Function

```{r create-spatial-layers-function}
# Function to create and save individual spatial layers for each pathway combination
create_spatial_layers <- function(summarized_data, save_files = FALSE, output_dir = NULL) {
  # Split data by pathway combinations
  split_data <- summarized_data %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    group_split()
  
  cat(sprintf("Created %d spatial layers for different pathway combinations\n", length(split_data)))
  
  # Create list to store the layers
  spatial_layers <- list()

  # Create scenario string that specifies IBLs
  scenario_string <- "IBL"
  
  # Process each split
  for (i in seq_along(split_data)) {
    current_split <- split_data[[i]]
    
    # Get combination values for naming
    combo <- current_split %>%
      dplyr::select(PATHWAY, container, product_pathway, product_subpathway) %>%
      slice(1) %>%
      as.list()
    
    # Create clean name for layer using abbreviations
    layer_name <- paste(
      substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 27),
      substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3),
      sep = "_"
    )
    
    # Store in list with clean name
    spatial_layers[[layer_name]] <- current_split
    
    # Save to file if requested
    if (save_files && !is.null(output_dir)) {
      # Ensure output directory exists
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Convert to POINT geometry (required for GeoPackage)
      export_layer <- current_split %>%
        sf::st_cast("POINT")
      
      # Include scenario in filename - use GeoPackage format for Shiny compatibility
      file_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".gpkg"))
      
      # Try to save the file
      tryCatch({
        sf::st_write(export_layer, file_path, delete_layer = TRUE)
        cat(sprintf("Saved layer %d/%d: %s\n", i, length(split_data), file_path))
      }, error = function(e) {
        cat(sprintf("ERROR saving layer %d/%d: %s - %s\n", 
                   i, length(split_data), layer_name, e$message))
      })
    }
  }
    # Calculate totals directly from the original summarized data to ensure consistency
  # This approach guarantees the same results as the manual grouping and summarization
  layer_summary <- summarized_data %>%
    sf::st_drop_geometry() %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    summarize(
      total_risk = sum(total_mean_kg, na.rm = TRUE),
      point_count = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      layer_name = paste(
        substr(gsub("[^[:alnum:]]", "", PATHWAY), 1, 27),
        substr(gsub("[^[:alnum:]]", "", container), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_pathway), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_subpathway), 1, 3),
        sep = "_"
      ),
      scenario = scenario_string
    ) %>%
    dplyr::select(layer_name, scenario, point_count, total_risk) %>%
    arrange(desc(total_risk))
  
  # Print summary
  cat(paste0("\nSummary of spatial layers for scenario '", scenario_string, "' (top 10 by risk):\n"))
  print(head(layer_summary, 10))
  
  # Also save the summary table as CSV for reference
  if (save_files && !is.null(output_dir)) {
    summary_path <- file.path(output_dir, paste0("layer_summary_", scenario_string, ".csv"))
    write.csv(layer_summary, summary_path, row.names = FALSE)
    cat(sprintf("\nSaved summary table to: %s\n", summary_path))
  }
  
  return(list(
    layers = spatial_layers,
    summary = layer_summary,
    scenario = scenario_string
  ))
}
```

# 8 Create and Save Spatial Layers
```{r create-port-risk-layers}
# Set output directory for port risk layers
port_risk_output_dir <- here::here("output", "spatial layers", "risk at ports")

# Create the spatial layers
spatial_layers <- create_spatial_layers(
  summarized_data = spatial_layers_summary,
  save_files = TRUE,
  output_dir = port_risk_output_dir
)
```




# currently spit an error until destination risk layers are corrected

```{r error until destination risk layers completed, eval=FALSE}
stop("Destination risk layers creation not yet implemented") 
```





# 9. Prepare Data for Final Destination Risk Layers

```{r prepare-final-destination-data}
# Summarize by pathway combination (dropping geometry)
pathway_spatial_layers_summary <- spatial_layers_summary %>% 
  sf::st_drop_geometry() %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    #total_mean_kg is not actually the mean; in the other pathways, the 'mean' is the mean simulations but in this case there is only 1 simulation - reality. So, keeping the name for naming consistency, but leaving a note in case of confusion
    total_mean_kg = sum(total_mean_kg, na.rm = TRUE),
    .groups = 'drop'
  )

setwd(here::here('output','report files'))
write.csv(pathway_spatial_layers_summary, paste("IBL pathway summary", min(params$years), "to", max(params$years),".csv", sep = "_"), row.names = FALSE)
```

# 10. Create Final Destination Risk Layers Function

```{r distribute-pathway-risks-function}
# Function to distribute pathway risks to final destination spatial grids
distribute_pathway_risks <- function(pathway_summary, kde_folder_path) {
  # List all available KDE files (kernel density estimation grids)
  kde_files <- list.files(path = kde_folder_path, pattern = "kde_scaled_.*\\.rds", full.names = TRUE)
  
  # Create output directory for final destination risk layers
  output_dir <- here::here('output', 'spatial layers', 'risk at final destinations')
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Initialize data structures
  distributed_layers <- list()
  mapping_df <- data.frame(
    short_name = character(),
    full_name = character(),
    pathway = character(),
    container = character(),
    product_pathway = character(),
    product_subpathway = character(),
    stringsAsFactors = FALSE
  )
  
  # Process each unique pathway
  for (pathway in unique(pathway_summary$PATHWAY)) {
    # Find the matching KDE file for this pathway
    kde_filename_pattern <- paste0("kde_scaled_", pathway, ".rds")
    matching_files <- grep(kde_filename_pattern, kde_files, value = TRUE, fixed = TRUE)
    
    if (length(matching_files) == 0) {
      cat(sprintf("No KDE file found for pathway: %s\n", pathway))
      next
    }
    
    # Load the KDE grid for this pathway
    kde_grid <- readRDS(matching_files[1])
    
    # Get all combinations for this pathway
    pathway_combos <- pathway_summary %>% filter(PATHWAY == pathway)
    
    # Process each combination
    for (i in seq_len(nrow(pathway_combos))) {
      combo <- pathway_combos[i,]
      risk_value <- combo$total_mean_kg
      
      # Create abbreviated name components for file naming
      pathway_abbr <- substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 27)
      container_abbr <- substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3)
      prod_path_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3)
      prod_subpath_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3)
      
      # Create short name for file (with unique index)
      short_combo_name <- paste(pathway_abbr, container_abbr, prod_path_abbr, prod_subpath_abbr, i, sep = "_")
      
      # Create full descriptive name for reference
      full_combo_name <- paste(
        gsub("[^[:alnum:]]", "_", combo$PATHWAY),
        gsub("[^[:alnum:]]", "_", combo$container),
        gsub("[^[:alnum:]]", "_", combo$product_pathway),
        gsub("[^[:alnum:]]", "_", combo$product_subpathway),
        sep = "_"
      )
      
      # Apply the risk value to the KDE grid to create distributed risk
      distributed_kde <- kde_grid %>%
        dplyr::mutate(risk_value = kde_value * risk_value)
      
      # Store in layers list
      distributed_layers[[full_combo_name]] <- distributed_kde
      
      # Add mapping entry
      new_mapping <- data.frame(
        short_name = short_combo_name,
        full_name = full_combo_name,
        pathway = combo$PATHWAY,
        container = combo$container,
        product_pathway = combo$product_pathway,
        product_subpathway = combo$product_subpathway,
        stringsAsFactors = FALSE
      )
      mapping_df <- rbind(mapping_df, new_mapping)
      
      # Save as GeoPackage for Shiny compatibility
      output_file <- file.path(output_dir, paste0("risk_", short_combo_name, ".gpkg"))
      
      tryCatch({
        sf::st_write(distributed_kde, output_file, delete_layer = TRUE)
        cat(sprintf("Processed %s → %s with risk %f\n", full_combo_name, short_combo_name, risk_value))
      }, error = function(e) {
        cat(sprintf("ERROR: Could not write %s - %s\n", short_combo_name, e$message))
      })
    }
  }
  
  # Save the name mapping file for reference
  write.csv(mapping_df, file.path(output_dir, "filename_mapping.csv"), row.names = FALSE)
  
  # Create a layer catalog with additional metadata
  layer_catalog <- mapping_df %>%
    mutate(
      file_path = file.path(output_dir, paste0("risk_", short_name, ".gpkg")),
      risk_value = sapply(distributed_layers[full_name], function(x) sum(x$risk_value, na.rm=TRUE)),
      cell_count = sapply(distributed_layers[full_name], nrow)
    ) %>%
    arrange(desc(risk_value))
  
  # Save the catalog
  write.csv(layer_catalog, file.path(output_dir, "layer_catalog.csv"), row.names = FALSE)
  
  return(list(
    layers = distributed_layers,
    mapping = mapping_df,
    catalog = layer_catalog
  ))
}
```

# 11. Create Final Destination Risk Layers

```{r create-final-destination-layers}
# Apply the function to distribute risks to final destinations
distributed_risks <- distribute_pathway_risks(
  pathway_summary = pathway_spatial_layers_summary,
  kde_folder_path = here::here("output", "proportional destination layers")
)

# Save a compact version of the catalog for Shiny app use
risk_summary <- distributed_risks$catalog %>%
  dplyr::select(
    layer_name = full_name, 
    short_name,
    pathway,
    container, 
    product_pathway,
    product_subpathway,
    total_risk = risk_value, 
    cell_count
  )

# Save the summary for reference
write.csv(risk_summary, 
          here::here("output", "aqim total distributed risk", paste0("distributed_risk_summary_", params$affected_countries,".csv")), 
          row.names = FALSE)

cat("Spatial risk layers created and ready for use in Shiny app\n")
```
