---
title: "IBL data prep"
author: "Dalon White"
date: "2025-09-30"
output: 


params:
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  ibl_geolocated_filename: "ibl_records_geocoded"
---

```{r setup, include=FALSE}
input_dir <- file.path(here::here(),"input", "data", "illegal boat landings")
output_dir <- file.path(here::here(),"output", "IBLs")
layer_dir <- file.path(here::here(),"output", "spatial_layers")
```

```{r parameters, include=FALSE}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      begin_date = "2000-01-01",
      end_date = "2024-12-31",
      US_grid_cell_size = 30000,
      ibl_geolocated_filename = "ibl_records_geocoded"
    )
  }
}
```

```{r load data}
library(readr)
library(dplyr)
library(stringr)
library(readxl)  # Added for reading Excel files

# Read Excel files instead of CSV
locs <- read_excel(file.path(input_dir, "IBL 8-13-2025 2-32-25 PM.xlsx"))
dat <- read_excel(file.path(input_dir, "All IBL Tasks 8-13-2025 2-33-08 PM.xlsx"))


```
```{r clean data}
#---- Clean and select relevant columns ----
locs <- locs |> dplyr::select(`Structure name`, `Structure Entry Latitude`, `Structure Entry Longitude`, `Access Group`) |>
rename(structure_name = `Structure name`, lat = `Structure Entry Latitude`, lon = `Structure Entry Longitude`, access_group = `Access Group`)


dat <- dat |> 
dplyr::select(`TK: ID#`,Owner, `Sub-Category`, `Start Date`, `Completion comments`) |>
rename(task = `TK: ID#`, owner = Owner, sub_category = `Sub-Category`, start_date = `Start Date`, completion_comments = `Completion comments`)
```

```{r extract kgs}
#---- Create animal and plant kg data columns---- 
dat$completion_comments <- str_replace_all(dat$completion_comments, "(?i)anima\\b", "animal")  # Correct 'anima' to 'animal'
dat$completion_comments <- str_replace_all(dat$completion_comments, "(?i)plan\\b", "plant")   # Correct 'plan' to 'plant'
dat$completion_comments <- str_replace_all(dat$completion_comments, "(?<=\\s|^)\\.(\\d+)", "0.\\1")  # Replace .5 with 0.5

#--Regex patterns to extract numbers
animal_pattern <- "(?i)animal(?:s|\\s*products?)?\\s*[:#;\\-]?\\s*(\\d+(\\.\\d+)?)\\s*(pounds|lbs|#)?"
plant_pattern <- "(?i)plants?(?:\\s*products?)?\\s*[:#;\\-]?\\s*(\\d+(\\.\\d+)?)\\s*(pounds|lbs|#)?"
 
#--Additional regex patterns to capture numbers after "#:" or "#: " separators
animal_separator_pattern <- "(?i)animal\\s*#:\\s*(\\d+(\\.\\d+)?)"
plant_separator_pattern <- "(?i)plant[s]?\\s*#:\\s*(\\d+(\\.\\d+)?)"
 
#--First, extract using the main patterns
dat$Animal.kg <- as.numeric(str_extract(dat$completion_comments, animal_pattern) %>%
                              str_extract("\\d+(\\.\\d+)?"))
 
dat$Plant.kg <- as.numeric(str_extract(dat$completion_comments, plant_pattern) %>%
                             str_extract("\\d+(\\.\\d+)?"))
 
#--Now, extract using the separator patterns
dat$Animal.kg[is.na(dat$Animal.kg)] <- as.numeric(str_extract(dat$completion_comments[is.na(dat$Animal.kg)], animal_separator_pattern) %>%
                                                    str_extract("\\d+(\\.\\d+)?"))

dat$Plant.kg[is.na(dat$Plant.kg)] <- as.numeric(str_extract(dat$completion_comments[is.na(dat$Plant.kg)], plant_separator_pattern) %>%
                                                  str_extract("\\d+(\\.\\d+)?"))
 
#--Convert pounds to kilograms
dat$Animal.kg <- dat$Animal.kg * 0.453592
dat$Plant.kg <- dat$Plant.kg * 0.453592
```

```{r merge data}
# join
ibl_data <- dat %>%
  # First, do a left join based on structure name
  left_join(locs, by = c("sub_category" = "structure_name", "owner" = "access_group"))

ibl_data |> count(sub_category, owner) |> filter(n >1)

#View(ibl_data |> count(sub_category, owner) |> filter(n >1)) 
# There are repeated incidents, but not many. I'm not seeing any items on them. I think they are perhaps multiple visits to the same site, or a broken vessel arriving, perhaps.
```

# assign PATHWAY name
```{r assign pathway name}
# Assign PATHWAY name based on sub-category
ibl_data <- ibl_data %>%
  mutate(PATHWAY = "IBL"
  )
```
Load pathway to entry type reference table:

```{r load pathway to entry type reference table}
setwd(here::here('input', 'data', 'PATHWAY name reference'))
# Load the pathway reference table
pathway_ref <- read.csv("pathway to entry type reference table.csv", stringsAsFactors = FALSE)
ibl_data <- ibl_data %>%
  left_join(pathway_ref, by = "PATHWAY", relationship = "many-to-one")

get_entry_type <- function(pathway_str) {
  for (i in 1:nrow(pathway_ref)) {
    ref_pathway <- pathway_ref$PATHWAY[i]
    if (grepl(ref_pathway, pathway_str, fixed = TRUE)) {
      return(pathway_ref$entry_type[i])
    }
  }
  return(NA_character_)
}

# Apply the function to each pathway in ibl_data
ibl_data <- ibl_data %>%
  rowwise() %>%
  mutate(entry_type = get_entry_type(PATHWAY)) %>%
  ungroup()
```


# Match with the product reference list, which incldues a special category I created(ITEM == 'Animal_IBL') which has 'HIGH' ASF risk

## IBL product name classification
IBLs don't have a classfication other than animal product in #'s, converted to kgs. FOr other pathways, this is pulled straight from the Commodity Brg table. Here, we will assign it directly.

- these steps just help it fit in the pipeline in case something changes later - it's a long way of doing it
```{r assign product pathway}
ibl_data <- ibl_data %>%
  mutate(COMMODITY_DISPLAY_NAME = "Animal_IBL")
```

load product reference list
```{r identify products of risk}
setwd(here::here("input","data","asf affected product names"))
asf_product_ref <- read.csv(params$affected_product_file)

match_asf_risk <- function(data, reference_table) {
  # Try matching on COMMODITY_DISPLAY_NAME
  asf_product_match <- data %>%
    inner_join(reference_table, 
               by = c("COMMODITY_DISPLAY_NAME" = "ITEM", "entry_type" = "entry_type"), 
               relationship = "many-to-many",
               copy = TRUE)
  
  return(asf_product_match)
}

# Apply to eans
at_risk_ibl <- match_asf_risk(ibl_data, asf_product_ref)
```

# Could do sampling to predict based on a volume
- Not currently helpful, we don't know how many IBLs might come in
  - otherwise, it would be 1) change the at_risk_ibl to be only if the item exists, 2) predict the likelihood of the IBL having risk in FL vs PR, 3) create distribution of kgs, 4) provide a volume number, and 5) sample from that approach rate and size distribution to create a predicted volume of risk coming in

- BUT for now, just leave it as is

# Filter dates
```{r filter dates}
#---- Filter by years specified in params ----
ibl_data_filtered <- at_risk_ibl %>%
  mutate(start_date = as.Date(start_date)) %>%
  filter(lubridate::year(start_date) %in% params$years)

```

```{r save data}
#---- Save cleaned and merged data ----

#save the R file for picking up in the next script
saveRDS(ibl_data_filtered, here::here("output", "intermediate files",(params$ibl_geolocated_filename)))

#save a copy of the full dataset as a .csv
write_csv(ibl_data, file.path(output_dir, "all_ibl_data.csv"))

#save a copy of the filtered data set as a .csv
write_csv(ibl_data_filtered, file.path(output_dir, paste0(min(params$years), "_to_", max(params$years),"_ibl_data.csv")))
```