---
title: "IBL geolocations and layers for port and destination"
author: "Dalon White"
date: "2025-09-30"
output: html_document

params:
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  ibl_geolocated_filename: "ibl_records_geocoded"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(here)
library(odbc)
library(DBI)
```

# Parameters

```{r parameters, include=FALSE}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

```

# Access lat lon data for the correct years

```{r load-ibl-data}
ibl_data_filtered <- readRDS(here::here("output", "intermediate files", params$ibl_geolocated_filename))

## convert to spatial with the same CRS
ibl_data_sf <- ibl_data_filtered |>
  filter(!is.na(lon) & !is.na(lat)) |> 
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  # Transform to NAD83/Albers (EPSG:5070) for analysis
  sf::st_transform(crs = 5070)

# Save spatial object for future use
saveRDS(ibl_data_sf, here::here("output", "intermediate files", paste0(params$ibl_geolocated_filename,"_sf")))

```

## 1.2 Filter to US Boundaries

```{r us-boundaries}
# Load US boundaries 
us_bounds_sf <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_bounds.rds"))
us_grid <- readRDS(here::here("output", "intermediate files", "US hex grid", "us_grid.rds"))

# Filter spatial data to US boundaries
ibl_data_sf_filtered <- ibl_data_sf %>%
  st_join(us_bounds_sf, left = TRUE) 

# Report filtering statistics
total_points <- nrow(ibl_data_sf)
kept_points <- nrow(ibl_data_sf_filtered)
removed_points <- total_points - kept_points

cat(sprintf("Filtered to %d points within US boundaries (%d removed)\n", 
          kept_points, removed_points))
```

# Next Steps

9/30/25
Need to convert the IBLs to closest port locatoin
 - this would require import PoE locations and assigning closest to locations
    - then summarize by PoE, pathway, container, product_pathway, product_subpathway

```{r collect poe locs}

geo_conn <- dbConnect(odbc::odbc(),.connection_string = 
                     "Driver=SQL Server;
                      Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                      Database=GeoLocations;
                      trusted_connection=yes")


cbp_poe_cols <- c("CBP_NAME", "CBP_CODE", "LAT", "LON", "Port of Entry Name")


# Get CBP Port of Entry data with coordinates
cbp_poe <- tbl(geo_conn, sql("SELECT * FROM [GeoLocations].[dbo].[tbl_CBP_POEs]")) %>% 
           dplyr::select(all_of(cbp_poe_cols)) %>% 
           collect()


```

```{r assign closest POE by lat lon}
# Convert CBP POE to sf object
cbp_poe_sf <- cbp_poe  %>%
  filter(!is.na(LON) & !is.na(LAT)) %>%
  # Add row_id column to use for joining later
  mutate(row_id = row_number()) %>%
  sf::st_as_sf(coords = c("LON", "LAT"), crs = 4326) %>%
  sf::st_transform(crs = 5070)
  
# Find nearest POE for each IBL point
ibl_with_poe <- sf::st_nearest_feature(ibl_data_sf_filtered, cbp_poe_sf)

# Join POE details back to IBL data
ibl_data_with_poe <- ibl_data_sf_filtered %>%
  mutate(POE_ID = ibl_with_poe)
  
# Join using the row_id we created
ibl_data_with_poe <- ibl_data_with_poe %>%
  left_join(cbp_poe_sf %>%
              st_set_geometry(NULL),
            by = c("POE_ID" = "row_id"))

# Extract state code from "Port of Entry Name" which typically has format "Name, ST"
ibl_data_with_poe <- ibl_data_with_poe %>%
  mutate(LOCATION_STATE_CODE = str_extract(`Port of Entry Name`, "(?<=, )\\w{2}$"))
```


# 1.3 Summarize Data for Spatial Layers

```{r summarize-for-spatial-layers}
# Group by location and pathway combinations and summarize
spatial_layers_summary <- ibl_data_with_poe %>% 
  dplyr::rename(LOCATION_ID = CBP_CODE,
                LOCATION_NAME = CBP_NAME) %>%
  # Convert LOCATION_ID to character to match other scenarios (EAN, current, etc.)
  # CBP_CODE can be character (e.g., "0101") and needs consistent type across all port layers
  mutate(LOCATION_ID = as.character(LOCATION_ID)) %>%
  group_by(LOCATION_ID, LOCATION_NAME, LOCATION_STATE_CODE, PATHWAY, 
           container, product_pathway, product_subpathway) %>%
  summarize(
    total_mean_kg = sum(Animal.kg, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Filter out incomplete records
  filter(!is.na(container) & !is.na(product_pathway) & !is.na(product_subpathway))


```



# 7. Create and Save Spatial Layers Function

```{r create-spatial-layers-function}
# Function to create and save individual spatial layers for each pathway combination
create_spatial_layers <- function(summarized_data, save_files = FALSE, output_dir = NULL) {
  # Split data by pathway combinations
  split_data <- summarized_data %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    group_split()
  
  cat(sprintf("Created %d spatial layers for different pathway combinations\n", length(split_data)))
  
  # Create list to store the layers
  spatial_layers <- list()

  # Initialize mapping dataframe
  mapping_df <- data.frame(
    short_name = character(),
    full_name = character(),
    pathway = character(),
    container = character(),
    product_pathway = character(),
    product_subpathway = character(),
    stringsAsFactors = FALSE
  )

  # Create scenario string that specifies IBLs
  scenario_string <- "IBL"
  
  # Process each split
  for (i in seq_along(split_data)) {
    current_split <- split_data[[i]]
    
    # Get combination values for naming
    combo <- current_split %>%
      dplyr::select(PATHWAY, container, product_pathway, product_subpathway) %>%
      slice(1) %>%
      as.list()
    
    # Create clean name for layer using abbreviations (short name)
    layer_name <- paste(
      substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 31),
      substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3),
      substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3),
      sep = "_"
    )
    
    # Create full descriptive name for reference
    full_combo_name <- paste(
      gsub("[^[:alnum:]]", "_", combo$PATHWAY),
      gsub("[^[:alnum:]]", "_", combo$container),
      gsub("[^[:alnum:]]", "_", combo$product_pathway),
      gsub("[^[:alnum:]]", "_", combo$product_subpathway),
      sep = "_"
    )
    
    # Store in list with short name
    spatial_layers[[layer_name]] <- current_split
    
    # Add mapping entry
    new_mapping <- data.frame(
      short_name = layer_name,
      full_name = full_combo_name,
      pathway = combo$PATHWAY,
      container = combo$container,
      product_pathway = combo$product_pathway,
      product_subpathway = combo$product_subpathway,
      stringsAsFactors = FALSE
    )
    mapping_df <- rbind(mapping_df, new_mapping)
    
    # Save to file if requested
    if (save_files && !is.null(output_dir)) {
      # Ensure output directory exists
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Convert to POINT geometry (required for GeoPackage)
      export_layer <- current_split %>%
        sf::st_cast("POINT")
      
      # Include scenario in filename - use GeoPackage format for Shiny compatibility
      file_path <- file.path(output_dir, paste0(layer_name, "_", scenario_string, ".gpkg"))
      
      # Try to save the file
      tryCatch({
        sf::st_write(export_layer, file_path, delete_layer = TRUE)
        cat(sprintf("Saved layer %d/%d: %s\n", i, length(split_data), file_path))
      }, error = function(e) {
        cat(sprintf("ERROR saving layer %d/%d: %s - %s\n", 
                   i, length(split_data), layer_name, e$message))
      })
    }
  }
    # Calculate totals directly from the original summarized data to ensure consistency
  # This approach guarantees the same results as the manual grouping and summarization
  layer_summary <- summarized_data %>%
    sf::st_drop_geometry() %>%
    group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
    summarize(
      total_risk = sum(total_mean_kg, na.rm = TRUE),
      point_count = n(),
      .groups = 'drop'
    ) %>%
    mutate(
      layer_name = paste(
        substr(gsub("[^[:alnum:]]", "", PATHWAY), 1, 31),
        substr(gsub("[^[:alnum:]]", "", container), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_pathway), 1, 3),
        substr(gsub("[^[:alnum:]]", "", product_subpathway), 1, 3),
        sep = "_"
      ),
      scenario = scenario_string
    ) %>%
    dplyr::select(layer_name, scenario, point_count, total_risk) %>%
    arrange(desc(total_risk))  
  # Print summary
  cat(paste0("\nSummary of spatial layers for scenario '", scenario_string, "' (top 10 by risk):\n"))
  print(head(layer_summary, 10))
  
  # Also save the summary table and mapping table as CSV for reference
  if (save_files && !is.null(output_dir)) {
    summary_path <- file.path(output_dir, paste0("layer_summary_", scenario_string, ".csv"))
    write.csv(layer_summary, summary_path, row.names = FALSE)
    cat(sprintf("\nSaved summary table to: %s\n", summary_path))
    
    # Save the name mapping file
    mapping_path <- file.path(output_dir, paste0("filename_mapping_", scenario_string, ".csv"))
    write.csv(mapping_df, mapping_path, row.names = FALSE)
    cat(sprintf("Saved mapping table to: %s\n", mapping_path))
  }
  
  return(list(
    layers = spatial_layers,
    summary = layer_summary,
    mapping = mapping_df,
    scenario = scenario_string
  ))
}
```

# 8 Create and Save Spatial Layers
```{r create-port-risk-layers}
# Set output directory for port risk layers
port_risk_output_dir <- here::here("output", "spatial layers", "risk at ports")

# Create the spatial layers
spatial_layers <- create_spatial_layers(
  summarized_data = spatial_layers_summary,
  save_files = TRUE,
  output_dir = port_risk_output_dir
)
```


# destination layer

The destination layer needs to take all observations and assign them to the hex grid polygons
In essence, this process should just duplicate the destination layer creation in 01_01
 In this case, we're aggregating point data to the hex grid

```{r working on destinations}
# Aggregate IBL points to hex grid cells by summing Animal.kg values
us_grid <- us_grid %>% mutate(grid_id = row_number())
# Perform spatial join to assign each IBL point to a grid cell
ibl_with_grid <- ibl_data_sf_filtered %>%
  st_join(us_grid, left = FALSE)  # left = FALSE drops points outside the grid

# Check how many points were assigned to grid cells
cat(sprintf("Total IBL points: %d\n", nrow(ibl_data_sf_filtered)))
cat(sprintf("Points assigned to grid cells: %d\n", nrow(ibl_with_grid)))
cat(sprintf("Points outside grid: %d\n", nrow(ibl_data_sf_filtered) - nrow(ibl_with_grid)))

# Group by grid cell and sum Animal.kg values
ibl_grid_aggregated <- ibl_with_grid %>%
  st_drop_geometry() %>%  # Drop point geometry
  group_by(grid_id) %>%  # Assuming us_grid has a grid_id column; adjust if different
  summarize(
    total_animal_kg = sum(Animal.kg, na.rm = TRUE),
    n_points = n(),
    .groups = 'drop'
  )

# Join back to the grid to get polygon geometries
ibl_destination_layer <- us_grid %>%
  left_join(ibl_grid_aggregated, by = "grid_id") %>%  # Adjust join column as needed
  # Replace NA values with 0 for cells with no IBL points
  mutate(
    total_animal_kg = ifelse(is.na(total_animal_kg), 0, total_animal_kg),
    n_points = ifelse(is.na(n_points), 0, n_points)
  ) %>%
  # Filter to only cells with risk > 0
  filter(total_animal_kg > 0)

# Summary statistics
cat(sprintf("\nDestination layer summary:\n"))
cat(sprintf("- Grid cells with IBL risk: %d\n", nrow(ibl_destination_layer)))
cat(sprintf("- Total Animal.kg: %.2f\n", sum(ibl_destination_layer$total_animal_kg, na.rm = TRUE)))
cat(sprintf("- Mean Animal.kg per cell: %.2f\n", mean(ibl_destination_layer$total_animal_kg, na.rm = TRUE)))
cat(sprintf("- Max Animal.kg in a cell: %.2f\n", max(ibl_destination_layer$total_animal_kg, na.rm = TRUE)))

# View the aggregated data
mapview::mapview(ibl_destination_layer, zcol = "total_animal_kg")
```


# Destination risk layers have been implemented above using hex grid aggregation













# 9. Prepare Data for Final Destination Risk Layers

```{r prepare-final-destination-data}
# Summarize by pathway combination (dropping geometry)
pathway_spatial_layers_summary <- spatial_layers_summary %>% 
  sf::st_drop_geometry() %>%
  group_by(PATHWAY, container, product_pathway, product_subpathway) %>%
  summarize(
    #total_mean_kg is not actually the mean; in the other pathways, the 'mean' is the mean simulations but in this case there is only 1 simulation - reality. So, keeping the name for naming consistency, but leaving a note in case of confusion
    total_mean_kg = sum(total_mean_kg, na.rm = TRUE),
    .groups = 'drop'
  )

setwd(here::here('output','report files'))
write.csv(pathway_spatial_layers_summary, paste("IBL pathway summary", min(params$years), "to", max(params$years),".csv", sep = "_"), row.names = FALSE)
```

# 10. Create Final Destination Risk Layers Function

```{r distribute-pathway-risks-function}
# Function to create IBL destination risk layers from aggregated hex grid data
distribute_pathway_risks <- function(pathway_summary, aggregated_grid_data) {
  # Create output directory for final destination risk layers
  output_dir <- here::here('output', 'spatial layers', 'risk at final destinations')
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Initialize data structures
  distributed_layers <- list()
  mapping_df <- data.frame(
    short_name = character(),
    full_name = character(),
    pathway = character(),
    container = character(),
    product_pathway = character(),
    product_subpathway = character(),
    stringsAsFactors = FALSE
  )
    # For IBL data, we need to create destination layers from the aggregated hex grid
  # The aggregated_grid_data contains the total_animal_kg summed across all pathways
  # We'll distribute this proportionally based on the pathway_summary proportions
  
  # Process each combination in pathway_summary
  for (i in seq_len(nrow(pathway_summary))) {
    combo <- pathway_summary[i,]
    
    cat(sprintf("Processing combination %d/%d: %s\n", i, nrow(pathway_summary), combo$PATHWAY))
    
    # Create abbreviated name components for file naming
    pathway_abbr <- substr(gsub("[^[:alnum:]]", "", combo$PATHWAY), 1, 31)
    container_abbr <- substr(gsub("[^[:alnum:]]", "", combo$container), 1, 3)
    prod_path_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_pathway), 1, 3)
    prod_subpath_abbr <- substr(gsub("[^[:alnum:]]", "", combo$product_subpathway), 1, 3)
    
    # Create short name for file (with unique index)
    short_combo_name <- paste(pathway_abbr, container_abbr, prod_path_abbr, prod_subpath_abbr, i, sep = "_")
    
    # Create full descriptive name for reference
    full_combo_name <- paste(
      gsub("[^[:alnum:]]", "_", combo$PATHWAY),
      gsub("[^[:alnum:]]", "_", combo$container),
      gsub("[^[:alnum:]]", "_", combo$product_pathway),
      gsub("[^[:alnum:]]", "_", combo$product_subpathway),
      sep = "_"
    )
    
    # Calculate the proportion of total risk for this combination
    total_risk_all_combos <- sum(pathway_summary$total_mean_kg, na.rm = TRUE)
    combo_proportion <- combo$total_mean_kg / total_risk_all_combos
    
    # Apply this proportion to the aggregated grid data
    # This distributes the total risk across the grid based on where IBL points were found
    distributed_layer <- aggregated_grid_data %>%
      mutate(
        risk_value = total_animal_kg * combo_proportion,
        pathway = combo$PATHWAY,
        container = combo$container,
        product_pathway = combo$product_pathway,
        product_subpathway = combo$product_subpathway
      ) %>%
      # Only keep cells with risk > 0
      filter(risk_value > 0)
    
    if(nrow(distributed_layer) == 0) {
      cat(sprintf("No grid cells with risk for combination: %s\n", full_combo_name))
      next
    }
      
      # Store in layers list
      distributed_layers[[full_combo_name]] <- distributed_layer
      
      # Add mapping entry
      new_mapping <- data.frame(
        short_name = short_combo_name,
        full_name = full_combo_name,
        pathway = combo$PATHWAY,
        container = combo$container,
        product_pathway = combo$product_pathway,
        product_subpathway = combo$product_subpathway,
        stringsAsFactors = FALSE
      )
      mapping_df <- rbind(mapping_df, new_mapping)
      
      # Save as GeoPackage for Shiny compatibility
      output_file <- file.path(output_dir, paste0("risk_", short_combo_name, ".gpkg"))
      
      tryCatch({
        sf::st_write(distributed_layer, output_file, delete_layer = TRUE)
        total_risk <- sum(distributed_layer$risk_value, na.rm=TRUE)
        cat(sprintf("Processed %s → %s with risk %f (%d points)\n", 
                   full_combo_name, short_combo_name, total_risk, nrow(distributed_layer)))
      }, error = function(e) {
        cat(sprintf("ERROR: Could not write %s - %s\n", short_combo_name, e$message))
      })
    }
  
  
  # Save the name mapping file for reference
  write.csv(mapping_df, file.path(output_dir, "filename_mapping.csv"), row.names = FALSE)
  
  # Create a layer catalog with additional metadata
  layer_catalog <- mapping_df %>%
    mutate(
      file_path = file.path(output_dir, paste0("risk_", short_name, ".gpkg")),
      risk_value = sapply(distributed_layers[full_name], function(x) sum(x$risk_value, na.rm=TRUE)),
      cell_count = sapply(distributed_layers[full_name], nrow)
    ) %>%
    arrange(desc(risk_value))
  
  # Save the catalog
  write.csv(layer_catalog, file.path(output_dir, "layer_catalog.csv"), row.names = FALSE)
  
  # Mark these as IBL scenario data
  scenario_string <- "IBL"
  
  return(list(
    layers = distributed_layers,
    mapping = mapping_df,
    catalog = layer_catalog,
    scenario = scenario_string
  ))
}
```

# 11. Create Final Destination Risk Layers

```{r create-final-destination-layers}
# Apply the function to distribute risks to final destinations
distributed_risks <- distribute_pathway_risks(
  pathway_summary = pathway_spatial_layers_summary,
  aggregated_grid_data = ibl_destination_layer
)

# Save a compact version of the catalog for Shiny app use
risk_summary <- distributed_risks$catalog %>%
  dplyr::select(
    layer_name = full_name, 
    short_name,
    pathway,
    container, 
    product_pathway,
    product_subpathway,
    total_risk = risk_value, 
    cell_count
  )

# Save the summary for reference
write.csv(risk_summary, 
          here::here("output", "aqim total distributed risk", paste0("distributed_risk_summary_IBL.csv")), 
          row.names = FALSE)

cat("Spatial risk layers created and ready for use in Shiny app\n")
```


```{r}
library(sf)
library(here)

# Path to the IBL GeoPackage file
gpkg_path <- here::here("output", "spatial layers", "risk at final destinations", "risk_IBL_bag_swi_per_1.gpkg")

# Check what layers are in the file
layers <- st_layers(gpkg_path)
print(layers)

# Read the data
ibl_data <- st_read(gpkg_path)

# View the data structure
str(ibl_data)

# Plot a basic map
plot(st_geometry(ibl_data))





# Path to the IBL GeoPackage file
gpkg_path <- here::here("output", "spatial layers", "risk at final destinations", "risk_PedestrianSB_bag_oth_uns_1.gpkg")

# Check what layers are in the file
layers <- st_layers(gpkg_path)
print(layers)

# Read the data
ped_data <- st_read(gpkg_path)

# View the data structure
str(ped_data)

# Plot a basic map
plot(st_geometry(ped_data))
```
