---
title: "ASF Risk Spatial Visualization"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 8)
library(tidyverse)
library(sf)
library(leaflet)
library(RColorBrewer)
library(here)
library(DT)
```

## Load Risk Distribution Layers

```{r load_data}
# Set paths
risk_dir <- here::here('output', 'spatial layers', 'risk at final destinations')
new_layers_dir <- here::here('output', 'additional layers')  # Directory for your additional layers

# Load the layer catalog for easier access
layer_catalog <- read.csv(file.path(risk_dir, "layer_catalog.csv"))

# Function to load a specific layer by its short name
load_layer <- function(short_name) {
  file_path <- file.path(risk_dir, paste0("risk_", short_name, ".gpkg"))
  if(file.exists(file_path)) {
    return(st_read(file_path))
  } else {
    warning(paste("Layer not found:", file_path))
    return(NULL)
  }
}

# Function to discover and load additional layers
load_additional_layers <- function(directory, pattern = "*.gpkg") {
  if(!dir.exists(directory)) {
    warning(paste("Directory not found:", directory))
    return(list())
  }
  
  # List all matching files
  layer_files <- list.files(directory, pattern = pattern, full.names = TRUE)
  
  # Load each file into a named list
  layers <- list()
  for(file in layer_files) {
    name <- tools::file_path_sans_ext(basename(file))
    layers[[name]] <- st_read(file)
  }
  
  return(layers)
}

# Optional: Load the complete distributed_risks object if needed
# distributed_risks <- readRDS(file.path(risk_dir, "all_distributed_risks.rds"))
```

## Risk Layer Overview

```{r layer_overview}
# Display catalog of available layers
datatable(layer_catalog %>% 
            select(short_name, pathway, container, product_pathway, product_subpathway, risk_value) %>%
            arrange(desc(risk_value)),
          options = list(pageLength = 10),
          caption = "Available Risk Distribution Layers (Click to Select)")
```

## Map Highest Risk Layers

```{r top_risk_map}
# Get the top 5 highest risk layers
top_layers <- head(layer_catalog, 15)

# US states for background
us_states <- tigris::states(cb = TRUE)

# Load and plot top risk layers
for(i in 1:nrow(top_layers)) {
  layer_name <- top_layers$short_name[i]
  layer_data <- load_layer(layer_name)
  
  if(!is.null(layer_data)) {
    # Create title with details
    title <- paste0(
      "Risk Distribution: ", top_layers$pathway[i], 
      " - ", top_layers$container[i],
      " (", round(top_layers$risk_value[i], 2), " kg at risk)"
    )
    
    # Plot using ggplot2
    p <- ggplot() +
      geom_sf(data = us_states, fill = "white", color = "gray70") +
      geom_sf(data = layer_data, aes(fill = risk_value), alpha = 0.7) +
      scale_fill_viridis_c(name = "Risk (kg)", 
                          option = "plasma", 
                          trans = "log1p") +
      labs(title = title,
           subtitle = paste("Product:", top_layers$product_pathway[i], "-", top_layers$product_subpathway[i])) +
      theme_minimal() +
      theme(legend.position = "right")
    
    print(p)
  }
}
```

## Interactive Map of Selected Layer

Select a specific layer to visualize interactively:

```{r interactive_map, fig.width=12, fig.height=8}
# Change this to visualize a different layer
selected_layer_name <- layer_catalog$short_name[1]  # Default: highest risk layer

# Load the selected layer
selected_layer <- load_layer(selected_layer_name)
layer_details <- layer_catalog[layer_catalog$short_name == selected_layer_name, ]

if(!is.null(selected_layer)) {
  # Create leaflet map
  pal <- colorNumeric(
    palette = "YlOrRd",
    domain = selected_layer$risk_value,
    na.color = "transparent"
  )
  
  # Convert to web mercator for leaflet
  selected_layer_wgs <- st_transform(selected_layer, 4326)
  
  leaflet(selected_layer_wgs) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    addPolygons(
      fillColor = ~pal(risk_value),
      weight = 1,
      opacity = 1,
      color = "white",
      dashArray = "3",
      fillOpacity = 0.7,
      highlight = highlightOptions(
        weight = 2,
        color = "#666",
        dashArray = "",
        fillOpacity = 0.9,
        bringToFront = TRUE),
      popup = ~paste0(
        "<strong>Risk Value:</strong> ", round(risk_value, 4), " kg<br>"
      ),
      label = ~paste0("Risk: ", round(risk_value, 4), " kg")
    ) %>%
    addLegend(
      pal = pal,
      values = ~risk_value,
      opacity = 0.7,
      title = "Risk (kg)",
      position = "bottomright"
    ) %>%
    setView(lng = -95, lat = 39, zoom = 4) %>%
    addMiniMap(
      toggleDisplay = TRUE,
      position = "bottomleft"
    )
}
```

## Combining Multiple Risk Layers

You can combine multiple layers to see their cumulative effect:

```{r combine_layers}
# Select multiple layers to combine (e.g., top 3)
layers_to_combine <- head(layer_catalog$short_name, 3)

# Function to combine multiple spatial layers
combine_risk_layers <- function(layer_names) {
  # Use a different approach for creating the base grid
  # Get US counties as the base grid instead of trying to create Voronoi cells
  us_grid <- tigris::counties(cb = TRUE) %>%
    st_transform(5070) %>%  # Use Albers projection
    # No need for complex voronoi/casting operations
    st_sf() %>%
    mutate(combined_risk = 0)
  
  # Add risk from each layer
  for(name in layer_names) {
    layer <- load_layer(name)
    if(!is.null(layer)) {
      # Ensure compatible CRS
      layer <- st_transform(layer, st_crs(us_grid))
      
      # Spatial join and sum risk
      temp_join <- st_join(us_grid, layer %>% select(risk_value))
      
      # Aggregate risk by county
      us_grid <- temp_join %>%
        group_by(GEOID) %>%  # Group by county ID
        summarize(
          combined_risk = sum(combined_risk, na.rm = TRUE) + sum(risk_value, na.rm = TRUE),
          .groups = "drop"
        )
    }
  }
  
  return(us_grid)
}

# Create combined layer
combined_layer <- combine_risk_layers(layers_to_combine)

# Plot combined layer
ggplot() +
  geom_sf(data = us_states %>% st_transform(st_crs(combined_layer)), 
          fill = "white", color = "gray70") +
  geom_sf(data = combined_layer, aes(fill = combined_risk), alpha = 0.7) +
  scale_fill_viridis_c(name = "Combined Risk (kg)", 
                       option = "plasma", 
                       trans = "log1p") +
  labs(title = paste("Combined Risk from", length(layers_to_combine), "Highest Risk Pathways"),
       subtitle = paste("Total kg at risk:", round(sum(combined_layer$combined_risk, na.rm=TRUE), 2))) +
  theme_minimal() +
  theme(legend.position = "right")
```

## Incorporate New Layers

You can load and incorporate additional layers from other sources:

```{r additional_layers}
# Load additional layers (add your specific layers here)
additional_layers <- load_additional_layers(new_layers_dir)

# Example: If you have a layer with farm locations
if(length(additional_layers) > 0) {
  # Get the first layer for demonstration
  add_layer_name <- names(additional_layers)[1]
  add_layer <- additional_layers[[add_layer_name]]
  
  # Get one of our risk layers
  risk_layer <- load_layer(layer_catalog$short_name[1])
  
  if(!is.null(risk_layer) && !is.null(add_layer)) {
    # Ensure compatible CRS
    add_layer <- st_transform(add_layer, st_crs(risk_layer))
    
    # Plot together
    ggplot() +
      geom_sf(data = us_states %>% st_transform(st_crs(risk_layer)), 
              fill = "white", color = "gray70") +
      geom_sf(data = risk_layer, aes(fill = risk_value), alpha = 0.7) +
      geom_sf(data = add_layer, color = "black", size = 1) +
      scale_fill_viridis_c(name = "Risk (kg)", 
                           option = "plasma", 
                           trans = "log1p") +
      labs(title = paste("Risk Layer with Additional", add_layer_name, "Layer"),
           subtitle = "Combining risk distribution with external data") +
      theme_minimal()
  } else {
    cat("Either risk layer or additional layer not available\n")
  }
} else {
  cat("No additional layers found in the specified directory\n")
}
```

## Creating Custom Visualizations

You can create custom visualizations by filtering and combining layers:

```{r custom_viz}
# Example: Select layers by pathway type
passenger_layers <- layer_catalog %>%
  filter(grepl("Person|Passenger", pathway)) %>%
  pull(short_name)

cargo_layers <- layer_catalog %>%
  filter(grepl("Cargo|Shipment", pathway)) %>%
  pull(short_name)

# Function to summarize risk by pathway type
summarize_pathway_risk <- function(layer_names) {
  total_risk <- 0
  
  for(name in layer_names) {
    layer <- load_layer(name)
    if(!is.null(layer)) {
      layer_risk <- sum(layer$risk_value, na.rm = TRUE)
      total_risk <- total_risk + layer_risk
    }
  }
  
  return(total_risk)
}

# Calculate total risk by pathway type
passenger_risk <- summarize_pathway_risk(passenger_layers)
cargo_risk <- summarize_pathway_risk(cargo_layers)

# Create comparative bar chart
comparison_data <- data.frame(
  Pathway_Type = c("Passenger", "Cargo"),
  Risk_kg = c(passenger_risk, cargo_risk)
)

ggplot(comparison_data, aes(x = Pathway_Type, y = Risk_kg, fill = Pathway_Type)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Risk_kg, 1)), vjust = -0.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Comparison of Risk by Pathway Type",
       subtitle = "Total kg at risk by passenger vs cargo pathways",
       x = "", y = "Risk (kg)") +
  theme_minimal() +
  theme(legend.position = "none")
```



```{r multi_layer_map, fig.width=12, fig.height=8}
# Create layer categories for better organization
layer_categories <- list(
  "High Risk (Top 5)" = head(layer_catalog$short_name, 5),
  "Passenger Pathways" = layer_catalog %>% 
    filter(grepl("Person|Passenger", pathway)) %>% 
    pull(short_name),
  "Cargo Pathways" = layer_catalog %>% 
    filter(grepl("Cargo|Shipment", pathway)) %>% 
    pull(short_name),
  "Swine Products" = layer_catalog %>%
    filter(grepl("swine", product_pathway, ignore.case = TRUE)) %>%
    pull(short_name)
)

# Create base map
map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = -95, lat = 39, zoom = 4) %>%
  addMiniMap(toggleDisplay = TRUE, position = "bottomleft")

# Create color palette function that will work across all layers
max_risk <- max(layer_catalog$risk_value, na.rm = TRUE)
pal <- colorNumeric(palette = "YlOrRd", domain = c(0, max_risk))

# Track all layer groups for the layer control
all_groups <- c()
active_layers <- c() # Layers to show initially

# Add individual layers by category
for(category_name in names(layer_categories)) {
  # Get layers for this category
  category_layers <- layer_categories[[category_name]]
  
  # Create a collapsed section heading for this category
  map <- map %>% addLayersControl(
    overlayGroups = category_name,
    options = layersControlOptions(collapsed = FALSE)
  )
  
  # Prepare a combined layer for this category (for efficiency)
  combined_layer <- combine_risk_layers(category_layers) 
  combined_name <- paste0("ALL: ", category_name)
  
  # Add the combined layer
  if(!is.null(combined_layer)) {
    map <- map %>% addPolygons(
      data = st_transform(combined_layer, 4326),
      fillColor = ~pal(combined_risk),
      weight = 1,
      opacity = 0.8,
      color = "white",
      dashArray = "3",
      fillOpacity = 0.7,
      group = combined_name,
      popup = ~paste0("<strong>", category_name, " Combined Risk:</strong> ", 
                    round(combined_risk, 4), " kg"),
      label = ~paste0(category_name, ": ", round(combined_risk, 4), " kg")
    )
    all_groups <- c(all_groups, combined_name)
    active_layers <- c(active_layers, combined_name)  # Show combined layers by default
  }
  
  # Add individual layers (limit to first 10 for performance)
  for(i in 1:min(10, length(category_layers))) {
    layer_name <- category_layers[i]
    layer_details <- layer_catalog[layer_catalog$short_name == layer_name, ]
    
    # Skip if no details found
    if(nrow(layer_details) == 0) next
    
    # Create descriptive name
    display_name <- paste0(
      substr(layer_details$pathway, 1, 15), "...: ",
      substr(layer_details$product_pathway, 1, 15), " (", 
      round(layer_details$risk_value, 2), " kg)"
    )
    
    # Load layer
    layer <- load_layer(layer_name)
    if(is.null(layer)) next
    
    # Add layer to map
    map <- map %>% addPolygons(
      data = st_transform(layer, 4326),
      fillColor = ~pal(risk_value),
      weight = 1,
      opacity = 0.8,
      color = "black",  # Different color for individual layers
      dashArray = "3",
      fillOpacity = 0.7,
      group = display_name,
      popup = ~paste0(
        "<strong>Layer:</strong> ", display_name, "<br>",
        "<strong>Risk:</strong> ", round(risk_value, 4), " kg"
      )
    )
    
    all_groups <- c(all_groups, display_name)
    # Don't add to active_layers - we'll show only combined by default
  }
}

# Add final layer control with all groups
map %>%
  addLayersControl(
    overlayGroups = all_groups,
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    pal = pal,
    values = c(0, max_risk),
    title = "Risk (kg)",
    position = "bottomright"
  ) %>%
  # Hide individual layers by default, show only combined
  hideGroup(setdiff(all_groups, active_layers))
  ```






## Conclusion

This document provides a framework for visualizing and analyzing ASF risk distribution across the US. You can:

1. View individual risk layers 
2. Combine multiple layers to see cumulative effects
3. Incorporate additional external layers
4. Create custom analyses by pathway or other attributes

For additional analyses, you can extend this document with more specific visualizations or statistical summaries based on your specific research questions.