---
title: "Enhanced GLMM by Pathway Approach for ASF Risk Prediction"
author: "Dalon White"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(ggplot2)
library(knitr)
library(kableExtra)
library(here)
library(scales)
library(broom.mixed) # For model diagnostics
library(DHARMa)      # For residual diagnostics
library(car)         # For model diagnostics
library(boot)        # For bootstrapping
```

## Introduction

This document implements an enhanced GLMM approach to the ASF risk prediction problem by:

1. Splitting the data by PATHWAY
2. Fitting a separate GLMM for each pathway
3. Providing multiple methods for calculating confidence intervals
4. Adding comprehensive model diagnostics
5. Implementing fallback models for pathways with convergence issues
6. Comparing results with the Bayesian approach (if available)

This approach simplifies the model structure compared to a full hierarchical Bayesian model while still accounting for the hierarchical structure within each pathway.

## Data Preparation

First, let's load the same data used in the Bayesian approach:

```{r load_data}
# Load the aqim_events_summary data
if (file.exists(here::here("output", "intermediate files", "aqim_events_summary.rds"))) {
  aqim_events_summary <- readRDS(here::here("output", "intermediate files", "aqim_events_summary.rds"))
} else {
  stop("Could not find aqim_events_summary data. Please run the data preparation steps first.")
}

# Prepare the data using the same process as in the Bayesian model
bayes_data <- aqim_events_summary %>%
  # Remove any rows with missing values or where total_obs is 0
  filter(!is.na(n_positive), !is.na(total_obs), total_obs > 0) %>%
  # Ensure n_positive doesn't exceed total_obs (data validation)
  mutate(n_positive = pmin(n_positive, total_obs)) %>%
  # Create a unique ID for each observation
  mutate(obs_id = row_number())

# Display data dimensions and structure
cat("Data dimensions:", dim(bayes_data), "\n")
cat("Number of unique pathways:", length(unique(bayes_data$PATHWAY)), "\n")
cat("Number of locations:", length(unique(bayes_data$INSPECTION_LOCATION_NAME)), "\n")
cat("Total positive events:", sum(bayes_data$n_positive), "\n")
cat("Average positive rate:", 
    round(100 * sum(bayes_data$n_positive) / sum(bayes_data$total_obs), 4), "%\n")

# Show distribution of observations by pathway
pathway_counts <- bayes_data %>%
  group_by(PATHWAY) %>%
  summarize(
    observations = n(),
    locations = n_distinct(INSPECTION_LOCATION_NAME),
    positives = sum(n_positive),
    positive_rate = round(100 * sum(n_positive) / sum(total_obs), 2)
  ) %>%
  arrange(desc(observations))

kable(pathway_counts, caption = "Data summary by pathway") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Enhanced GLMM by Pathway Approach

We'll implement an enhanced version of the pathway-specific GLMM approach with multiple methods for calculating confidence intervals and improved diagnostics:

```{r define_enhanced_function}
# Function to fit GLMM for a single pathway with multiple CI calculation methods
fit_pathway_glmm <- function(pathway_data, pathway_name, 
                           method = c("simple", "bootstrap", "profile"),
                           bootstrap_reps = 100,  # Reduced for demonstration
                           fallback = TRUE) {
  # Default to simple method if not specified
  method <- match.arg(method)
  
  # Print info about this pathway
  cat("\n--------------------------------------------------------\n")
  cat("Processing pathway:", pathway_name, "\n")
  cat("Number of observations:", nrow(pathway_data), "\n")
  cat("Number of locations:", length(unique(pathway_data$INSPECTION_LOCATION_NAME)), "\n")
  cat("Total positive events:", sum(pathway_data$n_positive), "\n")
  cat("Average positive rate:", 
      round(100 * sum(pathway_data$n_positive) / sum(pathway_data$total_obs), 4), "%\n")
  
  # Skip pathways with insufficient data
  if (nrow(pathway_data) < 5) {
    cat("WARNING: Too few observations (< 5) for pathway", pathway_name, "- skipping\n")
    return(NULL)
  }
  
  if (sum(pathway_data$n_positive) == 0) {
    cat("WARNING: No positive events for pathway", pathway_name, "- skipping\n")
    return(NULL)
  }
  
  # Try to fit the model with error handling
  model_result <- tryCatch({
    # Fit the GLMM model
    model <- glmer(
      cbind(n_positive, total_obs - n_positive) ~ 1 + 
        (1 | INSPECTION_LOCATION_NAME) + 
        (1 | container) + 
        (1 | product_pathway) +
        (1 | product_pathway:product_subpathway),
      data = pathway_data,
      family = binomial("logit"),
      control = glmerControl(optimizer = "bobyqa")
    )
    
    # Get fixed effect (global intercept)
    fixed_effect <- fixef(model)[1]
    cat("Fixed effect (intercept):", round(fixed_effect, 4), "\n")
    cat("Overall baseline probability:", round(plogis(fixed_effect), 4), "\n")
    
    # Calculate predicted probabilities based on the selected method
    if (method == "simple") {
      # Method 1: Simple fitted values with delta method CI
      fitted_values <- predict(model, newdata = pathway_data, type = "link")
      fitted_probs <- plogis(fitted_values)
      
      # Calculate standard errors using the delta method
      se_link <- sqrt(fitted_probs * (1 - fitted_probs) / pathway_data$total_obs)
      
      # Calculate confidence intervals
      lower_probs <- plogis(fitted_values - 1.96 * se_link)
      upper_probs <- plogis(fitted_values + 1.96 * se_link)
      
    } else if (method == "bootstrap") {
      # For demonstration, we'll use a simplified bootstrap approach
      # In a real application, you would use parametric bootstrap with the full model
      
      cat("Running parametric bootstrap with", bootstrap_reps, "replications...\n")
      
      # Get predicted values on the link scale
      fitted_values <- predict(model, newdata = pathway_data, type = "link")
      fitted_probs <- plogis(fitted_values)
      
      # Function to generate bootstrap samples
      boot_func <- function(data, indices) {
        # Create bootstrap sample
        boot_sample <- data[indices, ]
        
        # Fit model to bootstrap sample
        boot_model <- tryCatch({
          glmer(
            cbind(n_positive, total_obs - n_positive) ~ 1 + 
              (1 | INSPECTION_LOCATION_NAME) + 
              (1 | container) + 
              (1 | product_pathway) +
              (1 | product_pathway:product_subpathway),
            data = boot_sample,
            family = binomial("logit"),
            control = glmerControl(optimizer = "bobyqa")
          )
        }, error = function(e) {
          # If complex model fails, try simpler model
          glmer(
            cbind(n_positive, total_obs - n_positive) ~ 1 + 
              (1 | INSPECTION_LOCATION_NAME),
            data = boot_sample,
            family = binomial("logit"),
            control = glmerControl(optimizer = "bobyqa")
          )
        })
        
        # Return fixed effect (intercept)
        return(fixef(boot_model)[1])
      }
      
      # Run bootstrap
      set.seed(12345)
      boot_results <- boot(data = pathway_data, statistic = boot_func, R = bootstrap_reps)
      
      # Calculate bootstrap confidence intervals for the intercept
      boot_ci <- boot.ci(boot_results, type = "perc")
      
      # Apply the CI to the predictions
      intercept_lower <- boot_ci$percent[4]
      intercept_upper <- boot_ci$percent[5]
      
      # Calculate CI for each prediction
      intercept_shift_lower <- intercept_lower - fixed_effect
      intercept_shift_upper <- intercept_upper - fixed_effect
      
      lower_probs <- plogis(fitted_values + intercept_shift_lower)
      upper_probs <- plogis(fitted_values + intercept_shift_upper)
      
    } else if (method == "profile") {
      # Profile likelihood method
      cat("Using profile likelihood for confidence intervals...\n")
      
      # Get predictions on link scale
      fitted_values <- predict(model, newdata = pathway_data, type = "link")
      fitted_probs <- plogis(fitted_values)
      
      # For demonstration, we'll use a simplified approach
      # In practice, you would compute profile CIs for each prediction
      
      # Get profile CI for fixed effect
      prof <- suppressWarnings(profile(model))
      ci_fixed <- confint(prof, level = 0.95)
      
      # Use the CI of the fixed effect to calculate prediction intervals
      intercept_lower <- ci_fixed[1, 1]
      intercept_upper <- ci_fixed[1, 2]
      
      # Calculate CI for each prediction
      intercept_shift_lower <- intercept_lower - fixed_effect
      intercept_shift_upper <- intercept_upper - fixed_effect
      
      lower_probs <- plogis(fitted_values + intercept_shift_lower)
      upper_probs <- plogis(fitted_values + intercept_shift_upper)
    }
    
    # Calculate model diagnostics
    model_summary <- summary(model)
    AIC_val <- AIC(model)
    BIC_val <- BIC(model)
    
    # Get random effects variances
    var_comps <- as.data.frame(VarCorr(model))
    var_summary <- var_comps %>%
      select(grp, vcov) %>%
      rename(Random_Effect = grp, Variance = vcov)
    
    # Create diagnostics dataframe
    diagnostics <- data.frame(
      pathway = pathway_name,
      AIC = AIC_val,
      BIC = BIC_val,
      n_obs = nrow(pathway_data),
      n_positive = sum(pathway_data$n_positive),
      positive_rate = sum(pathway_data$n_positive) / sum(pathway_data$total_obs),
      intercept = fixed_effect,
      baseline_prob = plogis(fixed_effect),
      method = method
    )
    
    # Create results data frame
    results <- pathway_data %>%
      mutate(
        pathway = pathway_name,
        pred_prob = fitted_probs,
        lower_prob = lower_probs,
        upper_prob = upper_probs,
        observed_rate = n_positive / total_obs,
        method = method
      )
    
    cat("Model successfully fit for pathway:", pathway_name, "\n")
    return(list(
      model = model, 
      results = results, 
      diagnostics = diagnostics,
      var_components = var_summary
    ))
    
  }, error = function(e) {
    cat("ERROR fitting model for pathway", pathway_name, ":", conditionMessage(e), "\n")
    
    if (fallback) {
      cat("Trying simplified binomial model without random effects...\n")
      
      # Try a simple binomial GLM as fallback
      tryCatch({
        # Fit a simple binomial model
        simple_model <- glm(
          cbind(n_positive, total_obs - n_positive) ~ 1,
          data = pathway_data,
          family = binomial("logit")
        )
        
        # Get predictions
        fitted_probs <- predict(simple_model, type = "response")
        ci <- confint(simple_model)
        lower_probs <- plogis(qlogis(fitted_probs) + ci[1] - coef(simple_model)[1])
        upper_probs <- plogis(qlogis(fitted_probs) + ci[2] - coef(simple_model)[1])
        
        # Create results data frame
        results <- pathway_data %>%
          mutate(
            pathway = pathway_name,
            pred_prob = fitted_probs,
            lower_prob = lower_probs,
            upper_prob = upper_probs,
            observed_rate = n_positive / total_obs,
            method = "fallback_glm"
          )
        
        # Create diagnostics dataframe
        diagnostics <- data.frame(
          pathway = pathway_name,
          AIC = AIC(simple_model),
          BIC = BIC(simple_model),
          n_obs = nrow(pathway_data),
          n_positive = sum(pathway_data$n_positive),
          positive_rate = sum(pathway_data$n_positive) / sum(pathway_data$total_obs),
          intercept = coef(simple_model)[1],
          baseline_prob = plogis(coef(simple_model)[1]),
          method = "fallback_glm"
        )
        
        cat("Fallback model successfully fit for pathway:", pathway_name, "\n")
        return(list(
          model = simple_model, 
          results = results, 
          diagnostics = diagnostics,
          var_components = data.frame(Random_Effect = "None", Variance = 0)
        ))
        
      }, error = function(e2) {
        cat("ERROR fitting fallback model:", conditionMessage(e2), "\n")
        return(NULL)
      })
    } else {
      return(NULL)
    }
  })
  
  return(model_result)
}
```

## Demonstration with a Single Pathway

Let's demonstrate the approach with a single pathway before applying it to all pathways:

```{r demonstrate_single_pathway}
# Pick a pathway with sufficient data
test_pathway <- pathway_counts %>% 
  filter(positives >= 10) %>% 
  slice(1) %>% 
  pull(PATHWAY)

# Extract data for this pathway
test_data <- bayes_data %>% filter(PATHWAY == test_pathway)

# Print information about this test pathway
cat("Testing with pathway:", test_pathway, "\n")
cat("Number of observations:", nrow(test_data), "\n")
cat("Number of locations:", length(unique(test_data$INSPECTION_LOCATION_NAME)), "\n")
cat("Total positive events:", sum(test_data$n_positive), "\n")
cat("Average positive rate:", 
    round(100 * sum(test_data$n_positive) / sum(test_data$total_obs), 4), "%\n")

# Fit models using different methods
simple_result <- fit_pathway_glmm(test_data, test_pathway, method = "simple")
bootstrap_result <- fit_pathway_glmm(test_data, test_pathway, method = "bootstrap", bootstrap_reps = 100)
profile_result <- fit_pathway_glmm(test_data, test_pathway, method = "profile")

# Compare baseline probabilities
if (!is.null(simple_result) && !is.null(bootstrap_result) && !is.null(profile_result)) {
  comparison_df <- data.frame(
    Method = c("Simple", "Bootstrap", "Profile"),
    Baseline_Probability = c(
      simple_result$diagnostics$baseline_prob,
      bootstrap_result$diagnostics$baseline_prob,
      profile_result$diagnostics$baseline_prob
    ),
    Lower_CI = c(
      min(simple_result$results$lower_prob),
      min(bootstrap_result$results$lower_prob),
      min(profile_result$results$lower_prob)
    ),
    Upper_CI = c(
      max(simple_result$results$upper_prob),
      max(bootstrap_result$results$upper_prob),
      max(profile_result$results$upper_prob)
    ),
    CI_Width = c(
      max(simple_result$results$upper_prob) - min(simple_result$results$lower_prob),
      max(bootstrap_result$results$upper_prob) - min(bootstrap_result$results$lower_prob),
      max(profile_result$results$upper_prob) - min(profile_result$results$lower_prob)
    )
  )
  
  kable(comparison_df, digits = 4, caption = "Comparison of CI methods") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
  # Compare variance components
  var_comp_comparison <- list(
    Simple = simple_result$var_components,
    Bootstrap = bootstrap_result$var_components,
    Profile = profile_result$var_components
  )
  
  # Print variance components for the simple method
  kable(var_comp_comparison$Simple, digits = 4, caption = "Variance components (Simple method)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
  # Create a plot comparing the three methods
  combined_results <- bind_rows(
    simple_result$results %>% mutate(Method = "Simple"),
    bootstrap_result$results %>% mutate(Method = "Bootstrap"),
    profile_result$results %>% mutate(Method = "Profile")
  )
  
  # Plot comparing CI widths
  ggplot(combined_results, aes(x = Method, y = pred_prob, color = Method)) +
    geom_boxplot() +
    geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), width = 0.2, position = position_dodge(0.75)) +
    labs(
      title = paste("Comparison of CI Methods for Pathway:", test_pathway),
      subtitle = "Showing predicted probabilities and confidence intervals",
      y = "Predicted Probability"
    ) +
    theme_minimal()
}
```

## Processing All Pathways

Now let's implement the full workflow to process all pathways:

```{r process_all_pathways, eval=FALSE}
# Function to process all pathways
process_all_pathways <- function(bayes_data, 
                               method = c("simple", "bootstrap", "profile"),
                               bootstrap_reps = 100,
                               fallback = TRUE) {
  # Default to simple method if not specified
  method <- match.arg(method)
  
  # Split data by PATHWAY
  pathways <- unique(bayes_data$PATHWAY)
  cat("Found", length(pathways), "unique pathways to process\n")
  
  # Initialize lists to store results
  all_models <- list()
  all_results <- list()
  all_diagnostics <- list()
  all_var_components <- list()
  
  # Process each pathway
  for (pathway in pathways) {
    # Extract data for this pathway
    pathway_data <- bayes_data %>% filter(PATHWAY == pathway)
    
    # Fit model for this pathway
    result <- fit_pathway_glmm(
      pathway_data = pathway_data, 
      pathway_name = pathway, 
      method = method,
      bootstrap_reps = bootstrap_reps,
      fallback = fallback
    )
    
    # Store results if successful
    if (!is.null(result)) {
      all_models[[pathway]] <- result$model
      all_results[[pathway]] <- result$results
      all_diagnostics[[pathway]] <- result$diagnostics
      all_var_components[[pathway]] <- result$var_components
    }
  }
  
  # Combine all results
  combined_results <- bind_rows(all_results)
  combined_diagnostics <- bind_rows(all_diagnostics)
  combined_var_components <- bind_rows(
    lapply(names(all_var_components), function(p) {
      cbind(pathway = p, all_var_components[[p]])
    })
  )
  
  # Create output directory if it doesn't exist
  output_dir <- here::here("output", "aqim pathway approach rates")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Create method-specific suffix
  method_suffix <- switch(method,
                         simple = "",
                         bootstrap = "_bootstrap",
                         profile = "_profile")
  
  # Save the combined results
  results_file <- file.path(output_dir, paste0("glmm_results_aqim_fraction_at_risk", method_suffix, ".csv"))
  write.csv(combined_results, results_file, row.names = FALSE)
  saveRDS(combined_results, file.path(output_dir, paste0("glmm_results_aqim_fraction_at_risk", method_suffix, ".rds")))
  
  # Save the diagnostics
  diag_file <- file.path(output_dir, paste0("glmm_diagnostics_by_pathway", method_suffix, ".csv"))
  write.csv(combined_diagnostics, diag_file, row.names = FALSE)
  
  # Save the variance components
  var_file <- file.path(output_dir, paste0("glmm_variance_components", method_suffix, ".csv"))
  write.csv(combined_var_components, var_file, row.names = FALSE)
  
  # Save the models
  models_file <- file.path(output_dir, paste0("glmm_models_by_pathway", method_suffix, ".rds"))
  saveRDS(all_models, models_file)
  
  cat("\nResults saved to:", results_file, "\n")
  cat("Diagnostics saved to:", diag_file, "\n")
  cat("Variance components saved to:", var_file, "\n")
  cat("Models saved to:", models_file, "\n")
  
  return(list(
    models = all_models,
    results = combined_results,
    diagnostics = combined_diagnostics,
    var_components = combined_var_components
  ))
}
```

```{r run_subset, eval=FALSE}
# For demonstration, let's process a subset of pathways
# In practice, you would process all pathways

# Select top 5 pathways with positive events
top_pathways <- pathway_counts %>%
  filter(positives > 0) %>%
  head(5) %>%
  pull(PATHWAY)

# Filter data to these pathways
subset_data <- bayes_data %>%
  filter(PATHWAY %in% top_pathways)

# Process this subset
results <- process_all_pathways(subset_data, method = "simple", fallback = TRUE)

# Create visualizations
visualize_results <- function(results, diagnostics = NULL) {
  # Create output directory if it doesn't exist
  output_dir <- here::here("output", "aqim pathway approach rates", "glmm_validation")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # 1. Observed vs. Predicted Plot
  obs_vs_pred_plot <- ggplot(results, aes(x = observed_rate, y = pred_prob, size = total_obs, color = pathway)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    scale_x_continuous(labels = percent_format()) +
    scale_y_continuous(labels = percent_format()) +
    labs(
      title = "Observed vs. Predicted Rates by Pathway",
      subtitle = "Point size represents number of observations",
      x = "Observed Rate (n_positive / total_obs)",
      y = "Predicted Rate",
      size = "Sample Size"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  print(obs_vs_pred_plot)
  
  # 2. Pathway-level summary
  pathway_summary <- results %>%
    group_by(pathway) %>%
    summarize(
      total_obs = sum(total_obs),
      n_positive = sum(n_positive),
      observed_rate = n_positive / total_obs,
      pred_prob = mean(pred_prob),
      lower_prob = mean(lower_prob),
      upper_prob = mean(upper_prob),
      .groups = "drop"
    ) %>%
    arrange(desc(pred_prob))
  
  # Pathway predictions plot
  pathway_plot <- ggplot(pathway_summary, 
                       aes(x = reorder(pathway, pred_prob), y = pred_prob)) +
    geom_point() +
    geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), width = 0.2) +
    geom_point(aes(y = observed_rate), shape = 4, size = 2) +  # Add observed rates
    coord_flip() +
    scale_y_continuous(labels = percent_format()) +
    labs(
      title = "Pathway Risk Probabilities",
      subtitle = "Error bars represent 95% confidence intervals. X marks show observed rates.",
      x = "Pathway",
      y = "Estimated Probability"
    ) +
    theme_minimal()
  
  print(pathway_plot)
  
  # 3. Model diagnostics plot (if available)
  if (!is.null(diagnostics)) {
    # Plot model statistics by pathway
    diag_plot <- ggplot(diagnostics, aes(x = reorder(pathway, baseline_prob), y = baseline_prob)) +
      geom_point(aes(size = n_obs, color = method)) +
      coord_flip() +
      scale_y_continuous(labels = percent_format()) +
      labs(
        title = "Model Diagnostics by Pathway",
        subtitle = "Baseline probability with sample size and method used",
        x = "Pathway",
        y = "Baseline Probability",
        size = "Sample Size",
        color = "Method"
      ) +
      theme_minimal()
    
    print(diag_plot)
  }
  
  return(list(
    pathway_summary = pathway_summary
  ))
}

# Visualize results
summaries <- visualize_results(results$results, results$diagnostics)
```

## Comparing with the Bayesian Approach

If Bayesian results are available, we can compare them with our GLMM approach:

```{r compare_bayesian, eval=FALSE}
# Function to compare GLMM results with the Bayesian model results
compare_with_bayesian <- function(glmm_results, bayesian_file) {
  # Check if Bayesian results exist
  if (!file.exists(bayesian_file)) {
    cat("Bayesian results file not found:", bayesian_file, "\n")
    return(NULL)
  }
  
  # Load Bayesian results
  bayes_results <- readRDS(bayesian_file)
  
  # Merge the results
  if ("pathway" %in% names(glmm_results) && 
      "pathway" %in% names(bayes_results) && 
      "pred_prob" %in% names(glmm_results) && 
      "pred_prob" %in% names(bayes_results)) {
    
    # Prepare Bayesian results
    bayes_summary <- bayes_results %>%
      group_by(pathway) %>%
      summarize(
        bayes_prob = mean(pred_prob, na.rm = TRUE),
        bayes_lower = mean(lower_prob, na.rm = TRUE),
        bayes_upper = mean(upper_prob, na.rm = TRUE),
        .groups = "drop"
      )
    
    # Prepare GLMM results
    glmm_summary <- glmm_results %>%
      group_by(pathway) %>%
      summarize(
        glmm_prob = mean(pred_prob, na.rm = TRUE),
        glmm_lower = mean(lower_prob, na.rm = TRUE),
        glmm_upper = mean(upper_prob, na.rm = TRUE),
        .groups = "drop"
      )
    
    # Merge the results
    comparison <- full_join(glmm_summary, bayes_summary, by = "pathway") %>%
      mutate(
        diff = glmm_prob - bayes_prob,
        relative_diff = diff / bayes_prob,
        overlap = pmin(glmm_upper, bayes_upper) - pmax(glmm_lower, bayes_lower),
        has_overlap = overlap > 0
      )
    
    # Create comparison plot
    comparison_plot <- ggplot(comparison, 
                             aes(x = bayes_prob, y = glmm_prob, size = abs(diff))) +
      geom_point(aes(color = has_overlap)) +
      geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
      scale_x_continuous(labels = percent_format()) +
      scale_y_continuous(labels = percent_format()) +
      scale_color_manual(values = c("red", "blue")) +
      labs(
        title = "GLMM vs. Bayesian Probability Estimates",
        subtitle = "Point size represents absolute difference, color indicates CI overlap",
        x = "Bayesian Estimate",
        y = "GLMM Estimate",
        size = "Absolute Difference",
        color = "CIs Overlap"
      ) +
      theme_minimal()
    
    print(comparison_plot)
    
    return(comparison)
  } else {
    cat("Error: Results data frames don't have the expected structure\n")
    return(NULL)
  }
}

# Look for Bayesian results
bayesian_file <- here::here("output", "aqim pathway approach rates", "bootstrap_results_aqim_fraction_at_risk.rds")
if (file.exists(bayesian_file)) {
  comparison <- compare_with_bayesian(results$results, bayesian_file)
  
  if (!is.null(comparison)) {
    # Show summary statistics of differences
    diff_summary <- comparison %>%
      summarize(
        mean_diff = mean(abs(diff), na.rm = TRUE),
        median_diff = median(abs(diff), na.rm = TRUE),
        max_diff = max(abs(diff), na.rm = TRUE),
        pct_overlap = mean(has_overlap, na.rm = TRUE) * 100
      )
    
    kable(diff_summary, caption = "Summary of differences between GLMM and Bayesian approaches", digits = 4) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  }
}
```

## Running the Full Analysis

To run the full analysis on all pathways, you can use the following code. Note that this may take some time to complete, especially if using the bootstrap or profile methods:

```{r full_analysis, eval=FALSE}
# This code block is not evaluated in this document
# To run the full analysis, set eval=TRUE or run this code separately

# Process all pathways using the simple method
results_simple <- process_all_pathways(bayes_data, method = "simple", fallback = TRUE)

# Create visualizations
summaries_simple <- visualize_results(results_simple$results, results_simple$diagnostics)

# Compare with Bayesian results if available
bayesian_file <- here::here("output", "aqim pathway approach rates", "bootstrap_results_aqim_fraction_at_risk.rds")
if (file.exists(bayesian_file)) {
  comparison <- compare_with_bayesian(results_simple$results, bayesian_file)
}

# Print summary
cat("\nAnalysis completed successfully\n")
cat("Processed", length(results_simple$models), "pathways\n")
cat("Total observations in results:", nrow(results_simple$results), "\n")
```

## Conclusion

This enhanced GLMM approach offers several advantages over the basic implementation:

1. **Multiple Confidence Interval Methods**: You can choose between simple delta method CIs, bootstrap CIs, or profile likelihood CIs depending on your accuracy needs.

2. **Improved Diagnostics**: The approach provides comprehensive model diagnostics including AIC, BIC, and variance component analysis.

3. **Robust Error Handling**: The implementation includes fallback models for pathways with convergence issues.

4. **Comparison with Bayesian Results**: When Bayesian results are available, you can directly compare the two approaches.

5. **Visualization Tools**: The approach includes various visualization tools for exploring the results.

While this approach is computationally simpler than the full Bayesian hierarchical model, it still accounts for the hierarchical structure within each pathway, making it a practical alternative for ASF risk prediction.

The key difference is that this approach splits the data by pathway and fits separate models, while the Bayesian approach fits a single hierarchical model across all pathways. The advantage of splitting is improved convergence and simplified interpretation, but at the potential cost of reduced power for pathways with limited data.
