---
title: "geolocate place names"
author: "Dalon White"
date: "2025-07-18"
output: html_document
---
## Geolocate place names

```{r packages}
pacman::p_load(tidyverse, dplyr)
```

# Get places that have already been geolocated by Amber

```{r get known locations}
#read in the file
setwd(here::here("input","data","place names and geolocation"))
#read all files within
files = file.info(list.files(here::here("input","data","place names and geolocation"), full.names = T))
geolocs = read_csv("city state geolocations.csv")

print('`geolocs` loaded with known geolocations by AT')
```

# Get unknown locations to geolocate
These are being shunted from 01_01.
  Any known matching locations will be in the above file. All AQIM locations are tested against that file. If there is not a match, it gets saved as the unknown location file to be geolocated here.

```{r get unknown locations}

#get locations
setwd(here::here("output","geoplaces",'unknown locations'))
unknown_destinations <- read.csv("unknown locations.csv") 

# #for testing purposes only
# unknown_destinations = unknown_destinations[1:4,]
```

# Correct names with fuzzy matching
Parse unknown locations to correct any misspellings before re-geolocating

## Adjust to use 2 types - first run using OSA which will correct simple typos, then JW for those not found

```{r correct names and re-geolocate}

# Reference table of cities and states (e.g., us_cities)
# us_cities <- data.frame(city = ..., state = ..., lat = ..., lon = ...)
setwd(here::here('input',"data",'place names and geolocation','US places'))
files_gaz = here::here("input","data","place names and geolocation",'US places') |> list.files(full.names = T) |> file.info()
gazetteer <- read.delim(base::rownames(files_gaz)[which.max(files_gaz$mtime)], sep = "\t", header = TRUE)


#reference table of geoplaces from geonames - right now I'm using cities1000 (12/5/25)
files_geo <- here::here("input","data","place names and geolocation",'geonames') |> list.files(full.names = T) |> file.info()
geonames <- read.delim(base::rownames(files_geo)[which.max(files_geo$mtime)], sep = "\t", header = FALSE) |> filter(grepl("US|PR|GU|VI|AS|MP", V9)) #v9 filters to US and territories, see 00_06

#geonames has territory state code in V9 and conus state code in v11; combine
geonames <- geonames |> mutate(
  state = case_when(
    V9 == "US" ~ V11,
    TRUE ~ V9
    )
  )


#standardize city state and lat longs
geonames <- geonames |> mutate(
  state = toupper(state),
  city = tools::toTitleCase(tolower(V3)),
  lat = as.numeric(V5),
  lon = as.numeric(V6)
) |> dplyr::select(state, city, lat, lon)

gazetteer <- gazetteer |> mutate(
  state = toupper(USPS),
  city = tools::toTitleCase(tolower(NAME)),
  lat = as.numeric(INTPTLAT),
  lon = as.numeric(INTPTLONG)
) |> dplyr::select(state, city, lat, lon)

combined_ref = geonames |> bind_rows(gazetteer)
nrow(combined_ref) #check number of rows
combined_ref <- combined_ref |> distinct(state, city, lat, lon, .keep_all = T)
nrow(combined_ref |> distinct(state, city)) #check number of rows again
# Theres ~400 with the same names but distinct lat longs between the 2 datasets

# Check for duplicates
duplicates <- combined_ref |> group_by(state, city) |> filter(n() > 1)
View(duplicates)

#A quick look tells me that these are slight variations in lat long for the same city between the 2 datasets, so we can just keep one of them
# Looking at the first few examples, geonames seems to have more precise lat longs, so we'll keep those when duplicates exist
#For example, Jefferson, GA was lat = 33.4 for gazetteer and 34.1 for geonames; I looked it up and found that google maps has 34.1 as the correct lat
# Since we are binding gazetteer to geonames, it will keep the geonames version when we distinct()
combined_ref <- combined_ref |> distinct(state, city, .keep_all = T)


# Fuzzy match function
correct_city <- function(city, state, ref_table, ref_city_col, ref_state_col, max_dist, method = 'jw') {
  # Handle NA inputs early
  if (is.na(city) || is.na(state)) {
    return(data.frame(corrected_city = NA_character_, lat = NA_real_, lon = NA_real_))
  }
  
  # Normalize input
  state_lookup <- toupper(state)
  city_lookup <- tools::toTitleCase(tolower(city))
  
  # Filter candidates by state (using the normalized state_lookup variable)
  candidates <- ref_table %>%
    mutate(
      city_clean = tools::toTitleCase(tolower(.data[[ref_city_col]])),
      state_clean = toupper(.data[[ref_state_col]])
    ) %>%
    filter(state_clean == state_lookup)
  
  if (nrow(candidates) == 0) {
    return(data.frame(corrected_city = NA_character_, lat = NA_real_, lon = NA_real_))
  }
    # Calculate string distances using Optimal String Alignment (OSA)
  # OSA is more forgiving for typos like substitutions (D -> S)
  dists <- stringdist::stringdist(city_lookup, candidates$city_clean, method = method)
  min_idx <- which.min(dists)
  
  # Only correct if distance is below threshold (for OSA, this is number of edits)
  # max_dist should be something like 2-3 for OSA (2-3 character changes allowed)
  if (length(min_idx) == 0 || dists[min_idx] > max_dist) {
    return(data.frame(corrected_city = NA_character_, lat = NA_real_, lon = NA_real_))
  }

  # # Return the corrected city name
  # candidates$city_clean[min_idx]
   # Return a dataframe with the corrected name, lat, and lon
  return(data.frame(
    corrected_city = candidates$city_clean[min_idx],
    lat = candidates$lat[min_idx],
    lon = candidates$lon[min_idx]
  ))
}
```


# correct the city names
```{r apply name correction}

# Initialize the corrected_city column
unknown_destinations$corrected_city <- NA_character_

unknown_destinations$lat <- NA_character_
unknown_destinations$lon <- NA_character_

# Define a helper function to apply correction only to unmatched rows
apply_correction_sequentially <- function(df, ...) {
  # Identify which rows still need a correction (are NA in corrected_city)
  rows_to_correct_idx <- which(is.na(df$corrected_city))
  
  # If all rows are corrected, do nothing
  if (length(rows_to_correct_idx) == 0) {
    return(df)
  }
  
  # Run mapply only on the subset of rows that need correction
  # mapply will return a list of data frames
  new_corrections_list <- mapply(correct_city, 
                                 df$CITY_DEST_NAME[rows_to_correct_idx], 
                                 df$STATE_DEST_CD[rows_to_correct_idx], 
                                 MoreArgs = list(...),
                                 SIMPLIFY = FALSE) # Return a list
  
  # Convert the list of data frames to a single data frame
  new_corrections_df <- do.call(rbind, new_corrections_list)
  
  # Place the new corrections back into the original dataframe
  df$corrected_city[rows_to_correct_idx] <- new_corrections_df$corrected_city
  df$lat[rows_to_correct_idx] <- new_corrections_df$lat
  df$lon[rows_to_correct_idx] <- new_corrections_df$lon
  
  # Return the updated dataframe
  return(df)



}

# --- Run correction steps sequentially in chunks ---

# Define chunk size and output directory
chunk_size <- 50

#This directory will save the corrected chunks, which will be used to compile the known and unknown locations, and then be deleted and cleared
output_dir <- here::here("output", "geoplaces", "corrected_chunks")
#This will save each correction permanently
saved_corrections_dir <- here::here("output", "geoplaces", "city name corrections")

# Create the output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Loop through the dataframe in chunks
for (i in seq(1, nrow(unknown_destinations), by = chunk_size)) {
  # Define the current chunk
  start_index <- i
  end_index <- min(i + chunk_size - 1, nrow(unknown_destinations))
  chunk <- unknown_destinations[start_index:end_index, ]
  
  # Initialize columns for the chunk
  chunk$corrected_city <- NA_character_
  chunk$lat <- NA_real_
  chunk$lon <- NA_real_
  
  # --- Apply correction logic to the current chunk ---
  # 1. Combined Ref with OSA (best for typos)
  chunk <- apply_correction_sequentially(chunk, ref_table = combined_ref, max_dist = 3, method = 'osa', ref_city_col = "city", ref_state_col = "state")
  
  # 2. Combined Ref with JW (best for different word orders or missing words)
  chunk <- apply_correction_sequentially(chunk, ref_table = combined_ref, max_dist = 0.15, method = 'jw', ref_city_col = "city", ref_state_col = "state")
  
  # --- Save the processed chunk ---
  chunk_filename <- file.path(output_dir, paste0("corrected_chunk_", start_index, "_to_", end_index, ".csv"))
  write.csv(chunk, chunk_filename, row.names = FALSE)

  correction_filename <- file.path(saved_corrections_dir, paste0("city_name_corrections_", start_index, "_to_", end_index, Sys.Date(), ".csv"))
  write.csv(chunk, correction_filename, row.names = FALSE)

  print(paste("Processed and saved chunk:", chunk_filename))
}
```



## Combine files, replace input file, then delete

```{r compile geolocation data and integrate in knowns}

# Point to the directory with the corrected chunk files
output_dir <- here::here("output", "geoplaces", "corrected_chunks")
files <- list.files(output_dir, full.names = TRUE, pattern = "\\.csv$")

# Read and combine all chunk files
all_data <- do.call(rbind, lapply(files, read.csv))

# --- The rest of your processing logic ---
known_places <- all_data |> 
  mutate(x = lon, y = lat) |> 
  #filter out unidentified places
  filter(!(is.na(x) | is.na(y))) |> 
  # Create the citystlow column from the *original* city/state names
  mutate(citystlow = tolower(
    paste(CITY_DEST_NAME, STATE_DEST_CD, sep = ", "))
    ) |> 
  mutate(UniqueID = paste("osm-", row_number())) |> 
  dplyr::select(citystlow, UniqueID, x, y)

# Set working directory to save the final file
setwd(here::here("input","data","place names and geolocation"))
known_places <- rbind(geolocs, known_places)
write.csv(known_places, "city state geolocations.csv", row.names = FALSE)

# Optional: Clean up the chunk files directory
# unlink(corrected_chunks_dir, recursive = TRUE)

```



# testing: ignore below here

```{r}




# to convert using gazetteer as reference table and jw method
test = head(unknown_destinations,5)
mapply(correct_city, test$CITY_DEST_NAME, test$STATE_DEST_CD, MoreArgs = list(ref_table = gazetteer, max_dist = .2, ref_city_col = "NAME", ref_state_col = "USPS"))

# to convert using gazetteer as reference table and osa method
test = head(unknown_destinations,5)
mapply(correct_city, test$CITY_DEST_NAME, test$STATE_DEST_CD, MoreArgs = list(ref_table = gazetteer, max_dist = 3, method = 'osa', ref_city_col = "NAME", ref_state_col = "USPS"))


# to convert using geonames as reference table and jw method

mapply(correct_city, test$CITY_DEST_NAME, test$STATE_DEST_CD, MoreArgs = list(ref_table = geonames, max_dist = .1, method = 'jw', ref_city_col = "V3", ref_state_col = "state"))

# to convert using geonames as reference table and osa method

mapply(correct_city, test$CITY_DEST_NAME, test$STATE_DEST_CD, MoreArgs = list(ref_table = geonames, max_dist = 3, method = 'osa', ref_city_col = "V3", ref_state_col = "state"))

# Apply correction
# Note: max_dist for OSA represents the number of character edits allowed (insertions, deletions, substitutions)
# For 'DAN YSIDRO' -> 'San Ysidro', only 1 substitution (D->S) is needed
# max_dist = 2 allows up to 2 character edits, which should catch most typos
unknown_destinations$corrected_city <- mapply(correct_city, unknown_destinations$CITY_DEST_NAME, unknown_destinations$STATE_DEST_CD, MoreArgs = list(ref_table = gazetteer, max_dist = 3, method = 'osa', ref_city_col = "NAME", ref_state_col = "USPS"))

unknown_destinations$corrected_city_jw <- mapply(correct_city, unknown_destinations$CITY_DEST_NAME, unknown_destinations$STATE_DEST_CD, MoreArgs = list(ref_table = gazetteer, max_dist = .15, ref_city_col = "NAME", ref_state_col = "USPS"))


# # Geocode using corrected names
# unknown_destinations <- tidygeocoder::geocode(unknown_destinations, city = corrected_city, state = STATE_DEST_CD, method = "osm")
```

```{r checking geocoding}
# DAN YSIDRO is not being corrected to San Ysidro
# I want to check the distances each at a time
test_correction_func <- function(city, state, ref_table, max_dist) {
  # Handle NA inputs early
  if (is.na(city) || is.na(state)) {
    return(NA_character_)
  }
  
  # Normalize input
  state_lookup <- toupper(state)
  city_lookup <- tools::toTitleCase(tolower(city))
  
  # Filter candidates by state (using the normalized state_lookup variable)
  candidates <- ref_table %>%
    mutate(
      city_clean = tools::toTitleCase(tolower(NAME)),
      state_clean = toupper(USPS)
    ) %>%
    filter(state_clean == state_lookup)

    # Calculate string distances using Optimal String Alignment (OSA)
  # OSA is more forgiving for typos like substitutions (D -> S)
  dists <- stringdist::stringdist(city_lookup, candidates$city_clean, method = "osa")

 return(data.frame(city_name = candidates$city_clean,
                   distance = dists)
 )
}

test = "Dan Ysidro"
ref <- gazetteer |> filter(grepl("Ysidro|Hanford", NAME))

#It is making 4 comparisons...
stringdist::stringdist(test, ref$NAME, method = "jw")
print(ref$NAME)

#But only Hanford shows up as the minimum distance...
#but this is because the state is given as "CA"
test_correction_func("Dan Ysidro", "CA", ref_table = ref, max_dist = 1)

#If the state is given as "NM", it finds the 3 San Ysidros ...
test_correction_func("Dan Ysidro", "NM", ref_table = ref, max_dist = 1)

# And the reason is because the Gazetter US Places list does not include San Ysidro CA!; it is probably listed as San Diego.


#Why isn't OSA working well?
test = "Hidalqgo"
ref <- gazetteer |> filter(grepl("Ysidro|Hanford|Hidalgo", NAME))

#It is making 4 comparisons...
stringdist::stringdist(test, ref$NAME, method = "osa")
print(ref$NAME)
# Ah - The NAME column in gazetteer includes CDP, city, or village as an ending, making the strings really far apart. Adding in geonames data frame with just city and state may help.
```
