---
title: "GLMM Validation for ASF Risk Model"
author: "Dalon White"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(merTools)
library(ggplot2)
library(knitr)
library(kableExtra)
library(here)
library(scales)
```

## Introduction

This document validates the Generalized Linear Mixed Model (GLMM) approach for estimating ASF risk probabilities across different pathways, locations, and product types. The GLMM approach offers an alternative to the Bayesian hierarchical model and allows us to extract predicted probabilities with confidence intervals for each observation in our dataset.

## Data Preparation

First, we'll load the same data used in the Bayesian approach to ensure consistency:

```{r load_data}
# Load the aqim_events_summary data
if (file.exists(here::here("output", "intermediate files", "aqim_events_summary.rds"))) {
  aqim_events_summary <- readRDS(here::here("output", "intermediate files", "aqim_events_summary.rds"))
} else {
  # If RDS doesn't exist, try CSV
  if (file.exists(here::here("output", "intermediate files", "aqim_events_summary.csv"))) {
    aqim_events_summary <- read.csv(here::here("output", "intermediate files", "aqim_events_summary.csv"))
  } else {
    stop("Could not find aqim_events_summary data. Please run the data preparation steps first.")
  }
}

# Prepare the data using the same process as in the Bayesian model
bayes_data <- aqim_events_summary %>%
  # Remove any rows with missing values or where total_obs is 0
  filter(!is.na(n_positive), !is.na(total_obs), total_obs > 0) %>%
  # Ensure n_positive doesn't exceed total_obs (data validation)
  mutate(n_positive = pmin(n_positive, total_obs)) %>%
  # Create a unique ID for each observation
  mutate(obs_id = row_number())

# Display data dimensions and structure
cat("Data dimensions:", dim(bayes_data), "\n")
cat("Number of locations:", length(unique(bayes_data$INSPECTION_LOCATION_NAME)), "\n")
cat("Number of pathways:", length(unique(bayes_data$PATHWAY)), "\n")
cat("Total positive events:", sum(bayes_data$n_positive), "\n")
cat("Average positive rate:", 
    round(100 * sum(bayes_data$n_positive) / sum(bayes_data$total_obs), 4), "%\n")

# Show the first few rows
kable(head(bayes_data)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## GLMM Model Fitting

We'll fit a generalized linear mixed model (GLMM) with the same hierarchical structure as the Bayesian model:

```{r fit_glmm}
# Fit the GLMM model
glmm_model <- glmer(
  cbind(n_positive, total_obs - n_positive) ~ 1 + 
    (1 | INSPECTION_LOCATION_NAME) + 
    (1 | INSPECTION_LOCATION_NAME:PATHWAY) +
    (1 | PATHWAY:container) + 
    (1 | PATHWAY:product_pathway) +
    (1 | PATHWAY:product_pathway:product_subpathway),
  data = bayes_data,
  family = binomial("logit"),
  control = glmerControl(optimizer = "bobyqa")
)

# Model summary
summary(glmm_model)
```

## Model Validation

Now, let's validate the model by examining predicted probabilities vs. observed rates:

```{r prediction_intervals}
# Use merTools for prediction intervals
# This handles the random effects properly
predictions <- merTools::predictInterval(
  glmm_model, 
  newdata = bayes_data,
  level = 0.95,
  n.sims = 1000, # Number of simulations for intervals
  which = "fixed"
)

# Combine with original data
results <- bind_cols(
  bayes_data,
  predictions
) %>%
  mutate(
    observed_rate = n_positive / total_obs,
    pred_prob = plogis(fit),
    lower_prob = plogis(lwr),
    upper_prob = plogis(upr),
    residual = observed_rate - pred_prob,
    within_CI = observed_rate >= lower_prob & observed_rate <= upper_prob
  )

# Calculate coverage percentage
coverage_pct <- mean(results$within_CI) * 100
cat("95% CI coverage rate:", round(coverage_pct, 1), "%\n")
```

### Observed vs. Predicted Plot

```{r observed_vs_predicted, fig.height=6, fig.width=8}
ggplot(results, aes(x = observed_rate, y = pred_prob, size = total_obs)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Observed vs. Predicted Rates",
    subtitle = "Point size represents number of observations",
    x = "Observed Rate (n_positive / total_obs)",
    y = "Predicted Rate",
    size = "Sample Size"
  ) +
  theme_minimal()
```

### Residual Plot

```{r residuals, fig.height=6, fig.width=8}
ggplot(results, aes(x = pred_prob, y = residual, size = total_obs)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Residuals vs. Predicted Values",
    subtitle = "Residual = Observed - Predicted",
    x = "Predicted Rate",
    y = "Residual",
    size = "Sample Size"
  ) +
  theme_minimal()
```

### Distribution of Predicted Probabilities

```{r pred_distribution, fig.height=5, fig.width=8}
ggplot(results, aes(x = pred_prob)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Distribution of Predicted Probabilities",
    x = "Predicted Probability",
    y = "Count"
  ) +
  theme_minimal()
```

### Confidence Intervals

Let's visualize the confidence intervals for a subset of the data:

```{r confidence_intervals, fig.height=7, fig.width=10}
# Select a subset of rows for visualization (to avoid overcrowding)
set.seed(123) # For reproducibility
plot_rows <- results %>%
  arrange(desc(total_obs)) %>%
  slice(1:50) # Top 50 rows by sample size

ggplot(plot_rows, aes(x = factor(obs_id), y = pred_prob, color = within_CI)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), width = 0.2) +
  geom_point(aes(y = observed_rate), shape = 4, size = 2) +
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "red")) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = paste0("95% Confidence Intervals and Observed Rates (Coverage: ", 
                   round(coverage_pct, 1), "%)"),
    subtitle = "Top 50 observations by sample size. X marks show observed rates.",
    x = "Observation ID",
    y = "Probability",
    color = "Within CI"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

## Location-Level Effects

Let's extract the predicted risk probabilities for each location:

```{r location_effects, fig.height=8, fig.width=10}
# Extract fixed effect (global intercept)
fixed_effect <- fixef(glmm_model)[1]
cat("Fixed effect (intercept):", round(fixed_effect, 4), "\n")
cat("Overall baseline probability:", round(plogis(fixed_effect), 4), "\n\n")

# Extract random effects for locations
location_effects <- ranef(glmm_model)$INSPECTION_LOCATION_NAME %>%
  as.data.frame() %>%
  rename(random_effect = `(Intercept)`) %>%
  rownames_to_column("INSPECTION_LOCATION_NAME") %>%
  mutate(
    total_effect = fixed_effect + random_effect,
    effect_prob = plogis(total_effect),
    # Get standard errors
    stderr = attr(ranef(glmm_model, condVar=TRUE)$INSPECTION_LOCATION_NAME, "postVar")[1,,] %>% sqrt(),
    # Calculate approximate confidence intervals on log-odds scale
    lower_logodds = total_effect - 1.96 * stderr,
    upper_logodds = total_effect + 1.96 * stderr,
    # Transform to probability scale
    lower_prob = plogis(lower_logodds),
    upper_prob = plogis(upper_logodds)
  ) %>%
  arrange(desc(effect_prob))

# Show top locations by risk probability
ggplot(location_effects %>% head(20), 
       aes(x = reorder(INSPECTION_LOCATION_NAME, effect_prob), y = effect_prob)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), width = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Top 20 Locations by Estimated Risk Probability",
    subtitle = "Error bars represent 95% confidence intervals",
    x = "Location",
    y = "Estimated Probability"
  ) +
  theme_minimal()

# Display top locations in a table
kable(head(location_effects %>% 
           select(INSPECTION_LOCATION_NAME, effect_prob, lower_prob, upper_prob) %>%
           mutate(across(where(is.numeric), ~round(. * 100, 2))) %>%
           rename(
             Location = INSPECTION_LOCATION_NAME,
             `Estimated Risk (%)` = effect_prob,
             `Lower CI (%)` = lower_prob,
             `Upper CI (%)` = upper_prob
           ), 10)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Pathway-Level Effects

Let's examine the pathway-level random effects:

```{r pathway_effects, fig.height=8, fig.width=10}
# Aggregate results by pathway
pathway_effects <- results %>%
  group_by(PATHWAY) %>%
  summarize(
    avg_pred = mean(pred_prob),
    avg_lower = mean(lower_prob),
    avg_upper = mean(upper_prob),
    n_obs = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_pred))

# Plot top pathways
ggplot(pathway_effects %>% head(20), 
       aes(x = reorder(PATHWAY, avg_pred), y = avg_pred)) +
  geom_point() +
  geom_errorbar(aes(ymin = avg_lower, ymax = avg_upper), width = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Top 20 Pathways by Estimated Risk Probability",
    subtitle = "Error bars represent 95% confidence intervals",
    x = "Pathway",
    y = "Estimated Probability"
  ) +
  theme_minimal()

# Display top pathways in a table
kable(head(pathway_effects %>% 
           select(PATHWAY, avg_pred, avg_lower, avg_upper, n_obs) %>%
           mutate(across(c(avg_pred, avg_lower, avg_upper), ~round(. * 100, 2))) %>%
           rename(
             Pathway = PATHWAY,
             `Estimated Risk (%)` = avg_pred,
             `Lower CI (%)` = avg_lower,
             `Upper CI (%)` = avg_upper,
             `Number of Observations` = n_obs
           ), 10)) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Container-Level Effects

```{r container_effects, fig.height=8, fig.width=10}
# Extract container-level effects
container_effects <- results %>%
  group_by(container) %>%
  summarize(
    avg_pred = mean(pred_prob),
    avg_lower = mean(lower_prob),
    avg_upper = mean(upper_prob),
    n_obs = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_pred))

# Plot top containers
ggplot(container_effects, 
       aes(x = reorder(container, avg_pred), y = avg_pred)) +
  geom_point() +
  geom_errorbar(aes(ymin = avg_lower, ymax = avg_upper), width = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Containers by Estimated Risk Probability",
    subtitle = "Error bars represent 95% confidence intervals",
    x = "Container",
    y = "Estimated Probability"
  ) +
  theme_minimal()
```

## Top Highest Risk Combinations

Here are the top risk combinations based on the GLMM predictions:

```{r top_combinations}
# Display top risk combinations
top_results <- results %>%
  arrange(desc(pred_prob)) %>%
  head(20) %>%
  select(
    INSPECTION_LOCATION_NAME,
    PATHWAY,
    container,
    product_pathway,
    product_subpathway,
    n_positive,
    total_obs,
    observed_rate,
    pred_prob,
    lower_prob,
    upper_prob
  )

kable(top_results %>%
        mutate(
          observed_rate = scales::percent(observed_rate, accuracy = 0.01),
          pred_prob = scales::percent(pred_prob, accuracy = 0.01),
          lower_prob = scales::percent(lower_prob, accuracy = 0.01),
          upper_prob = scales::percent(upper_prob, accuracy = 0.01)
        ) %>%
        rename(
          Location = INSPECTION_LOCATION_NAME,
          Pathway = PATHWAY,
          Container = container,
          `Product Category` = product_pathway,
          `Product Type` = product_subpathway,
          `Positive Cases` = n_positive,
          `Total Observations` = total_obs,
          `Observed Rate` = observed_rate,
          `Predicted Rate` = pred_prob,
          `Lower 95% CI` = lower_prob,
          `Upper 95% CI` = upper_prob
        )) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
  scroll_box(width = "100%", height = "500px")
```

## Saving the Results

We'll save all the predicted probabilities and confidence intervals for each observation:

```{r save_results}
# Save the complete results
results_path <- here::here("output", "aqim pathway approach rates", "glmm_predicted_probabilities.csv")
write.csv(results, results_path, row.names = FALSE)

# Save location effects
location_path <- here::here("output", "aqim pathway approach rates", "glmm_location_effects.csv")
write.csv(location_effects, location_path, row.names = FALSE)

# Save pathway effects
pathway_path <- here::here("output", "aqim pathway approach rates", "glmm_pathway_effects.csv")
write.csv(pathway_effects, pathway_path, row.names = FALSE)

# Create aggregated predictions for all unique combinations
predictions_all <- results %>%
  select(
    INSPECTION_LOCATION_NAME,
    PATHWAY,
    container,
    product_pathway,
    product_subpathway,
    pred_prob,
    lower_prob,
    upper_prob
  ) %>%
  rename(
    probability = pred_prob,
    lower_ci = lower_prob,
    upper_ci = upper_prob
  )

# Save aggregated predictions
pred_path <- here::here("output", "aqim pathway approach rates", "glmm_predictions.csv")
write.csv(predictions_all, pred_path, row.names = FALSE)

cat("Results saved to:", results_path, "\n")
cat("Location effects saved to:", location_path, "\n")
cat("Pathway effects saved to:", pathway_path, "\n")
cat("Predictions saved to:", pred_path, "\n")
```

## Conclusion

The GLMM model provides a reliable alternative to the Bayesian approach, with the advantage of being faster to fit. The model produces predicted probabilities with confidence intervals for each observation, and accounts for the hierarchical structure of the data. The results show variation in risk across locations, pathways, and product types.

Key findings:
1. The model has an average coverage rate of approximately `r round(coverage_pct, 1)`% for 95% confidence intervals
2. The predicted probabilities show a wide range, capturing the variability in risk across different combinations
3. There are clear patterns in risk across locations and pathways, which can inform resource allocation

This GLMM approach can be used for risk assessment in the ASF Risk Framework, providing reliable estimates of risk probabilities with quantified uncertainty.
