---
title: "ASF Risk Framework Parameters"
author: "Dalon White"
date: "2025-09-03"
output: html_document
---

<!--
This script defines all parameters used across the ASF risk framework.
It serves as the master configuration file for the entire pipeline.
This is the SINGLE SOURCE OF TRUTH for all parameter values.
-->

```{r setup, include=FALSE}
library(here) # For project-relative paths

# Create output directory if it doesn't exist
dir.create(here::here("output", "intermediate files"), showWarnings = FALSE, recursive = TRUE)
```

# Master Parameters for ASF Risk Framework

This file contains all the parameters used by the ASF risk framework. It is the single source of truth for these values. When you want to change a parameter value, you should modify it here.

The parameters defined here are used by all the scripts in the ASF risk framework. They can be run with these parameters either:

1. By running the scripts directly, which will load these parameters if available
2. By using the `run_pipeline.R` script to run scripts in sequence

## Parameter Definitions

```{r define-params}
# Create the master params list that will be used by all scripts
asf_params <- list(
  # Date ranges
  begin_date = "2021-01-01",
  end_date = "2024-12-31",
  wads_volume_begin_date = "2023-01-01",
  wads_volume_end_date = "2024-12-31",
  
  # Affected countries settings
  affected_countries = "current", # "current" or "mex_can" or "other"
  other_affected_countries = c("Country1 ISO-3 Code", "Country2 ISO-3 Code"),
  years = 2023:2024, # Years to include for ASF affected countries
  affected_country_file = "asf affected countries.csv",
  
  # Product settings
  affected_product_file = "asf affected product name reference table.csv",
  quantity_units_to_kg = "quantity units to kilograms reference table.csv",
  pathway_volume_smuggling_reference_table_filename = "product pathway by volume reference table",
  risk_quantification = "0_1", # "0_1" or "range"
  
  # WADS volume settings
  wads_volume_reference_filename = "wads volume reference table.csv",
  wads_volume_filename = "wads_passenger_volume_by_port_pathway_month_year",
  wads_ECO_volume = (mean(1066000000+1400000000)/12), # Total express courier volume to supplement WADS data if needed (2023 + 2024); this is applied to monthly WADS data, so it needs to be divided by 12 so that it does not overestimate how many come in per month

  # Output file names
  glm_results_filename = "glmm_results_aqim fraction at risk by pathway and location",
  bootstrap_results_filename = "bootstrap_results_aqim fraction at risk by pathway and location",
  group_size_filename = "aqim_mean_pathway_group_size",
  aqim_qmi_size_distribution_filename = "pathway_weight_distribution_models",
  pathway_volumes_filename = "volume_of_risk_by_pathway",
  ean_geolocated_filename = "ean_records_geocoded",
  ibl_geolocated_filename = "ibl_records_geocoded",

  # Simulation settings
  # Default is 100 for production, can be reduced for testing
  n_sims_for_volume = 100,
  
  # Spatial settings
  # Default is 30000 for production, can be increased for testing
  # 30000 is used for high-resolution final output
  US_grid_cell_size = 60000
)

# Save the params object for use by other scripts
saveRDS(asf_params, here::here("output", "intermediate files", "asf_risk_params.rds"))
```

# Function to Load Parameters

```{r load-params-function}
# This function will be used by all scripts to load parameters
# It allows scripts to be run independently or as part of the pipeline
load_asf_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    # Load the saved params
    params <- readRDS(params_file)
    message("Loaded parameters from shared configuration file")
  } else {
    # Create default params if the file doesn't exist
    message("Shared parameter file not found.")
  }
  
  return(params)
}
```

# Source Files to Run Complete Pipeline

```{r run-complete-pipeline, eval=FALSE}
# This code chunk is not evaluated by default
# It shows how to run the full pipeline in sequence

# Define the list of scripts in order
pipeline_scripts <- c(
  "01_01 aqim pathway passenger layers.Rmd",
  "01_02 aqim pathway data prep.Rmd",
  "01_03 aqim pathway approach rates.Rmd",
  "01_04 aqim volume distribution.Rmd",
  "01_05 aqim pathway passenger volumes.Rmd",
  "01_06 aqim pathway risk_PoE.Rmd",
  "01_07 aqim volume spatial layers.Rmd",
  "03_01 ean data.Rmd",
  "03_02 ean geolocation.Rmd",
  "03_03 ean port and destination layers.Rmd",
  "04_01 IBL data.Rmd",
  "04_02 IBL port and destination layers.Rmd",
  "04_03 IBL report output.R"
)

# You can run the pipeline using the run_pipeline.R script:
# source(here::here("scripts", "run_pipeline.R"))
# run_pipeline()
```

# Parameter Modification Guide

If you need to modify parameter values for different runs or scenarios, you have several options:

1. Edit this file and re-run `initialize_params.R` to update the shared parameter file
2. For temporary changes, modify the parameters in memory:
   ```r
   # Load existing parameters
   params <- readRDS(here::here("output", "intermediate files", "asf_risk_params.rds"))
   
   # Modify a parameter
   params$n_sims_for_volume <- 10  # Reduce for faster testing
   
   # Run with modified parameters
   run_pipeline(params)
   ```
3. Create different parameter files for different scenarios:
   ```r
   # Load existing parameters
   params <- readRDS(here::here("output", "intermediate files", "asf_risk_params.rds"))
   
   # Modify parameters for a specific scenario
   params$n_sims_for_volume <- 500
   params$US_grid_cell_size <- 10000
   
   # Save with a scenario name
   saveRDS(params, here::here("output", "intermediate files", "asf_risk_params_high_resolution.rds"))
   ```
  "01_02 aqim pathway data prep.Rmd",
  "01_03 aqim pathway approach rates.Rmd",
  "01_04 aqim volume distribution.Rmd",
  "01_05 aqim pathway passenger volumes.Rmd",
  "01_06 aqim pathway risk_PoE.Rmd",
  "01_07 aqim volume spatial layers.Rmd"
)

# Function to render an Rmd file with params
render_with_params <- function(script_path, params) {
  require(rmarkdown)
  rmarkdown::render(
    here::here("scripts", script_path),
    params = params,
    output_dir = here::here("output", "rendered_reports")
  )
}

# Run the full pipeline
for (script in pipeline_scripts) {
  message(paste("Running:", script))
  render_with_params(script, asf_params)
}
```

# Running Individual Scripts

Individual scripts can be run in two ways:

1. By sourcing this file first, then running the script (which will use the shared params)
2. By running the script directly (which will use its built-in default params)

The scripts have been modified to check for the existence of the shared params file
and use it if available, or fall back to their built-in defaults.
