---
title: "aqim pathway approach rates"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  risk_quantification: "0_1" # 0_1 or "range"
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  glm_results_filename: "glmm_results_aqim fraction at risk by pathway and location"
  bootstrap_results_filename: "bootstrap_results_aqim fraction at risk by pathway and location"
  group_size_filename: "aqim_mean_pathway_group_size"
---

-   prepped data already has the pathway, entry type, etc., and risk as no/low/med/high
-   need to use country of origin and product origin to determine if the inspection has any asf risk
-   then model the kg of product with asf risk by pathway, entry type, location, country of origin, product pathway, product subpathway

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(lme4)
library(ggplot2)
library(broom)
library(multcomp)
library(emmeans)
library(sjPlot)
library(broom.mixed)
library(boot)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      risk_quantification = "0_1", # 0_1 or "range"
      affected_countries = "current", # "current" or "mex_can" or "other" 
      years = 2024:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"), 
      glm_results_filename = "glmm_results_aqim fraction at risk by pathway and location",
      bootstrap_results_filename = "bootstrap_results_aqim fraction at risk by pathway and location",
      group_size_filename = "aqim_mean_pathway_group_size"
    )
  }
}
```

# Call AQIM data

```{r get data}

aqim_dat <- read.csv("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv")

```

# Set affected countries for analysis

Within params, can choose whether to use the current affected countries or just Mexico and Canada. The current ones, are then filtered by the years specified in params$years.
If only interested in the last year, then set params$years to the last year of interest. It is important to remember that we are not predicting where surveillance needs to be, such as for the CEAH group, but identifying the risk of ASF in products that are coming into the US, so whether an item corresponds to a year and country is not important. Instead, we are focused on estimating how much comes from each country and whether that country has ASF risk.

```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
if(params$affected_countries == "current") {
  affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
} else if (params$affected_countries == "mex_can") {
  affected_countries <- c("MEX", "CAN")
} else if (params$affected_countries == "other") {
  affected_countries <- params$other_affected_countries
} else {
  stop("Invalid value for params$affected_countries")
}
```


# Assign risk by country of origin and product origin

```{r assign risk}
# Filter for affected countries
aqim_dat <- aqim_dat %>% mutate(IS_ASF_RISK = case_when(
  edge_origin %in% affected_countries | product_origin %in% affected_countries ~ IS_ASF_RISK,
  .default = "NO"
         )
)
#once it is filtered by country, this is the at risk products by weight
aqim_dat |> filter(PATHWAY== "Pedestrian SB") |> count(TOTAL_KG > 0, IS_ASF_RISK)
#at risk product types exist, but they did not come from countries with ASF risk

```

# trouble shooting
Express courier is coming out too high in the final data layer - like 25% of packages are smuggled commercial prodcut -, and has entries from 'baggage' - there shouldn't be any baggage conveyances for ECO
and san francisco has 0.0311 smuggled commercial product AND 0.0311 swine products in passenger baggage; they shouldn't be the same necessarily so this might indicate a problem
- I fixed the baggage issue by correcting a reference file, it now says 'package'. Same with mail
- I have determined that the rate of at risk baggage in SF Int'l really was 328 out of ~10k (3%)
- I corrected an error in the reference table that wasnt assigning a value to TOTAL_KG for most AQAS aqim data because the UNIT name didn't match (mostly just missing 's')
- Need to check that this works now
```{r troubleshooting A}
# #Looking down at combinations_summary I can see that ECO does have some at-risk items but it's only a 51 out of 63k items
# #There is only one smuggled commercial product in an ECO package
# aqim_dat |> filter(PATHWAY== "Express_courier") |> count(TOTAL_KG > 0, IS_ASF_RISK)
# aqim_dat |> filter(PATHWAY== "Express_courier" & IS_ASF_RISK > 0)
# View(aqim_dat |> filter(PATHWAY== "Express_courier" & IS_ASF_RISK > 0))
# #Its a personal use item that has been appropraitely assigned to smuggling because it is 26 kilos

# # look at san francisco airport passenger baggage data
# SF_bags <- aqim_dat |> filter(PATHWAY == "Airport - Aircraft - Person - CBP" & grepl("SAN FRAN", INSPECTION_LOCATION_NAME)) |> count(TOTAL_KG > 0, IS_ASF_RISK)
# print(SF_bags)
# #at risk entries in SF Int'l Airport
# print(paste("at risk entries in SF Int'l Airport is", round(
# SF_bags[SF_bags$`TOTAL_KG > 0` == TRUE & SF_bags$IS_ASF_RISK == 1,]$n/sum(SF_bags$n)* 100, 2), "percent"
# ))


```

## handle risk quantification based on 0-1 or a scale

```{r handle risk quantification}
if(params$risk_quantification == "0_1") {
  # Convert risk to binary (0 or 1)
  aqim_dat <- aqim_dat %>%
    mutate(IS_ASF_RISK = case_when(
      IS_ASF_RISK == "HIGH" ~ 1,
      IS_ASF_RISK == "MEDIUM" ~ 1,
      IS_ASF_RISK == "LOW" ~ 1,
      .default = 0
    ))
} else if (params$risk_quantification == "range") {
  # Keep risk as is (no conversion)
  aqim_dat <- aqim_dat %>%
    mutate(IS_ASF_RISK = case_when(
      IS_ASF_RISK == "HIGH" ~ 1,
      IS_ASF_RISK == "MEDIUM" ~ .67,
      IS_ASF_RISK == "LOW" ~ .33,
      .default = 0
    ))
} else {
  stop("Invalid value for params$risk_quantification")
}

# modify TOTAL_KG by IS_ASF_RISK
aqim_dat <- aqim_dat |> mutate(TOTAL_KG = TOTAL_KG * IS_ASF_RISK)

# # modify IS_ASF_RISK to be binary
# aqim_dat <- aqim_dat |> mutate(IS_ASF_RISK = ifelse(IS_ASF_RISK > 0, 1, 0))

```

## Create Summary
This is to make sure I remember that not all these pathways have risk, so not all of them will be used in the 01_06 aqim pathway risk_PoE.Rmd script

```{r}
# Create summary statistics for combinations and risk counts
# Count all combinations by risk status
combinations_summary <- aqim_dat |>
  dplyr::count(PATHWAY, container, product_pathway, product_subpathway, IS_ASF_RISK)

# Get key statistics
pathway_counts <- combinations_summary |> distinct(PATHWAY) |> nrow()
total_combinations <- combinations_summary |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()
risk_combinations <- combinations_summary |> filter(IS_ASF_RISK > 0) |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()
no_risk_combinations <- combinations_summary |> filter(IS_ASF_RISK == 0) |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()

# Count observations
total_risk_count <- combinations_summary |> filter(IS_ASF_RISK > 0) |> summarize(total = sum(n)) |> pull(total)
total_no_risk_count <- combinations_summary |> filter(IS_ASF_RISK == 0) |> summarize(total = sum(n)) |> pull(total)
grand_total_count <- sum(combinations_summary$n)

# Create pathway-level summary
pathway_summary <- combinations_summary |>
  group_by(PATHWAY, IS_ASF_RISK) |>
  summarize(count = sum(n), .groups = "drop") |>
  pivot_wider(names_from = IS_ASF_RISK, values_from = count, 
              names_prefix = "risk_", values_fill = 0) |>
  mutate(total = risk_0 + risk_1,
         risk_percent = round(100 * risk_1 / total, 1))

# Print summary information
cat("\n### Summary of AQIM Data Combinations ###\n\n")
cat(paste("Total unique pathways:", pathway_counts, "\n"))
cat(paste("Total unique PATHWAY-container-product combinations:", total_combinations, "\n"))
cat(paste("- Combinations with ASF risk:", risk_combinations, 
          sprintf("(%.1f%%)", 100 * risk_combinations/total_combinations), "\n"))
cat(paste("- Combinations without ASF risk:", no_risk_combinations, "\n\n"))

cat("### Observation Counts ###\n\n")
cat(paste("Total observations:", format(grand_total_count, big.mark=","), "\n"))
cat(paste("- Observations with ASF risk:", format(total_risk_count, big.mark=","), 
          sprintf("(%.1f%%)", 100 * total_risk_count/grand_total_count), "\n"))
cat(paste("- Observations without ASF risk:", format(total_no_risk_count, big.mark=","), "\n\n"))

cat("###  Pathways by Risk Percentage ###\n\n")
top_risk_pathways <- pathway_summary |>
  filter(risk_1 > 0) |>
  arrange(desc(risk_percent)) 

print(top_risk_pathways |> dplyr::select(PATHWAY, risk_1, total, risk_percent))
print("The pathway combinations that exist are in the following table:")
View(combinations_summary)

```

# Need to predict both the incident rate per entry, and the kg of product that is intercepted.

I want to do all this on a per-pathway basis. For that, I need to first split the data by pathway into a list, and work from there

```{r split data by pathway}
data_for_risk <- aqim_dat |> 
  dplyr::select(CALENDAR_YEAR, INSPECTION_LOCATION_NAME, PATHWAY, group_size, IS_ASF_RISK, TOTAL_KG, container, product_pathway, product_subpathway)
aqim_dat_list <- split(data_for_risk, data_for_risk$PATHWAY)
aqim_dat_list <- lapply(aqim_dat_list, function(df) {
  df$group <- paste(df$PATHWAY, df$container, df$product_pathway, df$product_subpathway, sep = ";")
  return(df)
})

# Create a container for results
group_models <- list()

# Process each pathway separately
for (pathway_name in names(aqim_dat_list)) {
  pathway_data <- aqim_dat_list[[pathway_name]]
  
  # Get unique groups that aren't absences
  unique_groups <- pathway_data %>%
    filter(!endsWith(group, "NA;NA;NA")) %>%
    pull(group) %>%
    unique()
  
  # Create a sublist for this pathway's results
  group_models[[pathway_name]] <- list()
  
  # For each unique group in this pathway
  for (group_name in unique_groups) {
    cat(sprintf("\nAnalyzing group: %s\n", group_name))
    
    # Create binary outcome for this specific group
    modeling_data <- pathway_data %>%
      mutate(is_target_group = ifelse(group == group_name, 1, 0))
    
    # Check if we have enough positive examples
    positive_count <- sum(modeling_data$is_target_group)
    if (positive_count < 1) {
      cat("  Skipping due to insufficient positive examples (", positive_count, ")\n")
      next
    }
    
    # Store container for this group
    group_models[[pathway_name]][[group_name]] <- list()
      # 1. BOOTSTRAP APPROACH - simpler and more robust
    tryCatch({
      # Bootstrap function for this specific group
      boot_group_function <- function(data, indices) {
        d <- data[indices,]
        mean(d$IS_ASF_RISK)
      }
      
      # Run bootstrap
      boot_result <- boot(modeling_data, 
                          statistic = boot_group_function, 
                          R = 1000)
      
      # Calculate confidence intervals
      ci <- boot.ci(boot_result, type = "perc")
      
      # Store results
      group_models[[pathway_name]][[group_name]][["bootstrap"]] <- list(
        positive_count = positive_count,
        total_count = nrow(modeling_data),
        proportion = boot_result$t0,
        ci_lower = ci$percent[4],
        ci_upper = ci$percent[5]
      )
      
      cat(sprintf("  Bootstrap results: %.4f (%.4f, %.4f)\n", 
                  boot_result$t0, ci$percent[4], ci$percent[5]))
      
      # Now calculate bootstrap for each location within this group
      cat("  Analyzing locations within this group:\n")
      
      # Get all locations for this group
      group_specific_data <- modeling_data[modeling_data$is_target_group == 1, ]
      locations <- unique(group_specific_data$INSPECTION_LOCATION_NAME)
      
      # Create a container for location-specific bootstrap results
      location_bootstrap_results <- list()
      
      # Process each location
      for (loc_name in locations) {
        cat(sprintf("    Location: %s\n", loc_name))
        
        # Run bootstrap for this location
        location_result <- location_bootstrap(pathway_name, group_name, loc_name)
        
        # Store results if available
        if (!is.null(location_result)) {
          location_bootstrap_results[[loc_name]] <- location_result
          
          # Print summary
          cat(sprintf("      %s results: %.4f (%.4f, %.4f) from %d/%d observations\n", 
                      location_result$method,
                      location_result$proportion, 
                      location_result$ci_lower, 
                      location_result$ci_upper,
                      location_result$positive_count,
                      location_result$total_count))
        }
      }
      
      # Store the location-specific bootstrap results
      group_models[[pathway_name]][[group_name]][["location_bootstrap"]] <- location_bootstrap_results
      
    }, error = function(e) {
      cat("  Bootstrap analysis failed:", e$message, "\n")
    })
    
    # 2. GLMM APPROACH - for more complex modeling
    tryCatch({
      # Instead of building a complex model, we'll directly calculate the empirical probabilities
      # We'll still store the data in the same structure for compatibility
      
      # Get data for this specific group only
      group_specific_data <- modeling_data[modeling_data$is_target_group == 1, ]
      
      # Calculate rates by location
      location_rates <- group_specific_data %>%
        group_by(INSPECTION_LOCATION_NAME) %>%
        summarize(
          location_total = n(),
          location_positive = sum(IS_ASF_RISK, na.rm = TRUE),
          location_rate = location_positive / location_total,
          .groups = "drop"
        )
      
      # Log info about the data
      model_data_size <- nrow(group_specific_data)
      positive_risk_count <- sum(group_specific_data$IS_ASF_RISK, na.rm = TRUE)
      
      cat(sprintf("  Data for group model: %d observations (%d with risk)\n", 
                model_data_size, positive_risk_count))
      
      # Store a placeholder "model" with the data needed for later analysis
      # This ensures compatibility with the existing code structure
      placeholder_model <- list(
        data = group_specific_data,
        location_rates = location_rates,
        total_observations = model_data_size,
        positive_observations = positive_risk_count
      )
      class(placeholder_model) <- "empirical_rates"
      
      # Store this placeholder model
      group_models[[pathway_name]][[group_name]][["glmm"]] <- placeholder_model
      
      cat("  Direct empirical rate calculation successful\n")
      
    }, error = function(e) {
      cat("  Rate calculation failed:", e$message, "\n")
    })
  }
}

# Function to perform bootstrap for location-specific probabilities
location_bootstrap <- function(pathway_name, group_name, location_name) {
  # Get data for this specific location in this pathway and group
  data <- aqim_dat_list[[pathway_name]] %>%
    filter(group == group_name,
           INSPECTION_LOCATION_NAME == location_name)
  
  # Check if we have enough data
  if (nrow(data) < 10) {
    cat(sprintf("  Skip bootstrap for location %s: insufficient data (%d observations)\n", 
                location_name, nrow(data)))
    return(NULL)
  }
  
  # Count positives
  positive_count <- sum(data$IS_ASF_RISK, na.rm = TRUE)
  
  # If zero or all observations are positive, bootstrap may not be informative
  if (positive_count == 0 || positive_count == nrow(data)) {
    cat(sprintf("  Skip bootstrap for location %s: %d/%d positive (too homogeneous)\n", 
                location_name, positive_count, nrow(data)))
    
    # Return direct calculation instead
    p_hat <- positive_count / nrow(data)
    
    # Wilson score interval for the confidence bounds
    z_value <- 1.96  # 95% CI
    n <- nrow(data)
    
    denominator <- 1 + (z_value^2 / n)
    center <- (p_hat + (z_value^2 / (2 * n))) / denominator
    halfwidth <- z_value * sqrt((p_hat * (1 - p_hat) + (z_value^2 / (4 * n))) / n) / denominator
    
    ci_lower <- max(0, center - halfwidth)
    ci_upper <- min(1, center + halfwidth)
    
    if (positive_count == 0) {
      # For zero counts, use rule of three for upper bound
      ci_upper <- min(1, 3/n)
    }
    
    return(list(
      proportion = p_hat,
      ci_lower = ci_lower,
      ci_upper = ci_upper,
      total_count = nrow(data),
      positive_count = positive_count,
      method = "wilson_score"  # Indicate we used Wilson score instead of bootstrap
    ))
  }
  
  # Define bootstrap function
  boot_func <- function(data, indices) {
    d <- data[indices,]
    mean(d$IS_ASF_RISK, na.rm = TRUE)
  }
  
  # Run bootstrap
  tryCatch({
    boot_result <- boot(data, statistic = boot_func, R = 1000)
    
    # Calculate confidence intervals
    ci <- boot.ci(boot_result, type = "perc")
    
    return(list(
      proportion = boot_result$t0,
      ci_lower = ci$percent[4],
      ci_upper = ci$percent[5],
      total_count = nrow(data),
      positive_count = positive_count,
      method = "bootstrap"  # Indicate we used bootstrap
    ))
  }, error = function(e) {
    cat(sprintf("  Bootstrap failed for location %s: %s\n", location_name, e$message))
    
    # Fallback to direct calculation with Wilson score interval
    p_hat <- positive_count / nrow(data)
    
    # Wilson score interval
    z_value <- 1.96  # 95% CI
    n <- nrow(data)
    
    denominator <- 1 + (z_value^2 / n)
    center <- (p_hat + (z_value^2 / (2 * n))) / denominator
    halfwidth <- z_value * sqrt((p_hat * (1 - p_hat) + (z_value^2 / (4 * n))) / n) / denominator
    
    ci_lower <- max(0, center - halfwidth)
    ci_upper <- min(1, center + halfwidth)
    
    return(list(
      proportion = p_hat,
      ci_lower = ci_lower,
      ci_upper = ci_upper,
      total_count = nrow(data),
      positive_count = positive_count,
      method = "wilson_score_fallback"  # Indicate we used Wilson score as fallback
    ))
  })
}

# Get the name of the reference level and replace it in the location column
# Method 1: Extract the reference level directly from a model
get_reference_location <- function(model) {
  # Handle different model types
  if (inherits(model, "empirical_rates")) {
    # For empirical rates, there's no reference level concept
    # We'll use the data directly
    model_data <- model$data
    
    # Get all unique location names from the data
    all_locations <- unique(model_data$INSPECTION_LOCATION_NAME)
    
    # Get location names that appear in location_rates
    locations_in_rates <- model$location_rates$INSPECTION_LOCATION_NAME
    
    # The reference location is the one that doesn't appear in location_rates
    reference_location <- setdiff(all_locations, locations_in_rates)
    
    # If no reference location is found, return a placeholder
    if (length(reference_location) == 0) {
      return("N/A")
    }
    return(reference_location)
  } else {
    # Original GLM-based reference extraction
    # Get the data used in the model
    model_data <- model$model
    
    # Get all unique location names from the data
    all_locations <- unique(model_data$INSPECTION_LOCATION_NAME)
    
    # Get location names that appear in coefficients
    coef_names <- names(coef(model))
    location_coefs <- coef_names[grepl("INSPECTION_LOCATION_NAME", coef_names)]
    locations_in_coefs <- gsub("INSPECTION_LOCATION_NAME", "", location_coefs)
    
    # The reference location is the one that doesn't appear in coefficients
    reference_location <- setdiff(all_locations, locations_in_coefs)
    
    return(reference_location)
  }
}

# Apply to a specific model
for (pathway_name in names(group_models)) {
  for (group_name in names(group_models[[pathway_name]])) {
    if (!is.null(group_models[[pathway_name]][[group_name]][["glmm"]])) {
      model <- group_models[[pathway_name]][[group_name]][["glmm"]]
      ref_location <- get_reference_location(model)
      cat(sprintf("For pathway '%s', group '%s', the reference location is: %s\n", 
                 pathway_name, group_name, ref_location))
    }
  }
}


# Create a summary dataframe of all group results
group_summary <- tibble(
  pathway = character(),
  group = character(),
  positive_count = integer(),
  total_count = integer(),
  proportion = numeric(),
  ci_lower = numeric(),
  ci_upper = numeric()
)

# Extract bootstrap results for all groups
for (pathway_name in names(group_models)) {
  for (group_name in names(group_models[[pathway_name]])) {
    results <- group_models[[pathway_name]][[group_name]]
    
    if (!is.null(results[["bootstrap"]])) {
      group_summary <- group_summary %>%
        add_row(
          pathway = pathway_name,
          group = group_name,
          positive_count = results[["bootstrap"]]$positive_count,
          total_count = results[["bootstrap"]]$total_count,
          proportion = results[["bootstrap"]]$proportion,
          ci_lower = results[["bootstrap"]]$ci_lower,
          ci_upper = results[["bootstrap"]]$ci_upper
        )
    }
  }
}

get_location_specific_probabilities <- function(models_list, alpha = 0.1, conf_level = 0.95) {
  results_df <- data.frame()
  # Calculate z-value for confidence interval
  z_value <- qnorm(1 - (1 - conf_level)/2)  # 1.96 for 95% CI
  
  for (pathway_name in names(models_list)) {
    for (group_name in names(models_list[[pathway_name]])) {
      if (!is.null(models_list[[pathway_name]][[group_name]][["glmm"]])) {
        model <- models_list[[pathway_name]][[group_name]][["glmm"]]
        
        # Handle either the original GLM model or our new empirical_rates object
        if (inherits(model, "empirical_rates")) {
          # Use direct empirical calculations
          location_rates <- model$location_rates
          
          # Calculate overall probability for the group (used for the Average)
          overall_prob <- model$positive_observations / model$total_observations
          
          # Add each location to results
          for (i in 1:nrow(location_rates)) {
            loc <- location_rates[i, ]
            
            # Skip locations with too few observations
            if (loc$location_total < 5) next
            
            # Calculate Wilson score interval for confidence bounds
            # (better for small counts than normal approximation)
            if (loc$location_positive > 0 && loc$location_total > 0) {
              p_hat <- loc$location_positive / loc$location_total
              n <- loc$location_total
              
              # Wilson score interval
              denominator <- 1 + (z_value^2 / n)
              center <- (p_hat + (z_value^2 / (2 * n))) / denominator
              halfwidth <- z_value * sqrt((p_hat * (1 - p_hat) + (z_value^2 / (4 * n))) / n) / denominator
              
              ci_lower <- max(0, center - halfwidth)
              ci_upper <- min(1, center + halfwidth)
            } else {
              # For zero counts, use adjusted estimates
              n <- loc$location_total
              ci_lower <- 0
              ci_upper <- min(1, 3/n)  # Rule of three for upper bound with zero events
            }
            
            # Add to results
            results_df <- rbind(results_df, data.frame(
              pathway = pathway_name,
              group = group_name,
              location = loc$INSPECTION_LOCATION_NAME,
              estimate = log(loc$location_rate/(1-loc$location_rate)),  # Convert to log-odds for compatibility
              probability = loc$location_rate,
              ci_lower = ci_lower,
              ci_upper = ci_upper,
              p_value = NA,  # No p-value for empirical rates
              significant = NA,
              stringsAsFactors = FALSE
            ))
          }
          
          # Add the average (overall) rate for this group
          results_df <- rbind(results_df, data.frame(
            pathway = pathway_name,
            group = group_name,
            location = "Average",
            estimate = NA,
            probability = overall_prob,
            ci_lower = NA,  # Will be replaced by bootstrap CI later
            ci_upper = NA,
            p_value = NA,
            significant = NA,
            stringsAsFactors = FALSE
          ))
          
        } else {
          # Original GLM-based code - kept for compatibility
          # Get model summary and coefficients
          model_summary <- summary(model)
          coefs <- model_summary$coefficients
          
          # Get the intercept (baseline probability)
          intercept <- coefs["(Intercept)", "Estimate"]
          intercept_se <- coefs["(Intercept)", "Std. Error"]
          intercept_p <- coefs["(Intercept)", "Pr(>|z|)"]
          
          # Calculate CI for intercept on log-odds scale
          intercept_lower <- intercept - z_value * intercept_se
          intercept_upper <- intercept + z_value * intercept_se
          
          # Transform to probability scale
          baseline_prob <- plogis(intercept)
          ci_lower_prob <- plogis(intercept_lower)
          ci_upper_prob <- plogis(intercept_upper)
          
          # Add baseline/intercept to results
          results_df <- rbind(results_df, data.frame(
            pathway = pathway_name,
            group = group_name,
            location = "Baseline",
            estimate = intercept,
            probability = baseline_prob,
            ci_lower = ci_lower_prob,
            ci_upper = ci_upper_prob,
            p_value = intercept_p,
            significant = intercept_p < alpha,
            stringsAsFactors = FALSE
          ))
          
          # Add each location to results
          for (i in 1:nrow(coefs)) {
            coef_name <- rownames(coefs)[i]
            
            # Check if this is a location coefficient
            if (grepl("INSPECTION_LOCATION_NAME", coef_name) && coef_name != "(Intercept)") {
              # Extract the actual location name
              location_name <- gsub("INSPECTION_LOCATION_NAME", "", coef_name)
              
              # Get coefficient value, SE and p-value
              coef_value <- coefs[coef_name, "Estimate"]
              coef_se <- coefs[coef_name, "Std. Error"]
              p_value <- coefs[coef_name, "Pr(>|z|)"]
              
              # Calculate combined effect with intercept
              combined_estimate <- intercept + coef_value
              
              # Calculate SE for combined estimate (simplified - ignoring covariance)
              # For more precise CI, profile likelihood or bootstrap methods would be better
              combined_se <- sqrt(intercept_se^2 + coef_se^2)
              
              # Calculate CI for combined effect
              combined_lower <- combined_estimate - z_value * combined_se
              combined_upper <- combined_estimate + z_value * combined_se
              
              # Transform to probability scale
              location_prob <- plogis(combined_estimate)
              location_ci_lower <- plogis(combined_lower)
              location_ci_upper <- plogis(combined_upper)
              
              # Add to results
              results_df <- rbind(results_df, data.frame(
                pathway = pathway_name,
                group = group_name,
                location = location_name,
                estimate = coef_value,
                probability = location_prob,
                ci_lower = location_ci_lower,
                ci_upper = location_ci_upper,
                p_value = p_value,
                significant = p_value < alpha,
                stringsAsFactors = FALSE
              ))
            }
          }
          
          # Also include the average predicted probability across all observations
          avg_prob <- mean(fitted(model))
          
          # For average probability, we can use bootstrap to get CI
          # but for simplicity, we'll use NA for now
          results_df <- rbind(results_df, data.frame(
            pathway = pathway_name,
            group = group_name,
            location = "Average",
            estimate = NA,
            probability = avg_prob,
            ci_lower = NA,
            ci_upper = NA,
            p_value = NA,
            significant = NA,
            stringsAsFactors = FALSE
          ))
        }
      }
    }
  }
  
  return(results_df)
}

# Use the function
location_probabilities <- get_location_specific_probabilities(group_models)

# View the results with confidence intervals
head(location_probabilities)

# Get the bootstrap results for the average probabilities
extract_bootstrap_results <- function(models_list) {
  results_df <- data.frame()
  
  for (pathway_name in names(models_list)) {
    for (group_name in names(models_list[[pathway_name]])) {
      if (!is.null(models_list[[pathway_name]][[group_name]][["bootstrap"]])) {
        boot_results <- models_list[[pathway_name]][[group_name]][["bootstrap"]]
        
        results_df <- rbind(results_df, data.frame(
          pathway = pathway_name,
          group = group_name,
          proportion = boot_results$proportion,
          ci_lower = boot_results$ci_lower,
          ci_upper = boot_results$ci_upper
        ))
      }
    }
  }
  
  return(results_df)
}

bootstrap_results <- extract_bootstrap_results(group_models)
View(bootstrap_results)

# Replace the "Average" location probabilities with bootstrap results in cluding estimate and confidence intervals
location_probabilities_updated <- location_probabilities %>%
  left_join(
    bootstrap_results %>% 
      dplyr::select(pathway, group, bootstrap_proportion = proportion, 
             bootstrap_ci_lower = ci_lower, bootstrap_ci_upper = ci_upper),
    by = c("pathway", "group")
  ) %>%
  mutate(
    probability = ifelse(location == "Average" & !is.na(bootstrap_proportion), 
                       bootstrap_proportion, probability),
    ci_lower = ifelse(location == "Average" & !is.na(bootstrap_ci_lower), 
                     bootstrap_ci_lower, ci_lower),
    ci_upper = ifelse(location == "Average" & !is.na(bootstrap_ci_upper), 
                     bootstrap_ci_upper, ci_upper)
  ) %>%
  dplyr::select(-bootstrap_proportion, -bootstrap_ci_lower, -bootstrap_ci_upper)

# View the updated dataframe
print(location_probabilities_updated %>% filter(location == "Average"))
View(location_probabilities_updated)

# Create a dataframe of reference locations for each pathway and group
reference_locations <- data.frame()

# Extract all reference locations
for (pathway_name in names(group_models)) {
  for (group_name in names(group_models[[pathway_name]])) {
    if (!is.null(group_models[[pathway_name]][[group_name]][["glmm"]])) {
      model <- group_models[[pathway_name]][[group_name]][["glmm"]]
      ref_location <- get_reference_location(model)
      
      reference_locations <- rbind(reference_locations, data.frame(
        pathway = pathway_name,
        group = group_name,
        reference_location = ref_location
      ))
    }
  }
}

# Now update the location_probabilities_updated dataframe
location_probabilities_final <- location_probabilities_updated %>%
  left_join(reference_locations, by = c("pathway", "group")) %>%
  mutate(
    location = ifelse(location == "Baseline", reference_location, location)
  ) %>%
  dplyr::select(-reference_location)

# Check the updated dataframe
View(location_probabilities_final)

# Verify no more "Baseline" entries exist
if (any(location_probabilities_final$location == "Baseline")) {
  warning("Some Baseline entries could not be replaced")
}


```

## Validation Check: Check if the incidents found in vessel cargo are allowable at the time they were conducted

```{r}
#were the incidents found in vessel cargo allowable at the time they were conducted?
vessel_cargo_question <- aqim_dat |> filter(grepl("Maritime Port - Vessel - Cargo - CBP",PATHWAY)) |> filter(IS_ASF_RISK > 0) |> count(CALENDAR_YEAR,edge_origin,product_origin)

print(paste("these", nrow(vessel_cargo_question), "incidents were found in vessel cargo that were not allowed at the time they were conducted"))
head(vessel_cargo_question)
print("however, they were not seized, indicating that they were allowed to be imported with the correct paperwork, so these are not a risk for ASF")
print("In conclusion, no smuggling shows up in AQIM data of regulated cargo")


```

# save data

```{r save data}

setwd(here::here('output','aqim pathway approach rates')) 
# Save the final dataframe to a CSV file
write.csv(location_probabilities_final, paste(params$glm_results_filename,".csv"), row.names = FALSE)
saveRDS(location_probabilities_final, paste(params$glm_results_filename,".rds"))
write.csv(bootstrap_results, paste(params$bootstrap_results_filename,".csv"), row.names = FALSE)
saveRDS(bootstrap_results, paste(params$bootstrap_results_filename,".rds"))

```

# Save group size for later calculations
```{r save group size}
View(aqim_dat |> count(group_size, PATHWAY))
# If group_size is NA or 0, replace with 1
# for each pathway, find the average group size
group_size_summary <- aqim_dat %>%
  mutate(group_size = case_when(
    group_size == 0 ~ 1,
    is.na(group_size) ~ 1,
    .default = group_size
  )) %>%
  group_by(PATHWAY) %>%
  summarize(
    avg_group_size = mean(group_size, na.rm = TRUE),
    median_group_size = median(group_size, na.rm = TRUE),
    min_group_size = min(group_size, na.rm = TRUE),
    max_group_size = max(group_size, na.rm = TRUE),
    count = n()
  )

```

## save group size summary
```{r save group size summary}
setwd(here::here('output','aqim pathway approach rates'))
write.csv(group_size_summary, paste(params$group_size_filename,".csv", sep=""), row.names = FALSE)
saveRDS(group_size_summary, paste(params$group_size_filename,".rds", sep=""))


#save summary tables for validating the data
write.csv(combinations_summary, "aqim pathway by product combination summary of total vs at risk.csv", row.names = FALSE)
write.csv(pathway_summary, "aqim pathway summary of total vs at risk.csv", row.names = FALSE)
```


```{r}
aqim_dat |> filter(PATHWAY == "Airport - Aircraft - Person - CBP") |> filter(INSPECTION_LOCATION_NAME == "LOS ANGELES INT AIRPORT (2720)") |> count(TOTAL_KG > 0, IS_ASF_RISK)
```

## Validation: Compare actual counts to predicted probabilities
```{r validation}
# Validate probabilities for specific locations and groups to ensure accuracy
# Focus on Los Angeles Int Airport for the 'Airport - Aircraft - Person - CBP' pathway

# Target pathway and location
target_pathway <- "Airport - Aircraft - Person - CBP"
target_location <- "LOS ANGELES INT AIRPORT (2720)"

# Get actual counts from raw data
actual_counts <- aqim_dat %>% 
  filter(PATHWAY == target_pathway, 
         INSPECTION_LOCATION_NAME == target_location) %>%
  summarize(
    total_count = n(),
    risk_count = sum(IS_ASF_RISK, na.rm = TRUE),
    actual_rate = risk_count / total_count
  )

# Print actual counts and rate
cat("Actual counts for", target_location, "in pathway", target_pathway, ":\n")
cat("  Total observations:", actual_counts$total_count, "\n")
cat("  Risk observations:", actual_counts$risk_count, "\n")
cat("  Actual rate:", round(actual_counts$actual_rate * 100, 2), "%\n\n")

# Get predicted probabilities from our model for different groups
predicted_probs <- location_probabilities_final %>%
  filter(pathway == target_pathway, 
         location == target_location)

# Print predicted probabilities by group
cat("Predicted probabilities by group:\n")
for(i in 1:nrow(predicted_probs)) {
  group_parts <- strsplit(predicted_probs$group[i], ";")[[1]]
  container <- group_parts[2]
  product <- paste(group_parts[3], "-", group_parts[4])
  
  cat("  Group:", container, "/", product, "\n")
  cat("    Predicted probability:", round(predicted_probs$probability[i] * 100, 2), "%\n")
  cat("    95% CI: [", round(predicted_probs$ci_lower[i] * 100, 2), "%, ", 
      round(predicted_probs$ci_upper[i] * 100, 2), "%]\n")
}

# Compare with the previous GLM approach (for reference, already saved in output folder)
previous_probs_file <- file.path(here::here('output','aqim pathway approach rates'), 
                               "glmm_results_aqim fraction at risk by pathway and location.csv")

if(file.exists(previous_probs_file)) {
  previous_probs <- read.csv(previous_probs_file) %>%
    filter(pathway == target_pathway, 
           location == target_location)
  
  cat("\nPrevious GLM-based probabilities:\n")
  for(i in 1:nrow(previous_probs)) {
    group_parts <- strsplit(previous_probs$group[i], ";")[[1]]
    container <- group_parts[2]
    product <- paste(group_parts[3], "-", group_parts[4])
    
    cat("  Group:", container, "/", product, "\n")
    cat("    Predicted probability:", round(previous_probs$probability[i] * 100, 2), "%\n")
  }
}

# For top 5 locations with most observations, compare actual vs. predicted rates
top_locations <- aqim_dat %>%
  filter(PATHWAY == target_pathway) %>%
  count(INSPECTION_LOCATION_NAME) %>%
  arrange(desc(n)) %>%
  head(5) %>%
  pull(INSPECTION_LOCATION_NAME)

cat("\nValidation for top 5 busiest locations in", target_pathway, "pathway:\n")

validation_table <- data.frame()

for(loc in top_locations) {
  # Actual counts
  actual <- aqim_dat %>% 
    filter(PATHWAY == target_pathway, 
           INSPECTION_LOCATION_NAME == loc) %>%
    summarize(
      total_count = n(),
      risk_count = sum(IS_ASF_RISK, na.rm = TRUE),
      actual_rate = risk_count / total_count
    )
  
  # Predicted probabilities (averaged across all groups)
  predicted <- location_probabilities_final %>%
    filter(pathway == target_pathway, 
           location == loc,
           location != "Average") %>%
    summarize(
      avg_predicted = mean(probability, na.rm = TRUE)
    )
  
  # Add to validation table
  validation_table <- rbind(validation_table, data.frame(
    Location = loc,
    TotalCount = actual$total_count,
    RiskCount = actual$risk_count,
    ActualRate = actual$actual_rate,
    PredictedRate = if(nrow(predicted) > 0) predicted$avg_predicted else NA
  ))
}

# Print validation table
print(validation_table %>%
        mutate(
          ActualRate = paste0(round(ActualRate * 100, 2), "%"),
          PredictedRate = paste0(round(PredictedRate * 100, 2), "%")
         ))
```

## Summary of Probability Calculation Improvements

This script now calculates location-specific probabilities using a more accurate empirical approach rather than relying solely on GLM coefficients. The main improvements include:

1. **Direct Empirical Calculation**: Instead of relying on GLM coefficients which can produce extreme values on the log-odds scale, we now directly calculate empirical probabilities from actual counts for each location.

2. **Wilson Score Intervals**: For confidence intervals, we use the Wilson score method which is more appropriate for small counts and proportions near 0 or 1.

3. **Validation Against Raw Data**: We compare the predicted probabilities against actual observed rates to ensure accuracy.

4. **Compatibility with Original Approach**: The code maintains compatibility with the original GLM approach for comparison purposes.

These changes ensure that the location-specific probabilities in `location_probabilities_final` are realistic and accurately reflect the observed data, addressing the issue where GLM-based probabilities were unrealistically high for some locations like Los Angeles Int Airport.
