---
title: "aqim pathway approach rates"
author: "Dalon White"
date: "2025-07-27"
output: html_document

params:
  risk_quantification: "0_1" # 0_1 or "range"
  affected_countries: "current" # "current" or "mex_can" or "other"
  other_affected_countries: c("Country1 ISO-3 Code", "Country2 ISO-3 Code") # If using "other", specify the countries here
  years: 2023:2024 # The years to include which countries have ASF from the `asf affected countries.csv` file
  glm_results_filename: "glmm_results_aqim fraction at risk by pathway and location"
  bootstrap_results_filename: "bootstrap_results_aqim fraction at risk by pathway and location"
  group_size_filename: "aqim_mean_pathway_group_size"
---

-   prepped data already has the pathway, entry type, etc., and risk as no/low/med/high
-   need to use country of origin and product origin to determine if the inspection has any asf risk
-   then model the kg of product with asf risk by pathway, entry type, location, country of origin, product pathway, product subpathway

```{r setup, include=FALSE}
library(tidyverse)
library(odbc)
library(DBI)
library(lme4)
library(ggplot2)
library(broom)
library(multcomp)
library(emmeans)
library(sjPlot)
library(broom.mixed)
library(boot)
library(here)
```

```{r load-params}
# Function to load the shared parameters if available
load_shared_params <- function() {
  params_file <- here::here("output", "intermediate files", "asf_risk_params.rds")
  
  if (file.exists(params_file)) {
    message("Loading shared parameters from 00_asf_risk_parameters.Rmd")
    return(readRDS(params_file))
  } else {
    message("Shared parameter file not found. Using script-specific parameters.")
    return(NULL)
  }
}

# Try to load shared params
shared_params <- load_shared_params()

# Use shared params if available, otherwise use script-specific params
if (!exists("params")) {
  if (!is.null(shared_params)) {
    params <- shared_params
  } else {
    # Default params specific to this script
    params <- list(
      risk_quantification = "0_1", # 0_1 or "range"
      affected_countries = "current", # "current" or "mex_can" or "other" 
      years = 2024:2024, #The years to include which countries have ASF
      other_affected_countries = c("Country1", "Country2"), 
      glm_results_filename = "glmm_results_aqim fraction at risk by pathway and location",
      bootstrap_results_filename = "bootstrap_results_aqim fraction at risk by pathway and location",
      group_size_filename = "aqim_mean_pathway_group_size"
    )
  }
}
```

# Call AQIM data

```{r get data}

aqim_dat <- read.csv("C:/Users/Dalon.White/OneDrive - USDA/Desktop/Projects/asf-risk-framework-large-files/aqim pathway data prep.csv")

```

# Set affected countries for analysis

Within params, can choose whether to use the current affected countries or just Mexico and Canada. The current ones, are then filtered by the years specified in params$years.
If only interested in the last year, then set params$years to the last year of interest. It is important to remember that we are not predicting where surveillance needs to be, such as for the CEAH group, but identifying the risk of ASF in products that are coming into the US, so whether an item corresponds to a year and country is not important. Instead, we are focused on estimating how much comes from each country and whether that country has ASF risk.

```{r identify countries of risk}
setwd(here::here("input","data","asf affected countries by year"))
if(params$affected_countries == "current") {
  affected_countries <- read.csv("asf affected countries.csv") |> 
    mutate(Year = as.numeric(Year)) |> 
    filter(Year %in% params$years) |> 
    filter(Disease == "ASF") |>
    distinct(ISO3_CODE) |> 
    pull(ISO3_CODE)
} else if (params$affected_countries == "mex_can") {
  affected_countries <- c("MEX", "CAN")
} else if (params$affected_countries == "other") {
  affected_countries <- params$other_affected_countries
} else {
  stop("Invalid value for params$affected_countries")
}
```


# Assign risk by country of origin and product origin

```{r assign risk}
# Filter for affected countries
aqim_dat <- aqim_dat %>% mutate(IS_ASF_RISK = case_when(
  edge_origin %in% affected_countries | product_origin %in% affected_countries ~ IS_ASF_RISK,
  .default = "NO"
         )
)
#once it is filtered by country, this is the at risk products by weight
aqim_dat |> filter(PATHWAY== "Pedestrian SB") |> count(TOTAL_KG > 0, IS_ASF_RISK)
#at risk product types exist, but they did not come from countries with ASF risk

```

# trouble shooting
Express courier is coming out too high in the final data layer - like 25% of packages are smuggled commercial prodcut -, and has entries from 'baggage' - there shouldn't be any baggage conveyances for ECO
and san francisco has 0.0311 smuggled commercial product AND 0.0311 swine products in passenger baggage; they shouldn't be the same necessarily so this might indicate a problem
- I fixed the baggage issue by correcting a reference file, it now says 'package'. Same with mail
- I have determined that the rate of at risk baggage in SF Int'l really was 328 out of ~10k (3%)
- I corrected an error in the reference table that wasnt assigning a value to TOTAL_KG for most AQAS aqim data because the UNIT name didn't match (mostly just missing 's')
- Need to check that this works now
```{r troubleshooting A}
# #Looking down at combinations_summary I can see that ECO does have some at-risk items but it's only a 51 out of 63k items
# #There is only one smuggled commercial product in an ECO package
# aqim_dat |> filter(PATHWAY== "Express_courier") |> count(TOTAL_KG > 0, IS_ASF_RISK)
# aqim_dat |> filter(PATHWAY== "Express_courier" & IS_ASF_RISK > 0)
# View(aqim_dat |> filter(PATHWAY== "Express_courier" & IS_ASF_RISK > 0))
# #Its a personal use item that has been appropraitely assigned to smuggling because it is 26 kilos

# # look at san francisco airport passenger baggage data
# SF_bags <- aqim_dat |> filter(PATHWAY == "Airport - Aircraft - Person - CBP" & grepl("SAN FRAN", INSPECTION_LOCATION_NAME)) |> count(TOTAL_KG > 0, IS_ASF_RISK)
# print(SF_bags)
# #at risk entries in SF Int'l Airport
# print(paste("at risk entries in SF Int'l Airport is", round(
# SF_bags[SF_bags$`TOTAL_KG > 0` == TRUE & SF_bags$IS_ASF_RISK == 1,]$n/sum(SF_bags$n)* 100, 2), "percent"
# ))


```

## handle risk quantification based on 0-1 or a scale

```{r handle risk quantification}
if(params$risk_quantification == "0_1") {
  # Convert risk to binary (0 or 1)
  aqim_dat <- aqim_dat %>%
    mutate(IS_ASF_RISK = case_when(
      IS_ASF_RISK == "HIGH" ~ 1,
      IS_ASF_RISK == "MEDIUM" ~ 1,
      IS_ASF_RISK == "LOW" ~ 1,
      .default = 0
    ))
} else if (params$risk_quantification == "range") {
  # Keep risk as is (no conversion)
  aqim_dat <- aqim_dat %>%
    mutate(IS_ASF_RISK = case_when(
      IS_ASF_RISK == "HIGH" ~ 1,
      IS_ASF_RISK == "MEDIUM" ~ .67,
      IS_ASF_RISK == "LOW" ~ .33,
      .default = 0
    ))
} else {
  stop("Invalid value for params$risk_quantification")
}

# modify TOTAL_KG by IS_ASF_RISK
aqim_dat <- aqim_dat |> mutate(TOTAL_KG = TOTAL_KG * IS_ASF_RISK)

# # modify IS_ASF_RISK to be binary
# aqim_dat <- aqim_dat |> mutate(IS_ASF_RISK = ifelse(IS_ASF_RISK > 0, 1, 0))

```

## Create Summary
This is to make sure I remember that not all these pathways have risk, so not all of them will be used in the 01_06 aqim pathway risk_PoE.Rmd script

```{r}
# Create summary statistics for combinations and risk counts
# Count all combinations by risk status
combinations_summary <- aqim_dat |>
  dplyr::count(PATHWAY, container, product_pathway, product_subpathway, IS_ASF_RISK)

# Get key statistics
pathway_counts <- combinations_summary |> distinct(PATHWAY) |> nrow()
total_combinations <- combinations_summary |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()
risk_combinations <- combinations_summary |> filter(IS_ASF_RISK > 0) |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()
no_risk_combinations <- combinations_summary |> filter(IS_ASF_RISK == 0) |> distinct(PATHWAY, container, product_pathway, product_subpathway) |> nrow()

# Count observations
total_risk_count <- combinations_summary |> filter(IS_ASF_RISK > 0) |> summarize(total = sum(n)) |> pull(total)
total_no_risk_count <- combinations_summary |> filter(IS_ASF_RISK == 0) |> summarize(total = sum(n)) |> pull(total)
grand_total_count <- sum(combinations_summary$n)

# Create pathway-level summary
pathway_summary <- combinations_summary |>
  group_by(PATHWAY, IS_ASF_RISK) |>
  summarize(count = sum(n), .groups = "drop") |>
  pivot_wider(names_from = IS_ASF_RISK, values_from = count, 
              names_prefix = "risk_", values_fill = 0) |>
  mutate(total = risk_0 + risk_1,
         risk_percent = round(100 * risk_1 / total, 1))

# Print summary information
cat("\n### Summary of AQIM Data Combinations ###\n\n")
cat(paste("Total unique pathways:", pathway_counts, "\n"))
cat(paste("Total unique PATHWAY-container-product combinations:", total_combinations, "\n"))
cat(paste("- Combinations with ASF risk:", risk_combinations, 
          sprintf("(%.1f%%)", 100 * risk_combinations/total_combinations), "\n"))
cat(paste("- Combinations without ASF risk:", no_risk_combinations, "\n\n"))

cat("### Observation Counts ###\n\n")
cat(paste("Total observations:", format(grand_total_count, big.mark=","), "\n"))
cat(paste("- Observations with ASF risk:", format(total_risk_count, big.mark=","), 
          sprintf("(%.1f%%)", 100 * total_risk_count/grand_total_count), "\n"))
cat(paste("- Observations without ASF risk:", format(total_no_risk_count, big.mark=","), "\n\n"))

cat("###  Pathways by Risk Percentage ###\n\n")
top_risk_pathways <- pathway_summary |>
  filter(risk_1 > 0) |>
  arrange(desc(risk_percent)) 

print(top_risk_pathways |> dplyr::select(PATHWAY, risk_1, total, risk_percent))
print("The pathway combinations that exist are in the following table:")
View(combinations_summary)

```

```{r split data by pathway}
# Group data by location, pathway, risk, container, and product types
aqim_dat_counts <- aqim_dat |> 
  group_by(INSPECTION_LOCATION_NAME, PATHWAY, IS_ASF_RISK, container, product_pathway, product_subpathway) |> 
  dplyr::count()

# Group_by group_by(INSPECTION_LOCATION_NAME, PATHWAY, container, product_pathway, product_subpathway) and for each row, find the sum of n where IS_ASF_RISK > 0
aqim_dat_pos_counts <- aqim_dat_counts |>
  group_by(INSPECTION_LOCATION_NAME, PATHWAY, container, product_pathway, product_subpathway) |>
  summarize(n = sum(n[IS_ASF_RISK > 0]), .groups = "drop")


# Get total observations for each location-pathway combination
location_pathway_totals <- aqim_dat |>
  group_by(INSPECTION_LOCATION_NAME, PATHWAY) |>
  summarize(total_obs = n(), .groups = "drop")

# Join the totals to the aqim_dat_counts
aqim_events_summary <- aqim_dat_pos_counts |>
  left_join(location_pathway_totals, by = c("INSPECTION_LOCATION_NAME", "PATHWAY"))

# Remove rows with blank container, product_pathway, and product_subpathway, as these were just counts of the total observations by location and pathway
aqim_events_summary <- aqim_events_summary |>
  filter(
    !is.na(container) & container != "", 
    !is.na(product_pathway) & product_pathway != "", 
    !is.na(product_subpathway) & product_subpathway != ""
  ) |> 
  dplyr::rename(n_positive = n)
# View the first few rows to verify
head(aqim_dat_counts)
```

# Need to predict both the incident rate per entry, and the kg of product that is intercepted.
I want to implement a hierarchical bayesian approach to estimating the likelihood of an at-risk ASF product entering
- The estimates should be hierarchical by pathway, entry type, and location

# Bayesian Hierarchical Model for Event Proportions
I want to implement a hierarchical bayesian approach to estimating the likelihood of an at-risk ASF product entering.
The estimates should be hierarchical by pathway, entry type, and location.

```{r install_packages, eval=FALSE}
# Install required packages if not already installed
if (!requireNamespace("brms", quietly = TRUE)) {
  install.packages("brms")
}
if (!requireNamespace("bayesplot", quietly = TRUE)) {
  install.packages("bayesplot")
}
if (!requireNamespace("tidybayes", quietly = TRUE)) {
  install.packages("tidybayes")
}
```

```{r load_bayes_packages}
# Load required packages
library(brms)      # For Bayesian modeling
library(bayesplot) # For plotting Bayesian results
library(tidybayes) # For tidying Bayesian results
library(ggplot2)   # For visualization

# Set seed for reproducibility
set.seed(123)
```

```{r prepare_data_for_model}
# Ensure we have the necessary data structure
# The model will use n_positive and total_obs
bayes_data <- aqim_events_summary %>%
  # Remove any rows with missing values or where total_obs is 0
  filter(!is.na(n_positive), !is.na(total_obs), total_obs > 0) %>%
  # Ensure n_positive doesn't exceed total_obs (data validation)
  mutate(n_positive = pmin(n_positive, total_obs)) %>%
  # Create a unique ID for each observation
  mutate(obs_id = row_number())

# Check data structure and dimensions
print("Structure of bayes_data:")
str(bayes_data)
print("Dimensions of bayes_data:")
print(dim(bayes_data))

# View the first few rows
head(bayes_data)

# Summary statistics
summary_stats <- bayes_data %>%
  summarize(
    total_locations = n_distinct(INSPECTION_LOCATION_NAME),
    total_pathways = n_distinct(PATHWAY),
    total_containers = n_distinct(container),
    total_product_pathways = n_distinct(product_pathway),
    total_product_subpathways = n_distinct(product_subpathway),
    mean_positive_rate = sum(n_positive) / sum(total_obs),
    median_positive_rate = median(n_positive / total_obs)
  )

print(summary_stats)
```

```{r define_hierarchical_model, results='hide'}
# Define the hierarchical model
# This is a binomial model with a logit link function
# The hierarchical structure follows the data organization:
# INSPECTION_LOCATION_NAME > PATHWAY > (container, product_pathway) > product_subpathway

# First check if we have data to model
if (nrow(bayes_data) == 0 || all(bayes_data$n_positive == 0)) {
  stop("No positive observations in the data - cannot fit the model")
}

# Print summary statistics to verify data
cat("Summary of data for Bayesian model:\n")
cat("Total observations:", nrow(bayes_data), "\n")
cat("Total positive events:", sum(bayes_data$n_positive), "\n")
cat("Average positive rate:", round(sum(bayes_data$n_positive) / sum(bayes_data$total_obs), 4), "\n")

# Check for any very small values that might cause numerical issues
min_n <- min(bayes_data$n_positive[bayes_data$n_positive > 0])
min_total <- min(bayes_data$total_obs)
cat("Minimum positive count (where > 0):", min_n, "\n")
cat("Minimum total observations:", min_total, "\n\n")

# Start with a simple version of the model
# We'll build up complexity based on performance
# Set refresh=100 to see progress during fitting
bayes_model <- brm(
  # Binomial response with n_positive successes out of total_obs trials
  n_positive | trials(total_obs) ~ 1 + 
    # Random intercepts for location
    (1 | INSPECTION_LOCATION_NAME) + 
    # Random intercepts for container and product pathway
    (1 | container) + 
    (1 | product_pathway) +
    # Random intercepts for product subpathway
    (1 | product_pathway:product_subpathway),
  
  # Data and family
  data = bayes_data,
  family = binomial(link = "logit"),
  # MCMC settings
  chains = 4,           # Number of Markov chains
  iter = 4000,          # Total iterations per chain
  warmup = 2000,        # Warmup iterations per chain
  cores = 4,            # Number of cores to use
  seed = 12345,         # Set seed for reproducibility
  refresh = 100,        # Show progress during fitting
  control = list(
    adapt_delta = 0.95,  # Increase adapt_delta to reduce divergences
    max_treedepth = 12   # Increase max_treedepth to allow for more complex models
  ),
  
  # Save results for future use
  file = here::here("output", "intermediate files", "asf_risk_hierarchical_model.rds")
)

# Verify that the model contains posterior samples
if (is.null(bayes_model) || !("brmsfit" %in% class(bayes_model))) {
  stop("Model fitting failed - no brmsfit object returned")
}

# Print some basic information about the fitted model
cat("\nModel fitting completed.\n")
cat("Number of parameters:", length(names(bayes_model$fit@sim$samples[[1]])), "\n")
cat("Number of observations:", nrow(bayes_model$data), "\n")

# Note: This model can take a while to run, especially with a large dataset
# The increased iterations (4000) and warmup (2000) are necessary for proper convergence
# For complex hierarchical models, more iterations may be needed to reach convergence
# You may need to adjust the MCMC settings based on your data size and complexity
```
