---
title: "GLMM by Pathway Approach for ASF Risk Prediction"
author: "Dalon White"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lme4)
library(ggplot2)
library(knitr)
library(kableExtra)
library(here)
library(scales)
```

## Introduction

This document implements a simplified approach to the ASF risk prediction problem by:

1. Splitting the data by PATHWAY
2. Fitting a separate GLMM for each pathway
3. Combining the results for comprehensive analysis

This approach simplifies the model structure and improves convergence by treating each pathway independently.

## Data Preparation

First, let's load the same data used in the Bayesian approach:

```{r load_data}
# Load the aqim_events_summary data
if (file.exists(here::here("output", "intermediate files", "aqim_events_summary.rds"))) {
  aqim_events_summary <- readRDS(here::here("output", "intermediate files", "aqim_events_summary.rds"))
} else {
  stop("Could not find aqim_events_summary data. Please run the data preparation steps first.")
}

# Prepare the data using the same process as in the Bayesian model
bayes_data <- aqim_events_summary %>%
  # Remove any rows with missing values or where total_obs is 0
  filter(!is.na(n_positive), !is.na(total_obs), total_obs > 0) %>%
  # Ensure n_positive doesn't exceed total_obs (data validation)
  mutate(n_positive = pmin(n_positive, total_obs)) %>%
  # Create a unique ID for each observation
  mutate(obs_id = row_number())

# Display data dimensions and structure
cat("Data dimensions:", dim(bayes_data), "\n")
cat("Number of unique pathways:", length(unique(bayes_data$PATHWAY)), "\n")
cat("Number of locations:", length(unique(bayes_data$INSPECTION_LOCATION_NAME)), "\n")
cat("Total positive events:", sum(bayes_data$n_positive), "\n")
cat("Average positive rate:", 
    round(100 * sum(bayes_data$n_positive) / sum(bayes_data$total_obs), 4), "%\n")

# Show distribution of observations by pathway
pathway_counts <- bayes_data %>%
  group_by(PATHWAY) %>%
  summarize(
    observations = n(),
    locations = n_distinct(INSPECTION_LOCATION_NAME),
    positives = sum(n_positive),
    positive_rate = round(100 * sum(n_positive) / sum(total_obs), 2)
  ) %>%
  arrange(desc(observations))

kable(pathway_counts, caption = "Data summary by pathway") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## GLMM by Pathway Approach

Now we'll implement the approach where we fit a separate GLMM for each pathway:

```{r define_pathway_function}
# Function to fit GLMM for a single pathway
fit_pathway_glmm <- function(pathway_data, pathway_name) {
  # Print info about this pathway
  cat("\n--------------------------------------------------------\n")
  cat("Processing pathway:", pathway_name, "\n")
  cat("Number of observations:", nrow(pathway_data), "\n")
  cat("Number of locations:", length(unique(pathway_data$INSPECTION_LOCATION_NAME)), "\n")
  cat("Total positive events:", sum(pathway_data$n_positive), "\n")
  cat("Average positive rate:", 
      round(100 * sum(pathway_data$n_positive) / sum(pathway_data$total_obs), 4), "%\n")
  
  # Skip pathways with insufficient data
  if (nrow(pathway_data) < 5) {
    cat("WARNING: Too few observations (< 5) for pathway", pathway_name, "- skipping\n")
    return(NULL)
  }
  
  if (sum(pathway_data$n_positive) == 0) {
    cat("WARNING: No positive events for pathway", pathway_name, "- skipping\n")
    return(NULL)
  }
  
  # Try to fit the model with error handling
  tryCatch({
    # Fit the GLMM model
    model <- glmer(
      cbind(n_positive, total_obs - n_positive) ~ 1 + 
        (1 | INSPECTION_LOCATION_NAME) + 
        (1 | container) + 
        (1 | product_pathway) +
        (1 | product_pathway:product_subpathway),
      data = pathway_data,
      family = binomial("logit"),
      control = glmerControl(optimizer = "bobyqa")
    )
    
    # Get fixed effect (global intercept)
    fixed_effect <- fixef(model)[1]
    cat("Fixed effect (intercept):", round(fixed_effect, 4), "\n")
    cat("Overall baseline probability:", round(plogis(fixed_effect), 4), "\n")
    
    # Get predicted probabilities for each observation
    # Method 1: Simple fitted values
    fitted_values <- predict(model, newdata = pathway_data, type = "link")
    fitted_probs <- plogis(fitted_values)
    
    # Calculate standard errors using the delta method
    se_link <- sqrt(fitted_probs * (1 - fitted_probs) / pathway_data$total_obs)
    
    # Calculate confidence intervals
    lower_probs <- plogis(fitted_values - 1.96 * se_link)
    upper_probs <- plogis(fitted_values + 1.96 * se_link)
    
    # Create results data frame
    results <- pathway_data %>%
      mutate(
        pathway = pathway_name,
        pred_prob = fitted_probs,
        lower_prob = lower_probs,
        upper_prob = upper_probs,
        observed_rate = n_positive / total_obs
      )
    
    cat("Model successfully fit for pathway:", pathway_name, "\n")
    return(list(model = model, results = results))
    
  }, error = function(e) {
    cat("ERROR fitting model for pathway", pathway_name, ":", conditionMessage(e), "\n")
    return(NULL)
  })
}
```

Let's fit models for a few selected pathways to demonstrate the approach:

```{r fit_selected_models}
# Select a few interesting pathways
# You can adjust this list based on which pathways are important for your analysis
selected_pathways <- c(
  "Airport - Aircraft - Person - CBP",
  "Express_courier",
  "Mail",
  "Ship - Cargo - AQIM"
)

# Initialize lists to store results
selected_models <- list()
selected_results <- list()

# Process each selected pathway
for (pathway in selected_pathways) {
  # Skip if pathway doesn't exist in the data
  if (!(pathway %in% unique(bayes_data$PATHWAY))) {
    cat("Pathway", pathway, "not found in data - skipping\n")
    next
  }
  
  # Extract data for this pathway
  pathway_data <- bayes_data %>% filter(PATHWAY == pathway)
  
  # Fit model for this pathway
  result <- fit_pathway_glmm(pathway_data, pathway)
  
  # Store results if successful
  if (!is.null(result)) {
    selected_models[[pathway]] <- result$model
    selected_results[[pathway]] <- result$results
  }
}

# Combine results for selected pathways
combined_selected_results <- bind_rows(selected_results)
```

Let's examine the results for these pathways:

```{r examine_results, fig.height=8, fig.width=10}
# Create pathway summary
pathway_summary <- combined_selected_results %>%
  group_by(pathway) %>%
  summarize(
    total_obs = sum(total_obs),
    n_positive = sum(n_positive),
    observed_rate = n_positive / total_obs,
    pred_prob = mean(pred_prob),
    lower_prob = mean(lower_prob),
    upper_prob = mean(upper_prob),
    .groups = "drop"
  ) %>%
  arrange(desc(pred_prob))

# Display pathway summary
kable(pathway_summary %>%
      mutate(across(c(observed_rate, pred_prob, lower_prob, upper_prob), ~percent(.x, 0.01))),
      caption = "Predicted risk probabilities by pathway") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot pathway-level results
ggplot(pathway_summary, 
       aes(x = reorder(pathway, pred_prob), y = pred_prob)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), width = 0.2) +
  geom_point(aes(y = observed_rate), shape = 4, size = 3, color = "red") +  
  coord_flip() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Pathway Risk Probabilities",
    subtitle = "Error bars represent 95% confidence intervals. X marks show observed rates.",
    x = "Pathway",
    y = "Estimated Probability"
  ) +
  theme_minimal()

# Create location summary (for selected pathways)
location_summary <- combined_selected_results %>%
  group_by(pathway, INSPECTION_LOCATION_NAME) %>%
  summarize(
    total_obs = sum(total_obs),
    n_positive = sum(n_positive),
    observed_rate = n_positive / total_obs,
    pred_prob = mean(pred_prob),
    lower_prob = mean(lower_prob),
    upper_prob = mean(upper_prob),
    .groups = "drop"
  ) %>%
  arrange(pathway, desc(pred_prob))

# Display top locations for each pathway
for (path in unique(location_summary$pathway)) {
  cat("\n\n### Top Locations for Pathway:", path, "\n")
  
  top_locations <- location_summary %>% 
    filter(pathway == path) %>% 
    head(10)
  
  print(kable(top_locations %>%
        select(INSPECTION_LOCATION_NAME, total_obs, n_positive, observed_rate, pred_prob, lower_prob, upper_prob) %>%
        mutate(across(c(observed_rate, pred_prob, lower_prob, upper_prob), ~percent(.x, 0.01)))) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")))
}
```

## Running All Pathways

Now let's process all pathways and combine the results:

```{r process_all_pathways}
# Process all pathways (set eval=FALSE to skip this computation-intensive step when knitting)
process_all_pathways <- function(bayes_data) {
  # Split data by PATHWAY
  pathways <- unique(bayes_data$PATHWAY)
  cat("Found", length(pathways), "unique pathways to process\n")
  
  # Initialize lists to store results
  all_models <- list()
  all_results <- list()
  
  # Process each pathway
  for (pathway in pathways) {
    # Extract data for this pathway
    pathway_data <- bayes_data %>% filter(PATHWAY == pathway)
    
    # Fit model for this pathway
    result <- fit_pathway_glmm(pathway_data, pathway)
    
    # Store results if successful
    if (!is.null(result)) {
      all_models[[pathway]] <- result$model
      all_results[[pathway]] <- result$results
    }
  }
  
  # Combine all results
  combined_results <- bind_rows(all_results)
  
  # Create output directory if it doesn't exist
  output_dir <- here::here("output", "aqim pathway approach rates")
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # Save the combined results
  results_file <- file.path(output_dir, "glmm_results_aqim_fraction_at_risk.csv")
  write.csv(combined_results, results_file, row.names = FALSE)
  saveRDS(combined_results, file.path(output_dir, "glmm_results_aqim_fraction_at_risk.rds"))
  
  # Save the models
  models_file <- file.path(output_dir, "glmm_models_by_pathway.rds")
  saveRDS(all_models, models_file)
  
  cat("\nResults saved to:", results_file, "\n")
  cat("Models saved to:", models_file, "\n")
  
  return(list(
    models = all_models,
    results = combined_results
  ))
}

# Optionally run for all pathways
# Uncomment the line below to process all pathways
# all_results <- process_all_pathways(bayes_data)
```

## Different Methods for Converting to Predicted Probabilities

There are several methods to convert log-odds to predicted probabilities for each observation. Here are three options, from simplest to most advanced:

### Method 1: Direct Transformation (Already Implemented)

The simplest approach is to get the fitted values on the link (log-odds) scale and transform them using the inverse logit function:

```{r method1_example, eval=FALSE}
# Get fitted values on the link scale (log-odds)
fitted_values <- predict(model, newdata = data, type = "link")
# Transform to probability scale
fitted_probs <- plogis(fitted_values)
```

This is the method we used in the code above. It's fast and reliable but doesn't fully account for the uncertainty in the random effects.

### Method 2: Model Matrix Approach

A more direct approach using the model matrix:

```{r method2_example, eval=FALSE}
method2_predict <- function(model, newdata) {
  # Get fixed effect (intercept)
  fixed_effect <- fixef(model)[1]
  
  # Extract random effects
  ranef_location <- ranef(model)$INSPECTION_LOCATION_NAME
  ranef_container <- ranef(model)$container
  ranef_product <- ranef(model)$product_pathway
  ranef_subpathway <- ranef(model)$`product_pathway:product_subpathway`
  
  # Initialize predictions
  predictions <- rep(fixed_effect, nrow(newdata))
  
  # Add random effects for each observation
  for (i in 1:nrow(newdata)) {
    # Add location effect
    loc <- as.character(newdata$INSPECTION_LOCATION_NAME[i])
    if (loc %in% rownames(ranef_location)) {
      predictions[i] <- predictions[i] + ranef_location[loc, 1]
    }
    
    # Add container effect
    cont <- as.character(newdata$container[i])
    if (cont %in% rownames(ranef_container)) {
      predictions[i] <- predictions[i] + ranef_container[cont, 1]
    }
    
    # Add product_pathway effect
    prod <- as.character(newdata$product_pathway[i])
    if (prod %in% rownames(ranef_product)) {
      predictions[i] <- predictions[i] + ranef_product[prod, 1]
    }
    
    # Add product_pathway:product_subpathway effect
    subpath <- paste0(newdata$product_pathway[i], ":", newdata$product_subpathway[i])
    if (subpath %in% rownames(ranef_subpathway)) {
      predictions[i] <- predictions[i] + ranef_subpathway[subpath, 1]
    }
  }
  
  # Convert to probabilities
  return(plogis(predictions))
}
```

### Method 3: Simulation Approach (Bootstrap)

For the most accurate confidence intervals that account for uncertainty in both fixed and random effects:

```{r method3_example, eval=FALSE}
bootstrap_predict <- function(model, newdata, n_boot = 1000) {
  # Extract fixed effect
  fe <- fixef(model)[1]
  fe_se <- sqrt(vcov(model)[1,1])  # SE for fixed effect
  
  # Extract random effects
  ranef_location <- ranef(model)$INSPECTION_LOCATION_NAME
  ranef_container <- ranef(model)$container
  ranef_product <- ranef(model)$product_pathway
  ranef_subpathway <- ranef(model)$`product_pathway:product_subpathway`
  
  # Get SEs for random effects
  se_location <- sqrt(diag(attr(ranef(model, condVar=TRUE)$INSPECTION_LOCATION_NAME, "postVar")[1,,]))
  se_container <- sqrt(diag(attr(ranef(model, condVar=TRUE)$container, "postVar")[1,,]))
  se_product <- sqrt(diag(attr(ranef(model, condVar=TRUE)$product_pathway, "postVar")[1,,]))
  se_subpathway <- sqrt(diag(attr(ranef(model, condVar=TRUE)$`product_pathway:product_subpathway`, "postVar")[1,,]))
  
  # Initialize results matrix
  boot_results <- matrix(NA, nrow = nrow(newdata), ncol = n_boot)
  
  # Run bootstrap iterations
  for (i in 1:n_boot) {
    # Simulate new fixed effect
    new_fe <- rnorm(1, fe, fe_se)
    
    # Simulate new random effects
    new_ranef_location <- rnorm(length(ranef_location), ranef_location[,1], se_location)
    new_ranef_container <- rnorm(length(ranef_container), ranef_container[,1], se_container)
    new_ranef_product <- rnorm(length(ranef_product), ranef_product[,1], se_product)
    new_ranef_subpathway <- rnorm(length(ranef_subpathway), ranef_subpathway[,1], se_subpathway)
    
    # For each observation, compute prediction with simulated effects
    for (j in 1:nrow(newdata)) {
      # Start with global intercept
      logit_pred <- new_fe
      
      # Add location effect if applicable
      loc <- as.character(newdata$INSPECTION_LOCATION_NAME[j])
      if (loc %in% rownames(ranef_location)) {
        loc_idx <- which(rownames(ranef_location) == loc)
        logit_pred <- logit_pred + new_ranef_location[loc_idx]
      }
      
      # Add container effect if applicable
      cont <- as.character(newdata$container[j])
      if (cont %in% rownames(ranef_container)) {
        cont_idx <- which(rownames(ranef_container) == cont)
        logit_pred <- logit_pred + new_ranef_container[cont_idx]
      }
      
      # Add product_pathway effect if applicable
      prod <- as.character(newdata$product_pathway[j])
      if (prod %in% rownames(ranef_product)) {
        prod_idx <- which(rownames(ranef_product) == prod)
        logit_pred <- logit_pred + new_ranef_product[prod_idx]
      }
      
      # Add product_pathway:product_subpathway effect if applicable
      subpath <- paste0(newdata$product_pathway[j], ":", newdata$product_subpathway[j])
      if (subpath %in% rownames(ranef_subpathway)) {
        subpath_idx <- which(rownames(ranef_subpathway) == subpath)
        logit_pred <- logit_pred + new_ranef_subpathway[subpath_idx]
      }
      
      # Convert from logit to probability and store
      boot_results[j, i] <- plogis(logit_pred)
    }
  }
  
  # Calculate summary statistics
  pred_mean <- rowMeans(boot_results)
  pred_lower <- apply(boot_results, 1, quantile, probs = 0.025)
  pred_upper <- apply(boot_results, 1, quantile, probs = 0.975)
  
  return(data.frame(
    pred = pred_mean,
    lower = pred_lower,
    upper = pred_upper
  ))
}
```

## Conclusion

The pathway-specific GLMM approach offers several advantages:

1. **Simpler models**: By fitting separate models for each pathway, we avoid complex interaction terms
2. **Better convergence**: Each model is smaller and more focused, improving numerical stability
3. **Pathway-specific insights**: We can examine which factors matter most within each pathway
4. **Flexibility**: Different pathways can have different risk drivers and patterns

This approach provides the same hierarchical structure you wanted within each pathway, while avoiding the convergence issues that can happen with a single large model. The predicted probabilities can be calculated using any of the methods described above, with the tradeoff being between computational complexity and accuracy of uncertainty estimates.

For production use, we recommend the direct transformation method (Method 1) for most purposes, as it's fast and reliable. For more precise uncertainty quantification, the bootstrap approach (Method 3) provides the most comprehensive accounting for parameter uncertainty.
